---
title: "Project 3: What Characteristics Are Associated with Poor Hospital Performance?"
subtitle: "Merrimack College DSE6630: Healthcare & Life Sciences Analytics"
author: "Peter Gyorda"
date: "26 June 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      cache.comments = TRUE,
                      size = 13)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Turn off scientific notation
options(scipen = 999)

# Set seed for reproducibility
set.seed(50009)

# Clear workspace
rm(list = ls())

# Load required packages
pacman::p_load(
  tidyverse,
  readr,
  janitor,
  kableExtra,
  caret,
  randomForest,
  ranger
)
```

# 1 Background to the Project

## 1.1 Overview

Hospital quality has become an increasingly important focus for healthcare organizations, patients, and policymakers. The Centers for Medicare & Medicaid Services (CMS) publicly reports a variety of quality measures, including mortality, readmission, and patient safety outcomes. These measures provide an opportunity to identify characteristics associated with poor-performing hospitals and to better understand factors that may influence quality of care.

This project analyzes publicly available CMS hospital quality data to determine whether hospital characteristics, such as ownership, hospital type, emergency service availability, and geographic location, are associated with poor performance. The analysis also explores whether these characteristics can be used to predict poor-performing hospitals using machine learning techniques.

## 1.2 Background

CMS reports numerous quality measures for hospitals across the United States. Rather than relying solely on an overall hospital rating, this project creates a custom **Poor Performance Score** by combining the number of mortality, readmission, and patient safety measures in which each hospital performed worse than the national average. This approach provides a more detailed assessment of hospital performance across multiple quality domains.

## 1.3 Project Objectives

The objectives of this project are to:

* Explore the distribution of hospital quality performance across the United States.
* Identify hospital characteristics associated with poor performance.
* Evaluate differences in performance based on hospital ownership, hospital type, and emergency service availability.
* Examine geographic variation in hospital quality.
* Develop a Random Forest model to determine whether publicly available hospital characteristics can predict poor-performing hospitals.
* Provide recommendations that hospital administrators can use to prioritize quality improvement efforts.

## 1.4 Stakeholder

The primary stakeholders for this analysis are hospital administrators and healthcare system leadership teams seeking to identify organizational characteristics associated with poor hospital performance and opportunities for quality improvement.


# 2 Data Import and Initial Exploration

## 2.1 Load the CMS Hospital Dataset

```{r}
library(readr)

hospital <- read_csv("Hospital_General_Information.csv")
```

## 2.2 Examine the Dataset

```{r}
dim(hospital)

names(hospital)

glimpse(hospital)

summary(hospital)
```

## 2.3 Missing Data Assessment

```{r}
colSums(is.na(hospital))
```




### 2.4 Data Cleaning

```{r}
hospital_clean <- hospital %>%
  mutate(
    mort_worse = as.numeric(na_if(`Count of MORT Measures Worse`,
                                  "Not Available")),

    readm_worse = as.numeric(na_if(`Count of READM Measures Worse`,
                                   "Not Available")),

    safety_worse = as.numeric(na_if(`Count of Safety Measures Worse`,
                                    "Not Available"))
  )


```


### Verify the cleaned variables
```{r}
summary(hospital_clean[, c("mort_worse",
                           "readm_worse",
                           "safety_worse")])
```


The three variables were successfully converted to numeric values, with missing values representing hospitals where the measures were reported as "Not Available."

## 2.5 Create the Poor Performance Score
```{r}
hospital_clean <- hospital_clean %>%
  mutate(
    PoorPerformanceScore = rowSums(
      cbind(
        mort_worse,
        readm_worse,
        safety_worse
      ),
      na.rm = FALSE
    )
  )

summary(hospital_clean$PoorPerformanceScore)

sum(is.na(hospital_clean$PoorPerformanceScore))
```




## 2.6 Remove Missing Observations
```{r}
hospital_final <- hospital_clean %>%
  filter(!is.na(PoorPerformanceScore))

dim(hospital_final)
```

## 2.7 Distribution of the Poor Performance Score
```{r}
ggplot(hospital_final,
       aes(x = PoorPerformanceScore)) +
  geom_histogram(binwidth = 1,
                 fill = "steelblue",
                 color = "black") +
  labs(
    title = "Distribution of Poor Performance Scores",
    x = "Poor Performance Score",
    y = "Number of Hospitals"
  ) +
  theme_minimal()
```


# 3 Exploratory Data Analysis

## 3.1 Distribution of the Poor Performance Score

```{r}
# Remove hospitals with missing Poor Performance Scores
hospital_analysis <- hospital_clean %>%
  filter(!is.na(PoorPerformanceScore))

# Summary statistics
summary(hospital_analysis$PoorPerformanceScore)

# Histogram
ggplot(hospital_analysis,
       aes(x = PoorPerformanceScore)) +
  geom_histogram(
    binwidth = 1,
    fill = "steelblue",
    color = "black"
  ) +
  labs(
    title = "Distribution of Poor Performance Scores",
    x = "Poor Performance Score",
    y = "Number of Hospitals"
  ) +
  theme_minimal()
```

## 3.2 Hospital Ownership
```{r}
ownership_summary <- hospital_analysis %>%
  group_by(`Hospital Ownership`) %>%
  summarise(
    Hospitals = n(),
    MeanScore = mean(PoorPerformanceScore),
    MedianScore = median(PoorPerformanceScore),
    SD = sd(PoorPerformanceScore)
  ) %>%
  arrange(desc(MeanScore))

ownership_summary
```


```{r}
ggplot(hospital_analysis,
       aes(x = `Hospital Ownership`,
           y = PoorPerformanceScore)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Poor Performance Score by Hospital Ownership",
    x = "Hospital Ownership",
    y = "Poor Performance Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```


Hospital Ownership. Government-owned state hospitals had the highest Poor Performance Scores. Most ownership groups had median scores of 0 or 1, although outliers were present across several ownership types.


## 3.3 Hospital Type
The Poor Performance Score was compared across hospital types to determine whether certain types of hospitals were associated with higher quality concerns.

```{r}
type_summary <- hospital_analysis %>%
  group_by(`Hospital Type`) %>%
  summarise(
    Hospitals = n(),
    MeanScore = round(mean(PoorPerformanceScore), 2),
    MedianScore = round(median(PoorPerformanceScore), 2),
    SD = round(sd(PoorPerformanceScore), 2)
  ) %>%
  arrange(desc(MeanScore))

type_summary
```

```{r}
ggplot(hospital_analysis,
       aes(x = `Hospital Type`,
           y = PoorPerformanceScore)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Poor Performance Score by Hospital Type",
    x = "Hospital Type",
    y = "Poor Performance Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

*Hospital Type.* Hospital performance varied across hospital types. Differences in the median scores and several outliers suggest that hospital type may be associated with quality performance.


## 3.4 Emergency Services
The Poor Performance Score was compared between hospitals that provide emergency services and those that do not.

```{r}
emergency_summary <- hospital_analysis %>%
  group_by(`Emergency Services`) %>%
  summarise(
    Hospitals = n(),
    MeanScore = round(mean(PoorPerformanceScore), 2),
    MedianScore = round(median(PoorPerformanceScore), 2),
    SD = round(sd(PoorPerformanceScore), 2)
  )

emergency_summary
```

Emergency Services. Hospitals with emergency services had higher Poor Performance Scores than hospitals without emergency services. The average score was 0.99 for hospitals with emergency services compared with 0.66 for hospitals without emergency services.


## 3.5 State Analysis

The average Poor Performance Score was calculated for each state.
```{r}
# Average Poor Performance Score by state
state_summary <- hospital_analysis %>%
  group_by(State) %>%
  summarise(
    Hospitals = n(),
    MeanScore = round(mean(PoorPerformanceScore), 2),
    .groups = "drop"
  ) %>%
  filter(Hospitals >= 20) %>%
  arrange(desc(MeanScore))

# Top 10 states with at least 20 hospitals
top10_states <- state_summary %>%
  slice_head(n = 10)

top10_states

ggplot(top10_states,
       aes(x = reorder(State, MeanScore),
           y = MeanScore)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 States by Average Poor Performance Score",
    x = "State",
    y = "Average Poor Performance Score"
  ) +
  theme_minimal()
```
State Analysis. After limiting the analysis to states with at least 20 hospitals, New Jersey had the highest average Poor Performance Score, followed by New York and Massachusetts.

# 4 Correlation Analysis

The relationships between mortality, readmission, and safety measures were examined using Pearson correlation coefficients.

```{r}
# Select variables for correlation analysis
correlation_data <- hospital_analysis %>%
  select(
    mort_worse,
    readm_worse,
    safety_worse
  )

# Calculate correlation matrix
cor_matrix <- round(
  cor(
    correlation_data,
    use = "complete.obs"
  ),
  2
)

cor_matrix
```
Correlation Analysis. The mortality, readmission, and safety measures showed weak positive correlations (0.12–0.19). This suggests that each measure provides different information about hospital performance and supports combining them into the Poor Performance Score.


# 5 Poor Performer Classification

Hospitals were classified as poor performers if their Poor Performance Score was at or above the 75th percentile. This binary outcome variable will be used in the Random Forest classification model.

```{r}
# Calculate the 75th percentile cutoff
cutoff <- quantile(
  hospital_analysis$PoorPerformanceScore,
  probs = 0.75,
  na.rm = TRUE
)

# Display the cutoff value
cutoff

# Create the Poor Performer classification
hospital_analysis <- hospital_analysis %>%
  mutate(
    PoorPerformer = ifelse(
      PoorPerformanceScore >= cutoff,
      "Yes",
      "No"
    )
  )

# Convert to a factor for modeling
hospital_analysis$PoorPerformer <- factor(
  hospital_analysis$PoorPerformer,
  levels = c("No", "Yes")
)

# Display the frequency of each class
table(hospital_analysis$PoorPerformer)
```
**Poor Performer Classification.** Hospitals with Poor Performance Scores at or above the 75th percentile were classified as poor performers and used as the target variable for the Random Forest model.

Poor Performer Classification. Hospitals with a Poor Performance Score of 2 or greater (the top 25% of scores) were classified as poor performers. This resulted in 772 poor-performing hospitals and 2,316 non-poor-performing hospitals.

## 6.1 Prepare the Data


```{r}
# Select variables for the Random Forest model
rf_data <- hospital_analysis %>%
  select(
    PoorPerformer,
    `Hospital Ownership`,
    `Hospital Type`,
    `Emergency Services`,
    State,
    `Hospital overall rating`
  ) %>%
  drop_na() %>%
  filter(!State %in% c("GU", "VI"))

# Rename columns for modeling
colnames(rf_data) <- c(
  "PoorPerformer",
  "HospitalOwnership",
  "HospitalType",
  "EmergencyServices",
  "State",
  "HospitalRating"
)

str(rf_data)
```

## 6.2 Split the Data

The dataset was randomly divided into training (80%) and testing (20%) sets. Predictor variables were converted to factors prior to fitting the Random Forest model.
```{r}
# Split the data
set.seed(50009)

train_index <- createDataPartition(
  rf_data$PoorPerformer,
  p = 0.80,
  list = FALSE
)

train_data <- rf_data[train_index, ]
test_data  <- rf_data[-train_index, ]

# Convert predictors to factors
factor_vars <- c(
  "HospitalOwnership",
  "HospitalType",
  "EmergencyServices",
  "State",
  "HospitalRating"
)

train_data[factor_vars] <- lapply(
  train_data[factor_vars],
  factor
)

test_data[factor_vars] <- lapply(
  test_data[factor_vars],
  factor
)

# Force identical factor levels
for(v in factor_vars){
  test_data[[v]] <- factor(
    test_data[[v]],
    levels = levels(train_data[[v]])
  )
}

# Convert to data frames
train_data <- as.data.frame(train_data)
test_data  <- as.data.frame(test_data)

# Display dimensions
dim(train_data)
dim(test_data)
```


## 6.3 Build the Random Forest Model

```{r}
set.seed(50009)

rf_model <- randomForest(
  PoorPerformer ~ .,
  data = train_data,
  ntree = 500,
  importance = TRUE
)

rf_model
```
Random Forest Model. The Random Forest model achieved an out-of-bag error rate of 22.3%, corresponding to an estimated accuracy of approximately 77.7%. The model was more successful at identifying hospitals that were not poor performers than hospitals classified as poor performers.

## 6.4 Evaluate Model Performance

The Random Forest model was evaluated using the independent testing dataset. Model performance was summarized using a confusion matrix.

```{r}
# Predict Poor Performer status for the testing data
rf_predictions <- predict(
  rf_model,
  newdata = test_data
)

# Create the confusion matrix
rf_results <- confusionMatrix(
  rf_predictions,
  test_data$PoorPerformer
)

# Display the results
rf_results
```


Model Performance. The Random Forest model achieved an overall accuracy of 78.3% on the testing dataset. The model showed high sensitivity (88.6%) for identifying hospitals that were not poor performers but a lower specificity (47.4%) for identifying poor-performing hospitals.

## 6.5 Variable Importance

The Random Forest model calculates the relative importance of each predictor variable. Variables with higher importance contribute more to correctly classifying hospitals as poor performers.

```{r}
# Display the variable importance plot
varImpPlot(
  rf_model,
  main = "Variable Importance for Predicting Poor Hospital Performance"
)
```

**Variable Importance.** The variable importance plot ranks the predictor variables based on their contribution to the Random Forest model. Variables with higher importance scores have a greater influence on predicting whether a hospital is classified as a poor performer.


Variable Importance. Hospital overall rating was the most important predictor of poor hospital performance, followed by state and hospital ownership. Emergency services contributed the least to the Random Forest model.

# 7 Executive Summary

This project examined characteristics associated with poor hospital performance using publicly available CMS hospital quality data. A Poor Performance Score was developed by combining mortality, readmission, and patient safety measures in which hospitals performed worse than the national average.

Exploratory analyses found differences in Poor Performance Scores across hospital ownership, hospital type, emergency service availability, and geographic location. Government-owned state hospitals generally had the highest average Poor Performance Scores, while Critical Access Hospitals tended to have the lowest scores.

Hospitals with Poor Performance Scores in the top 25% were classified as poor performers. A Random Forest classification model was developed using hospital characteristics to predict poor-performing hospitals. The model achieved an overall accuracy of **78.3%** on the testing dataset.

Variable importance analysis showed that Hospital Overall Rating was the strongest predictor of poor hospital performance, followed by State and Hospital Ownership. Emergency Services contributed the least to the prediction model.

Overall, the results suggest that publicly available hospital characteristics can help identify hospitals that may be at greater risk for poor performance. Hospital administrators may use this information to prioritize quality improvement efforts and better understand factors associated with hospital quality.