---
title: "Project 3: What Characteristics Are Associated with Poor Hospital Performance?"
subtitle: "Merrimack College DSE6630: Healthcare & Life Sciences Analytics"
author: "Peter Gyorda"
date: "26 June 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      cache.comments = TRUE,
                      size = 13)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Turn off scientific notation
options(scipen = 999)

# Set seed for reproducibility
set.seed(50009)

# Clear workspace
rm(list = ls())

# Load required packages
pacman::p_load(
  tidyverse,
  readr,
  janitor,
  kableExtra,
  caret,
  randomForest,
  ranger
)
```

# 1 Background to the Project

## 1.1 Overview

Hospital quality has become an increasingly important focus for healthcare organizations, patients, and policymakers. The Centers for Medicare & Medicaid Services (CMS) publicly reports a variety of quality measures, including mortality, readmission, and patient safety outcomes. These measures provide an opportunity to identify characteristics associated with poor-performing hospitals and to better understand factors that may influence quality of care.

This project analyzes publicly available CMS hospital quality data to determine whether hospital characteristics, such as ownership, hospital type, emergency service availability, and geographic location, are associated with poor performance. The analysis also explores whether these characteristics can be used to predict poor-performing hospitals using machine learning techniques.

## 1.2 Background

CMS reports numerous quality measures for hospitals across the United States. Rather than relying solely on an overall hospital rating, this project creates a custom **Poor Performance Score** by combining the number of mortality, readmission, and patient safety measures in which each hospital performed worse than the national average. This approach provides a more detailed assessment of hospital performance across multiple quality domains.

## 1.3 Project Objectives

The objectives of this project are to:

* Explore the distribution of hospital quality performance across the United States.
* Identify hospital characteristics associated with poor performance.
* Evaluate differences in performance based on hospital ownership, hospital type, and emergency service availability.
* Examine geographic variation in hospital quality.
* Develop a Random Forest model to determine whether publicly available hospital characteristics can predict poor-performing hospitals.
* Provide recommendations that hospital administrators can use to prioritize quality improvement efforts.

## 1.4 Stakeholder

The primary stakeholders for this analysis are hospital administrators and healthcare system leadership teams seeking to identify organizational characteristics associated with poor hospital performance and opportunities for quality improvement.


# 2 Data Import and Initial Exploration

## 2.1 Load the CMS Hospital Dataset

```{r}
library(readr)

hospital <- read_csv("Hospital_General_Information.csv")
```

## 2.2 Examine the Dataset

```{r}
dim(hospital)

names(hospital)

glimpse(hospital)

summary(hospital)
```

## 2.3 Missing Data Assessment

```{r}
colSums(is.na(hospital))
```




### 2.4 Data Cleaning

```{r}
hospital_clean <- hospital %>%
  mutate(
    mort_worse = as.numeric(na_if(`Count of MORT Measures Worse`,
                                  "Not Available")),

    readm_worse = as.numeric(na_if(`Count of READM Measures Worse`,
                                   "Not Available")),

    safety_worse = as.numeric(na_if(`Count of Safety Measures Worse`,
                                    "Not Available"))
  )


```


### Verify the cleaned variables
```{r}
summary(hospital_clean[, c("mort_worse",
                           "readm_worse",
                           "safety_worse")])
```


The three variables were successfully converted to numeric values, with missing values representing hospitals where the measures were reported as "Not Available."

## 2.5 Create the Poor Performance Score
```{r}
hospital_clean <- hospital_clean %>%
  mutate(
    PoorPerformanceScore = rowSums(
      cbind(
        mort_worse,
        readm_worse,
        safety_worse
      ),
      na.rm = FALSE
    )
  )

summary(hospital_clean$PoorPerformanceScore)

sum(is.na(hospital_clean$PoorPerformanceScore))
```




## 2.6 Remove Missing Observations
```{r}
hospital_final <- hospital_clean %>%
  filter(!is.na(PoorPerformanceScore))

dim(hospital_final)
```

## 2.7 Distribution of the Poor Performance Score
```{r}
ggplot(hospital_final,
       aes(x = PoorPerformanceScore)) +
  geom_histogram(binwidth = 1,
                 fill = "steelblue",
                 color = "black") +
  labs(
    title = "Distribution of Poor Performance Scores",
    x = "Poor Performance Score",
    y = "Number of Hospitals"
  ) +
  theme_minimal()
```


# 3 Exploratory Data Analysis

## 3.1 Distribution of the Poor Performance Score

```{r}
# Remove hospitals with missing Poor Performance Scores
hospital_analysis <- hospital_clean %>%
  filter(!is.na(PoorPerformanceScore))

# Summary statistics
summary(hospital_analysis$PoorPerformanceScore)

# Histogram
ggplot(hospital_analysis,
       aes(x = PoorPerformanceScore)) +
  geom_histogram(
    binwidth = 1,
    fill = "steelblue",
    color = "black"
  ) +
  labs(
    title = "Distribution of Poor Performance Scores",
    x = "Poor Performance Score",
    y = "Number of Hospitals"
  ) +
  theme_minimal()
```

## 3.2 Hospital Ownership
```{r}
ownership_summary <- hospital_analysis %>%
  group_by(`Hospital Ownership`) %>%
  summarise(
    Hospitals = n(),
    MeanScore = mean(PoorPerformanceScore),
    MedianScore = median(PoorPerformanceScore),
    SD = sd(PoorPerformanceScore)
  ) %>%
  arrange(desc(MeanScore))

ownership_summary
```


```{r}
ggplot(hospital_analysis,
       aes(x = `Hospital Ownership`,
           y = PoorPerformanceScore)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Poor Performance Score by Hospital Ownership",
    x = "Hospital Ownership",
    y = "Poor Performance Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```


Hospital Ownership. Government-owned state hospitals had the highest Poor Performance Scores. Most ownership groups had median scores of 0 or 1, although outliers were present across several ownership types.


## 3.3 Hospital Type
The Poor Performance Score was compared across hospital types to determine whether certain types of hospitals were associated with higher quality concerns.

```{r}
type_summary <- hospital_analysis %>%
  group_by(`Hospital Type`) %>%
  summarise(
    Hospitals = n(),
    MeanScore = round(mean(PoorPerformanceScore), 2),
    MedianScore = round(median(PoorPerformanceScore), 2),
    SD = round(sd(PoorPerformanceScore), 2)
  ) %>%
  arrange(desc(MeanScore))

type_summary
```

```{r}
ggplot(hospital_analysis,
       aes(x = `Hospital Type`,
           y = PoorPerformanceScore)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Poor Performance Score by Hospital Type",
    x = "Hospital Type",
    y = "Poor Performance Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

*Hospital Type.* Hospital performance varied across hospital types. Differences in the median scores and several outliers suggest that hospital type may be associated with quality performance.


## 3.4 Emergency Services
The Poor Performance Score was compared between hospitals that provide emergency services and those that do not.

```{r}
emergency_summary <- hospital_analysis %>%
  group_by(`Emergency Services`) %>%
  summarise(
    Hospitals = n(),
    MeanScore = round(mean(PoorPerformanceScore), 2),
    MedianScore = round(median(PoorPerformanceScore), 2),
    SD = round(sd(PoorPerformanceScore), 2)
  )

emergency_summary
```

Emergency Services. Hospitals with emergency services had higher Poor Performance Scores than hospitals without emergency services. The average score was 0.99 for hospitals with emergency services compared with 0.66 for hospitals without emergency services.


## 3.5 State Analysis

The average Poor Performance Score was calculated for each state.
```{r}
# Average Poor Performance Score by state
state_summary <- hospital_analysis %>%
  group_by(State) %>%
  summarise(
    Hospitals = n(),
    MeanScore = round(mean(PoorPerformanceScore), 2),
    .groups = "drop"
  ) %>%
  filter(Hospitals >= 20) %>%
  arrange(desc(MeanScore))

# Top 10 states with at least 20 hospitals
top10_states <- state_summary %>%
  slice_head(n = 10)

top10_states

ggplot(top10_states,
       aes(x = reorder(State, MeanScore),
           y = MeanScore)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 States by Average Poor Performance Score",
    x = "State",
    y = "Average Poor Performance Score"
  ) +
  theme_minimal()
```
State Analysis. After limiting the analysis to states with at least 20 hospitals, New Jersey had the highest average Poor Performance Score, followed by New York and Massachusetts.




