---
title: "Demo 4: Patient Privacy & Data Integrity"
subtitle: "Merrimack College DSE6630: Healthcare & Life Sciences Analytics"
author: "Team Epsilon: Desi Spark (DS), Elizabeth Greenan (EG), Meghan Holden (MH)"
date: "26 June 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_crop: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,    
                      cache.comments = TRUE,
                      size = 13)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# clean, set up, and load
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
rm(list = ls(all = TRUE))

pacman::p_load(here,
               tidyverse, 
               ggplot2, 
               kableExtra,
               forcats,
               jsonlite, 
               lubridate,
               sqldf, 
               SQLite, 
               RSQLite,
               RMySQL,
               DBI)

## load the data files
load("patient-data.Rdata")
## a little tom-foolery to make a fake, local instance of a database to practice queries
con <- dbConnect(RSQLite::SQLite(), ":memory:")
dbWriteTable(con, "patient_info", patient_info)
dbWriteTable(con, "patient_disability", patient_disability)
dbWriteTable(con, "patient_clinical", patient_clinical)
dbWriteTable(con, "patient_insurance", patient_insurance)
dbWriteTable(con, "patient_binge_drinking", patient_binge_drinking)

## now unload the datafiles to enhance the simulation so it looks just like connecting to an external database!
rm(patient_info)
rm(patient_disability)
rm(patient_clinical)
rm(patient_insurance)
rm(patient_binge_drinking)
```

# 1 Introduction

You have just been brought onto a new project that is using an __mHealth__ (mobile health) application (**"iQOL"**, where QOL stands for "Quality of Life") to track the health of patients with and without disabilities. The goal of this mHealth application is to better understand how functional and quality of life differs day-to-day among patients with and without a disability.

As a first pass, you ask the back-end (database) development team to provide you with some descriptive data from the first 100,000 users of the mHealth application. You want to make a report to stakeholders about who is using the application, the nature of the (dis) ability, geographic breakdown, and descriptive statistics about other information gathered by the mHealth application when a patient first signs up for the mHealth Tracker. 

The app's back-end developer sends you several tables from the database in an Rdata file, with the following descriptions of fields:

```{r, echo=FALSE}
dict <- tribble(
  ~ Variable, ~ Description,
  "PT_KEY", "Patient unique identifier",
  "INS_KEY", "Insurance unique identifier",
  "FIRST_NAME", "Patient first name",
  "SURNAME",  "Patient surname",
  "DOB",  "Date-of-birth",
  "AGE_DAYS",   "Age in days at time of database pull",
  "AGE_YEARS",  "Age in years at time of database pull",
  "EMAIL_ADDRESS", "Patient email address",
  "STREET",   "Patient mailing address",
  "CITY",  "City of mailing address",
  "STATE",  "State of mailing address",
  "ZIPCODE", "Zipcode of mailing address",
  "TELEPHONE", "Patient primary phone number",
  "ADULT_CHILD", "Flag if user is under 18 (we're developing a separate app for teens)",
  "FLU_INNOC", "Patient states they received their influenza vaccine this year",
  "COVID_INNOC", "Patient states they received their COVID-19 vaccine this year",
  "TET_INNOC", "Patient states they received their tetanus vaccine in past 10 years",
  "DATE_LAST_SEEN", "Patient reported last date seen by primary care physician",
  "NEW_PATIENT", "New user to the application",
  "PUBLIC_INS", "Patient reports receiving public insurance (Medicaid, Medicare, VA Benefits)",
  "DISABILITY_STATUS", "Patient-reported disability status, per federal guidelines",
  "EMPLOYMT_STATUS", "Patient-reported employment status for past 6 months",
  "EDUC_LEVEL", "Patient-reported highest educational attainment",
  "BMI_CATEGORY", "Body mass index categorization; BMI < 18.5 Underweight, 18.5 < BMI < 25 Normal, 25 <= BMI < 30 Overweight, BMI >= 30 Obese",
  "SMOKING_STATUS", "Patient is a former or current smoker, or has never smoked",
  "BINGE_DRINKING", "Patient reports at least one episode of binge drinking in past 6 months"
)

kable(
    dict,
    format = "html",
    col.names = c("", ""),
    caption = "Table 1. Data dictionary of available fields in the mHealth database pull.") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

## 1.1 Database connections

If your developer had given you `.db` files, or if you were accessing a server, you would have needed to walk through two steps, which we will simulate here, in order to load the tables into R for analysis. 

**Importantly**, the general purpose of doing this work in R will generally be for **research/analysis only**. However, the advantage to writing it in SQL is two-fold:

1. You can use your SQL queries on the server itself to do the same work, say, for a report you need to generate regularly.

2. If you DO need to update tables in the database, you can actually do that from R!!

**A word of caution**, however: if you've been given access to the __production__ databases, do not ever alter or update the tables unless you've been given explicit permission to do so. 

### 1.1.1 Making connections

```{r, echo=TRUE, eval=FALSE}
## This is how you would set a connection to an external server

# con = dbConnect(MySQL(),
#                 dbname = 'my_db_name',
#                 host = 'my_host',
#                 port = 3306,   ## The typical port used for SQL and MySQL servers
#                 user = 'username',
#                 password = 'password')

## Can be helpful to establish which flavor of SQL

# options(sqldf.driver = "SQLite")
```

But here, I've already made a connection to our mock database, also stored into `con`. So we will use that.

### 1.1.2 How many tables?

Usually the first thing we want to know if we're unfamiliar with a database is **how many tables are in it**. We are going to make our first SQL query together to get the names of all the tables in the connection, `con`. Notice that I have to use `dbGetQuery()` as my function to set the query, then I have to specify the name of the connection, `con`. The actual query itself is then given as a string in quotations after that.

```{r}
dbGetQuery(con, "SELECT name FROM sqlite_master WHERE type='table'")
```

### 1.1.3 What's in each table?

Notice that, unlike when we work with dataframes directly in R, we can't __see__ the tables. I did this on purpose; I'm hiding them from you so that it's as much like a simulated database experience as possible.

So, if we're flying blind, how can we find out what's in the tables? Simple - we write a query! Since we have five tables, the easiest way to do this is probably to write a `for` loop, where we query each table as we loop. 

Although there are multiple ways we could choose to execute this, perhaps some more clean than others, we are going to use the SQLite statement `PRAGMA`, which is a SQLite-specific extension and can be used to modify an operation or query a SQLite library. We're using it for the second purpose. In other words, we will use `PRAGMA` to allow us to query each table using `dbGetQuery`. Confused yet? It's okay if you are!

1. Make a query using `dbGetQuery` to get all of the names of the tables on the connection `con`. We will store these table names into `tables`. (Note that I didn't call them `table` because that's the name of a base `R` function! Best programming practice will ALWAYS have us not overwrite base objects; but in `R`, there are no protected name spaces, like there are in some other programming languages. So, we as the programmer must be cautious.)

2. Start the `for` loop, looping through each table by its `name`. 

3. Make a second query using `dbGetQuery`, this time setting the connection `con` AND using `PRAGMA` to get the `table_info` of each table. `table_info` returns a set of pre-specified fields on information, but I am only going to print the first two fields for brevity.

And that's it! A little clearer? If not, work through this code trying to change variables to help yourself figure out how it works. It's a **query within a query!**

```{r}
## 1. Make a query to get the list of tables from the connection, which we store in tables
tables <- dbGetQuery(con, "SELECT name FROM sqlite_master WHERE type = 'table' ")

# Loop over each table by name and print the table information 
for (t in tables$name) {
  ## Store the table information after making a second table-specific query about what's in the table using PRAGMA and table_info()
  table_info <- dbGetQuery(con, paste0("PRAGMA table_info(", t, ")"))
  ## Print the name of the table so we know the order 
  print(t)
  ## Show the first two columns from the table_info() of each table in the connection
  print(table_info[, 1:2])
}
```

It might feel like we're starting a bit deep, but don't worry - we can get a lot of information using simpler queries too, including assessing how well our developer has been protecting __patient privacy__.


# 2 Ensuring Patient Privacy

You had already decided, since you're still learning **SQL (Structured Query Language)** to continue using that in R for practice rather than the more familiar `tidyverse`. See, we would have the option to __store our dataframes__ from the connection `con` and manipulate them in all the familiar ways! And we will, in good time. But first, we are going to continue to use `dbGetQuery()` from the `SQLite` package to help us to make our queries.

You noticed as soon as you saw the fields of some of the tables that the developer sent that there's a __big problem!!__ The identities of your mHealth users are __NOT__ being safeguarded correctly. Uh-oh!

##### Just how bad IS the problem?

## 2.1 Checking for duplicate keys

You decide to first check for duplicates because, well, you asked for 100,000 adult patients... did you actually get 100,000? You start in the `patient_info` table because you see a lot of problems there in terms of protecting patient privacy!

```{r}
dbGetQuery(con, "SELECT pt_key, 
  COUNT(*) AS patient_count
  FROM patient_info
  GROUP BY pt_key
  HAVING patient_count > 1")
```

Whew! No problems here. Notice first that, to write this query, we are using the `dbGetQuery()` function in R (which is serving as a __wrapper__) to call our SQL query on our database connection, `con`. Our query is given as a string in quotations.  Also notice that even though the __field__ in our table is `PT_KEY`, in SQL letter case doesn't matter. In fact, upper case is usually reserved for certain __actions__ in SQL: `SELECT`, `COUNT`, `FROM`, `GROUP BY`, and `HAVING`. These are the __functions__. However, you don't even have to use the case for your query to run! Notice, too, that we used the `(*)` wildcard to select ALL patient keys. 

Also, note that even though I put each statement on separate lines (which is how we often write it in a query on a SQL server), we do not have to. We just do that to improve _readability of the code_. 

#### Question 1 [1 point]:

Try to translate into English exactly what this SQL query does.

EG

> The above SQL query selects the patient unique identifier (pt_key) and counts the number of rows, which in this instance, represent patients, with that unique identifier. The results are grouped by pt_key. The HAVING function filters the results of the GROUP BY function on if there is more than 1 patient with that unique identifer. Therefore, if there were any duplicates, they would have been listed in the output.

#### Question 2 [1 point]:

You try it: check for duplicate insurance keys. Make sure to query the right table! (**Hint**: You should get 81.)

EG

> Here, I updated the SELECT and GROUP BY statements to look for ins_key and updated the FROM statement to look at the patient_insurance table.

```{r}
dbGetQuery(con, "SELECT ins_key,
  COUNT(*) AS patient_count
  FROM patient_insurance
  GROUP BY ins_key
  HAVING patient_count > 1")
```

> There are 81 duplicate insurance keys. These are located in the patient_insurance table.

#### Question 3 [1 point]:

Is this a problem? Why or why not? (Hint: Can multiple individuals be on the same insurance policy?)

EG

> Duplicate insurance unique idenitifers (ins_key) are not a problem. Since multiple people can be on a policy, duplicates would make sense. Therefore, this shows that there are 81 policies that have multiple people on the insurance policy. This is much different than the issue of a duplicate patient unique identifiers (pt_key). In that case, having a duplicate record for a patient would be a problem, as those must be unique.

## 2.2 Dropping fields with sensitive data

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_info") %>% 
head() %>% 
  kable(
    format = "html",
    caption = "Table 2. The patient_info table BEFORE removing any sensitive information. ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

Let's start by dropping ALL of the ultra sensitive data from the `patient_info` table: date-of-birth, age_days, name fields, street address, telephone, & email address from `patient_info`. 

In SQL, there is a function called `DROP COLUMN` that unfortunately is not supported by any of the packages we are using in R. Instead, we will have to use the `SELECT` function to choose the fields we want to keep. 

But note that it is not nearly as straightforward as if we were doing it in R. We will have to go through a series of steps to do it correctly; NOT ONLY THAT, __but it is making permanent changes to the database connection!__ PROCEED WITH CAUTION in the real world!

This is why there are often two versions of the database, one that is in __production__ and one that is used for __research/analysis__. Thus, we can think of this as we are making corrections to the analytical database. But we will also need to make sure to __give our suggestions__ back to the developer!!

* 1. Make a copy of the `patient_info` table, called `temp`, from just the columns we want to __keep__.

* 2. Now, this is blank. We have to actually `INSERT INTO` the new table, `temp`, the columns we want!

* 3. Remove, or `DROP TABLE` the original `patient_info` table

* 4. `RENAME` the table `temp` back to the original name, `patient_info`, by using the `ALTER TABLE` statement.

Notice too that to execute these table-level functions I am now using `dbExecute()` from `SQLite` rather than `dbGetQuery()` because these are not queries. These are executive statements.

```{r, results='hide'}
# Step 1
dbExecute(con, "CREATE TABLE temp AS 
  SELECT pt_key, age_years, city, state, zipcode 
  FROM patient_info")

# Step 2
dbExecute(con, "INSERT INTO temp (pt_key, age_years, city, state, zipcode)
  SELECT  pt_key, age_years, city, state, zipcode
  FROM patient_info")

# Step 3
dbExecute(con, "DROP TABLE patient_info")

# Step 4
dbExecute(con, "ALTER TABLE temp RENAME TO patient_info")
```

**Important**: We have permanently replaced the original `patient_info` table the developer gave us!!!

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_info") %>% 
head() %>% 
  kable(
    format = "html",
    caption = "Table 3. The patient_info table AFTER removing any sensitive information. ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

#### Question 4 [1 point]:
Your turn!

Remove all the sensitive information remaining in any of the other tables.

EG

```{r}

# Check the fields in each of the remaining tables

# patient_disability table
dbGetQuery(con, "PRAGMA table_info(patient_disability)")

# patient_clinical table
dbGetQuery(con, "PRAGMA table_info(patient_clinical)")

# patient_insurance table
dbGetQuery(con, "PRAGMA table_info(patient_insurance)")

# patient_binge_drinking table
dbGetQuery(con, "PRAGMA table_info(patient_binge_drinking)")
```

> The patient_disability, patient_insurance, and patient_binge_drinking do not contain any sensitive information that would need to be removed. They contain fields such as keys, status, and insurance information, none of which are required to be removed. The patient_clinical table has more fields, including some with sensitive information. The fields that need to be removed form the patient_clinical table include surname and date of birth (DOB).

```{r}

# Remove SURNAME AND DOB from the patient_clinical table

# Step 1
dbExecute(con, "CREATE TABLE temp AS 
  SELECT pt_key, flu_innoc, tet_innoc, covid_innoc, bmi_category, smoking_status 
  FROM patient_clinical")

# Step 2
dbExecute(con, "INSERT INTO temp (pt_key, flu_innoc, tet_innoc, covid_innoc, bmi_category, smoking_status)
  SELECT  pt_key, flu_innoc, tet_innoc, covid_innoc, bmi_category, smoking_status
  FROM patient_clinical")

# Step 3
dbExecute(con, "DROP TABLE patient_clinical")

# Step 4
dbExecute(con, "ALTER TABLE temp RENAME TO patient_clinical")

```

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_clinical") %>% 
head() %>% 
  kable(
    format = "html",
    caption = "Table 3. The patient_clinical table AFTER removing any sensitive information. ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

Not sure what to do? Load the R script and run it instead to proceed.

> EG: I ran this code and confirmed the patient_clinical table in the R file matches what I have above.

```{r}
source("q4_answer.R")
```

Make sure to demonstrate that the sensitive information is gone using the follow code that employs the `head()` function after removing the comments and making it executable. 

```{r, echo = TRUE}
 dbGetQuery(con, "select * from patient_info") %>% ## patient_info table
 head() %>% 
   kable(
     format = "html",
     caption = "Table 4. The table AFTER removing sensitive information. ") %>%
     kable_styling(bootstrap_options = c("hover"),
 )


 dbGetQuery(con, "select * from patient_disability") %>% ## patient_disability table
 head() %>% 
   kable(
     format = "html",
     caption = "Table 4. The table AFTER removing sensitive information. ") %>%
     kable_styling(bootstrap_options = c("hover"),
 )
 
  dbGetQuery(con, "select * from patient_clinical") %>% ## patient_clinical table
 head() %>% 
   kable(
     format = "html",
     caption = "Table 4. The table AFTER removing sensitive information. ") %>%
     kable_styling(bootstrap_options = c("hover"),
 )
  
    dbGetQuery(con, "select * from patient_insurance") %>% ## patient_insurance table
 head() %>% 
   kable(
     format = "html",
     caption = "Table 4. The table AFTER removing sensitive information. ") %>%
     kable_styling(bootstrap_options = c("hover"),
 )
    
      dbGetQuery(con, "select * from patient_binge_drinking") %>% ## patient_binge_drinking table
 head() %>% 
   kable(
     format = "html",
     caption = "Table 4. The table AFTER removing sensitive information. ") %>%
     kable_styling(bootstrap_options = c("hover"),
 )
```


#### Question 5 [1 point]:
**Have you noticed that I am having to first extract the table as a dataframe using `dbGetQuery()` before I can pass it through `head()`?**

Why do you think I have to do this?

EG

> I think you have to do this because head() works on dataframes. It would not work on the table itself without this step. dbGetQuery() ensures that the table is now a dataframe, so it will work when using head().

## 2.3 Preserving integrity while removing anything that can identify a patient.

When looking at the `ZIPCODE` column of the `patient_info` table, you start to worry about some of the entries. Some zipcodes have the four-digit US postal service identifiers, which can really narrow down a patient's identity to a much smaller geographic region. This makes you uneasy (and me too!). So, we decide to preserve zipcode (a much larger geographic area) while dropping the four-digits at the end when they exist. 

You know you could do this easily with `gsub()` as we've used before on a dataframe - but we aren't currently working with a dataframe! So, you decide to try doing it the "SQL-way" first to see if you can do it.

* 1. Create a `temp` table. This time, we are making it from everything in the `patient_info` table and we are specificying it as a `TEMPORARY TABLE` with our `CREATE` statement. This is to save us the two steps we had to do to make the `temp` table earlier.

```{r, results='hide'}
dbExecute(con, "CREATE TEMPORARY TABLE temp AS
  SELECT * FROM patient_info")
```

* 2. Now use the `UPDATE` statement with the `SUBSTR()` (substring) function to just keep the first 5 characters of the ZIP code. Note that we are doing this on the `temp` table and not `patient_info`.

```{r, results='show'}
dbExecute(con, 
          "UPDATE temp SET zipcode = SUBSTR(zipcode, 1, 5)")
```

Uh-oh... 200,000? Does that mean we __duplicated rows?__

```{r, results='show'}
dbGetQuery(con, 
           "SELECT COUNT(*) AS row_count FROM temp")
```

Noooooo! What happened? Weren't we just trying to lop off the "-xxxx" off the end of a 5-digit ZIP code? 

So, the problem is that SQL is not smart enough to know what to do when data match in more than one column. It errs on the side of caution, preserving the data - which inherently isn't a bad thing at all, just a bit annoying. Although one way to avoid it is to ensure that you're mapping any many-to-many relationships (if they exist), but sometimes it is easier to simply remove duplicated rows.

##### Enter your first major lesson in writing SQL:
Always, always, ALWAYS alter a **TEMPORARY** version of the table when you're working in your database connection. Because undo buttons do not exist (although back-end developers will write in safeguards and, chances are, you wouldn't be allowed to alter the production database unless you have a really sloppy back-end dev).

* 3. Okay, let's tidy up! Notice that what I am doing is using the `DISTINCT` statement to remove all duplicated rows from the temp table. Not only that, rather than having to use another `ALTER` statement, I am choosing to just overwrite `patient_info` so then we're done.

```{r, results='hide'}
## Drop the original
dbExecute(con, "DROP TABLE patient_info")

## Remove the duplicates from the temp and make a new patient_info
dbExecute(con, "CREATE TABLE patient_info 
  AS SELECT DISTINCT *
  FROM temp")

## Drop the temporary
dbExecute(con, "DROP TABLE temp")
```

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_info") %>% 
head(10) %>%
  kable(
    format = "html",
    caption = "Table 5. Did we fix the ZIP codes? ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

Yay, we did it! But let's pause to reflect a moment: Didn't this feel like A LOT more work than if we'd just made the dataframe in R and then used the one-liner `gsub()`? Arguably, I think so - but it depends on our ultimate goal. If the goal is to `UPDATE` the database, we would need to do it this way and not just `gsub()` it!

# 3 Descriptives

Next, since we're working in SQL and flying 'blind', we would also like to try to get some sense for what these data contain, if there are any issues we need to take back to the developer for integrity as well as privacy concerns. We start by going through and looking at the tables descriptively.

## 3.1 Geographic representation

```{r}
dbGetQuery(con, 
      "SELECT COUNT(DISTINCT state) AS num_states FROM patient_info")
```

##### Does 89 states seem a little odd to you?
It does to me! Let's see what they are:

```{r, echo = FALSE}
dbGetQuery(con, 
      "SELECT DISTINCT state FROM patient_info") %>% 
  arrange(STATE) %>% 
  head(12) %>% 
  kable(
    format = "html",
    caption = "Table 6. What's going on with the states??!?") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```
#### Question 6 [1 point]:

What is your best guess about what you think has happened? Do you need to include this in your report back to the developer?

EG

> When looking at a state field, the assumption is that there will be approximately 50 unique values (depending on if DC and territories are included). Therefore, seeing 89 values is a red flag. My first thought when seeing abbreviations like AB or BC is that the table is also including Canadian provinces. There are additional codes (ex. AG and BN) that could be data errors or could even be country ISO2 values. My best guess is that this dataset is not confined to just US states, and that these non-US abbreviations would need to be removed. However, this assumption would need to be verified. Information like this must be included in the report back to the developer, so it can be properly investigated and/or verified.

#### Question 7 [1 point]:

Confirm your suspicions by writing a query to print all the ZIP codes in **AB**. Use the `SELECT`, `DISTINCT`, and `WHERE` statements. For example, in SQL, `SELECT zipcodes FROM patient_info WHERE state = 'AL';` would return all of the ZIP codes in the state of Alabama. Look up one of the ZIP codes - does it confirm your suspicions about what happened with the data pull?

EG

Source: https://www.zipcode.com.ng/2022/12/alberta.html

```{r}

dbGetQuery(con, 
      "SELECT DISTINCT zipcode FROM patient_info WHERE state = 'AB'")
```

> This does confirm my suspicions about what happened with the data pull. When querying on the zipcodes in AB, they have the following format: XXX X. According to zipcode.com, zipcodes in Alberta range from T0A to T9X. Therefore, these all look like Canadian zipcodes and that the data is pulling in other countries, not just the US.

#### Question 8 [2 points]:
Decipher, into English, what you think this SQL query is doing.

```{r}
dbGetQuery(con, 
  "SELECT COUNT(DISTINCT pt_key) AS patient_count FROM patient_info WHERE state = 'AL' AND city = 'WETUMPKA' ")
```

DS
> This query is SELECTING a COUNT of how many unique or differing pt_key's are from the patient_info with all locations having AL and city of WETUMPKA. This comes out to a total of 12 patients that are located in Wetumpka, AL

Now write your own! Pick your hometown or your homestate and find out how many patients in the `patient_info` table are from there.

```{r}
# Your code here.
dbGetQuery(con, 
  "SELECT COUNT(DISTINCT pt_key) AS patient_count FROM patient_info WHERE state = 'HI' AND city = 'WAHIAWA' ")
```
DS
> I got a result of three distinct patient_count from Wahiawa, HI.

## 3.2 Age

SQL does not have a lot of the sophisticated statistics we have grown accustomed to having access to in R, but we can execute the basics like max, min, and mean. Let's find the average, max, and minimum age in years of our patients in the `patient_info` table.

```{r}
dbGetQuery(con, "SELECT 
    MAX(age_years) AS max_age,
    MIN(age_years) AS min_age,
    AVG(age_years) AS mean_age
  FROM patient_info")
```
Pretty easy, right? This wasn't so bad!

#### Question 9 [2 points]:

Can you find the max, min, and mean age of all patients who reside in Massachusetts? (**MA**). 

**Hint**: Use the `WHERE` statement! You should get a mean of 41.90843, max of 96.6, and min of 20.5.

DS
```{r}
dbGetQuery(con, "SELECT 
    MAX(age_years) AS max_age,
    MIN(age_years) AS min_age,
    AVG(age_years) AS mean_age
  FROM patient_info
  WHERE state = 'MA' ")
```

## 3.2 Disability Status

This is our primary interest, right? It would be nice to know the counts of folks in each of the disability categories, which we can return in SQL just as we could in R on a dataframe using the `table()` function or `dplyr`'s `count()` or the `n()` of `tidyverse`.

Notice we need to switch to the `patient_disability` table this time. We introduce the `GROUP BY` statement, which works very much like it does in `dplyr` and `tidyverse`! 

```{r}
dbGetQuery(con, 
  "SELECT disability_status,
  COUNT(*) AS patient_count
  FROM patient_disability
  GROUP BY disability_status")
```

#### Question 10 [2 points]:

Now do the same for `EMPLOYMT_STATUS` and `EDUC_LEVEL`, changing the tables as appropriate.

DS
```{r}
dbGetQuery(con, 
  "SELECT EMPLOYMT_STATUS,
  COUNT(*) AS patient_count
  FROM patient_disability
  GROUP BY EMPLOYMT_STATUS")
```

```{r}
#There is not a column for Education Level anymore since it can be considered sensitive information.
```

# 4 Joining data tables in SQL

So much of what you may need to do will involve joining, or merging, tables in some way. You may do this on the fly to produce unique tabulations of data. You may make a new, temporary table that you export to a dataframe to perform analysis upon. We will be doing it for the second purpose.

## 4.1 Left joins

By now in your data science career, the idea of left vs. right, inner vs. full joins are probably starting to feel pretty comfortable. Good! One could argue that **the table join was invented in SQL**, so what better place to practice them?

A `left join` will keep all of the data on the left but only the data that match from the right table. If any data are missing in the right table, they will be dropped. 

Let's perform a `left join` of the `patient_info` and `patient_disability` tables first.
```{r}
dbExecute(con, 
  "CREATE TABLE merged AS SELECT *
  FROM patient_info
  LEFT JOIN patient_disability
  ON patient_info.pt_key = patient_disability.pt_key")
```

Notice the syntax: `ON patient_info.pt_key = patient_disability.pt_key`. We name the left-hand table, then say what key we want to use; then we set that equal to the matching key in the right-hand table.

#### Question 11 [1 point]:

Write a quick query to report back how many rows successfully merged.

```{r, results='show'}
dbExecute(con, 
  "SELECT pt_key,
  COUNT(*) AS patient_count
  FROM patient_info")
```

Any problems to report back to the back-end developer?
> None of the pt_key's from disibility merged with patient_info.

## 4.2 Inner join with aliasing

Next, let's practice an inner join this time using what's called `aliasing` - because old-timey SQL programmers HATED writing out table names! They'd just call them 'a', 'b', or whatever else they felt like. Here, we will use `m` for the `merged` table and `c` for the `patient_clinical` table.

```{r, results = 'hide'}
dbExecute(con, 
  "CREATE TABLE merged2 AS
  SELECT 
    m.pt_key,
    m.ins_key,
    m.age_years,
    m.city,
    m.state,
    m.zipcode,
    m.disability_status, 
    m.employmt_status,
    c.flu_innoc,
    c.tet_innoc,
    c.covid_innoc,
    c.bmi_category,
    c.smoking_status
  FROM merged AS m
  INNER JOIN 
  patient_clinical AS c
  ON m.pt_key = c.pt_key")

## Now remove any duplicates:

## Drop the original
dbExecute(con, "DROP TABLE merged")

## Remove the duplicates from the newly merged file and make it the new merged
dbExecute(con, 
  "CREATE TABLE merged 
  AS SELECT DISTINCT *
  FROM merged2")

## Drop merged2 to tidy up
dbExecute(con, "DROP TABLE merged2")
```

And how did we do?
```{r}
dbGetQuery(con,
           "SELECT COUNT(*) AS row_count FROM merged")
```

Excellent! It looks like we haven't lost any rows in the merging process. Let's see where we are currently at.

```{r, echo = FALSE}
dbGetQuery(con, "SELECT * FROM merged") %>% 
  head() %>% 
  kable(
    format = "html",
    caption = "Table 7. Our beautifully merged table...") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

#### Question 12 [1 point]:

Write either a `LEFT JOIN` or `INNER JOIN` to merge the `merged` table with the `patient_insurance` table. Make sure to join on TWO keys; you will find the `AND` statement helpful for this. Try using aliases!

```{r, results='show'}
dbExecute(con, 
"CREATE TABLE merged2 AS
  SELECT
    m.pt_key,
    m.ins_key,
    m.age_years,
    m.city,
    m.state,
    m.zipcode,
    m.disability_status,
    m.employmt_status,
    m.flu_innoc,
    m.tet_innoc,
    m.covid_innoc,
    m.bmi_category,
    m.smoking_status,
    c.public_ins
  FROM merged AS m
  INNER JOIN
  patient_insurance AS c
  ON m.pt_key = c.pt_key AND m.ins_key = c.ins_key")


# CREATE TABLE merged2 AS
#   SELECT 
#     m.pt_key,
#     m.ins_key,
#     m.age_years,
#     m.city,
#     m.state,
#     m.zipcode,
#     m.disability_status, 
#     m.employmt_status,
#     m.flu_innoc,
#     m.tet_innoc,
#     m.covid_innoc,
#     m.bmi_category,
#     m.smoking_status
#      c.public_ins
#   FROM merged AS m
#   INNER JOIN 
#   patient_clinical AS c
#   ON m.pt_key = c.pt_key
```

```{r, echo = FALSE}
dbGetQuery(con, "SELECT * FROM merged2") %>% 
  head() %>% 
  kable(
    format = "html",
    caption = "Table 7. Our beautifully merged table...") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

Not sure what to do? Load the R script and run it instead to proceed.

```{r}
source("q12_answer.R")
```


#### Question 13 [1 point]:

Any problems to report back to the back-end developer?

> Your answer here.


## 4.3 Joining when there are duplicate keys

There is another table, `patient_binge_drinking`, that was pulled from the backend but there seems to be a problem with it. Notice that there is no `pt_key`!

```{r, echo = FALSE}
dbGetQuery(con, "SELECT * FROM patient_binge_drinking") %>% 
  head() %>% 
  kable(
    format = "html",
    caption = "Table 8. The patient binge-drinking table as pulled from the database.") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

#### Question 14 [1 point]:

Why is it problematic for merging if the patient key is missing and only the insurance key is available?

> Your answer here.


There are really a couple of ways we could deal with this issue. First, and probably most practical, would be to ask our back-end developer to make sure the tables are linked up in the database. Something seems off! (**Hint**: Make sure to give them this recommendation!).

But since we are also under a tight timeline to stakeholders, we are going to do our best to 'plow through' as quickly as we can with what we have. After all, if we remove the duplicated insurance keys entirely, we only lose 81 observations total - not so bad out of 100,000 for a first pass, right?

Let's do it!

## 4.3.1 Using `tidyverse` when we do not want to make any permanent changes to the connection.

I have mentioned this several times, but we are definitely at a point where we do **NOT** want to be sending any changes back to the database. We don't want to permanently delete any data, for one thing!

So, from this point forward, we will lean on our trusty friend from the past few weeks, `dplyr` and `tidyverse`.

* 1. Extract the `merged` table and store it as a dataframe. 

```{r}
merged <- dbGetQuery(con,
                     "SELECT * FROM merged")
```

* 2. Remove the duplicate keys from the `patient_binge_drinking` table. 

```{r}
dbExecute(con, 
  "CREATE TEMPORARY TABLE temp AS
  SELECT DISTINCT ins_key, binge_drinking
  FROM patient_binge_drinking")
```

* 3. Verify that the duplicates were removed.

```{r}
dbGetQuery(con, "SELECT COUNT(*) AS patient_count FROM temp")
```

* 4. Now, extract it to a dataframe we will call `binge`.

```{r}
binge <- dbGetQuery(con, "SELECT* FROM temp")
```


* 5. Join `binge` with the bigger `merged` table we already have. Note that I am doing an `inner_join` - think about why this is! Also note that we are merging on the `ins_key`s rather than the patient keys as before!
**Case matters again!!!**

```{r}
final <- inner_join(merged, binge, by = "INS_KEY", relationship = "many-to-many") %>% 
  distinct(INS_KEY, .keep_all = TRUE)
```

* 6. Verify that our join was successful.

```{r}
str(final)
```

# 4 Aggregation to protect patient privacy

One very important way that we can protect patient privacy is by aggregating across zip codes, regions, states, or any other grouping variable. Given that we often have very small sample sizes (e.g., 12 people with disabilities in Wetumpka, Alabama) it is ostensible that we STILL would not be doing a sufficiently good job at protecting their identity. What can we do?

#### Question 14 [5 points]:

Think about at least two ways we might aggregate the data here to protect patient privacy and conform with HIPAA regulations. Make sure to describe what you would do in detail and why.

> Your answer here.

# 5 Recommendations to the back-end development team

#### Question 15 [10 points]:

Draft a letter or brief report detailing the ways in which you would recommend that the database connection be improved to ensure patient privacy and data integrity. 

* Did you get the data you 'asked' for? (100,000 adult users from the United States)
* Were the data of high quality?
* Were there any problems with keys?
* Should you have had access to the production database? Why or why not?
* Are we in danger of accidentally committing a HIPAA violation?

Make sure to be detailed and specific. You are responsible for safe-guarding our patients' data - just as they are! The job doesn't fall to just one team. So be kind but firm in your recommendations.

> Your answer here.
