---
title: "Project 3: Bioinformatics of Gene Expression"
subtitle: "Merrimack College DSE6630: Healthcare & Life Sciences Analytics"
author: "Elizabeth Greenan"
date: "3 July 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: ../../Module_Demos/Demo_3/references.bib  
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE, # Change to T when knitting
                      cache.comments = F, # Change to T when knitting
                      size = 13)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Turn off scientific notation
options(scipen=999)

# Set seed; add CMRG for parallelization
set.seed(50009, "L'Ecuyer-CMRG")

# Clean, set up, and load
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
rm(list = ls(all = TRUE))

## Note - nestedcv is no longer available on CRAN. Check for and install via:
if (!require("nestedcv", character.only = TRUE)) {
  install.packages(
    "nestedcv",
    repos = c(
      "https://myles-lewis.r-universe.dev",
      "https://cloud.r-project.org"
    )
  )
} else {
  library(nestedcv)
}

pacman::p_load(tidyverse, 
               ggplot2, 
               kableExtra,
               BiocManager,
               DESeq2,
               RColorBrewer,
               gridExtra,
               ggrepel,
               e1071,
               caret,
               randomForest,
               ranger,
               multtest,
               # nestedcv,
               Rfast
)
```

# 1 Background to the Project
## 1.1 Overview
At the end of Demo_3, you hopefully concluded that the SVM was performing better than the Random Forest algorithm, at least OOB. But the biggest problem we saw with both algorithms is sign of __overfitting__: when the testing error was higher than the training error (i.e., testing accuracy < training accuracy). Well, thankfully, we can address this with __cross-validation__ and __tuning the hyperparameters__.

## 1.2 Overfitting
When we optimize the model for the training data, our model will score very well on the training set. We expect this to be true (if it doesn't, then training failed!). However, there may be times when the model in turn is not __generalizable__, meaning that it does not perform nearly as well on the test data set. When a model performs highly on the training set but poorly on the test set, this is known as __overfitting__. We have created a model that knows the training set very well but cannot be applied to new problems. Many textbooks use this analogy of overfitting, and I think it works very well: 
> We all knew a student in school who could memorize information, like answers to a quiz, but failed to apply those concepts in the real-world. That student's mind was overfitted!

Therefore, an overfitted model can look great on the training set but will be useless on other data because it is not generalizable. As mentioned above, we do have two keys ways of helping with this, though.

## 1.3 Cross-validation
You may have encountered $k$-fold CV before, in which case, excellent - you will already have a sense for what we are doing and why. But, if not, let's have a gentle introduction to cross-validation. You are all already familiar with the idea that we split our dataset into a training and a testing set. $k$-fold CV further splits the training set into a $k$ number of subsets also called __"folds"__. The model is fit iteratively $k$ times where each each time the training set contains $k-1$ folds and the model is evaluated on the $k^{th}$ fold. The most commonly used folds on personal computers are $k = 5$ and $k = 10$ for 5% hold-out and 10% hold-out at each iteration, respectively, but you can fit up to $n-1$ folds if you have the computational resources, where $n$ is the number of observations!

## 1.4 Hyperparameter Tuning
What is a __hyperparameter__? Model parameters are those that are learned during training, e.g., the slope and intercept in a linear regression. Hyperparameters are external and not learned, and therefore must be set (or generated by default) by the data scientist before training. In the case of a random forest, hyperparameters include the the number of features considered by each tree when splitting a node for random forest or the kernel in SVM. Since the best hyperparameters are not something we can choose _a priori_, this is why we __tune the hyperparameters__ as we train the model. This is done using trial-and-error, although there are some guidelines can often keep in mind to help ensure that we our models are on a good learning path.

As we (or the package we are using) performs the tuning, we are performing many iterations of the entire cross-validation process, whether that is $k$-fold or another CV method. We then compare all models, select the best one based on a performance metric (typically RMSE for regression or accuracy for classification), and  train it on the full training set. Once we have our best model, we will then evaluate it on the testing set. Thus, even though you may not necessarily see all of the iterations involved, be aware of how extensive this process actually is!

## 1.5 Objectives
We will spend our final class session this term briefly discussing overfitting, cross-validation, and hyperparameter tuning in the context of project planning. From here down, I give you a demo how to tune the Random Forest we fit last week to the VST expression data for the bumblebee chill-coma experiment. **Your objective** will be to apply all the same methods to the SVM and compare and contrast your findings. Specifically, we will compare and contrast **nested cross-validation** versus **$k$-fold cross-validation** to see how well, if at all, we are able to overcome the __P >> n__ problem we discussed in Demo_3. 

## 1.5.1 Your Tasks:

### 1. Answer questions scattered throughout the Random Forest CV and Tuning Sections to ensure your understanding (you may copy these answers, without code, to your project document.)

### 2. Your own novel analysis! Choose from:

  * [Easier] Perform $k$-fold & nested CV, as well as hyperparameter tuning, on SVM a different VST (from the 30-minute experiment only)

  * [Harder] Ready to do go farther? Try performing $k$-fold & nested CV, as well as hyperparameter tuning, on SVM.

## 1.5.2 Housekeeping

### 1.5.2.1 Loading Data
To make your life easier, we will load some pre-prepared datasets from our work in Demo_3. That way, we don't need to spend the computational time running our DESeq and previous machine learning models all over again!

```{r, echo = TRUE, message=FALSE, warning=FALSE}
# Updated file path to match my local
load("../../Module_Demos/Demo_3/vst_all_timepoints.Rdata")  ## These are the all time-points 
                                            ## used for Demo_3 
load("./data/DESeqResults.Rdata")                ## Load the results of the all time 
                                            ## points DESeq analysis (so you 
                                            ## don't have to run it again)
load("./data/OOB_RF_Importance.Rdata")   ## Load the OOB RF importance from 
                                            ## Demo_3 for comparison
load("./data/OOB_SVM_Importance.Rdata")  ## Load the OOB SVM importance from
                                            ## Demo_3 for comparison
```

### 1.5.2.2 Functions
Similarly, I have written some functions to make your life easier. Although I do not require you to take the time to annotate them, it can be great practice in code reading to take a stab at trying to understand exactly *how* these functions do what they do.

#### 1. `doSplits()`
This function quickly and easily performs data partitioning \& pre-processing of the datasets, depending on algorithm chosen. This function could also facilitate automated testing of filtering cutoffs or split-ratios by algorithm, if desired.

```{r, echo=FALSE}
doSplits <- function(vst, algorithm, splitRatio, filterCutoff) {
  ### @vst = vst dataset as extracted from DESeq2
  ### @algorithm = ML algorithm used; currently set up for rf and svm
  ### @splitRatio = the split ratio to employ (training size)
  ### @filterCutoff = the filter cutoff for median number of VST gene counts
  
  ## According to the Valabas et al. (2019) paper, make sure that we are 
  ## filtering in TRAINING set only! 

  # Extract the VST data and transpose
  tVST <- t(assay(vst))
  
  ## We do have gene names, e.g., TRNAV-UAC that are malformed for ranger 
  ## and randomForest. We will fix that before proceeding:
  for (c in 1:ncol(tVST)) {
    colName <- colnames(tVST)[c]
    colName <- gsub("-", "_", colName)
    colName -> colnames(tVST)[c]
  }
  
  ## Add the metadata as columns & merge
  df1 <- cbind(colData(vst)[1], colData(vst)[3], 
               colData(vst)[2])   ## We don't need the size factors
  tVST <- merge(tVST, df1, by = "row.names")

  ## The merge turns back into a dataframe and removes the sample names from 
  ## the rows; let's put them back:
  rownames(tVST) <- tVST[,1]
  tVST <- tVST[,-1]
  
  if(algorithm == "svm") {
    ## Make the factors unordered
    tVST <- tVST %>% 
      mutate_if(is.ordered, factor, ordered = FALSE)
  }
  
  ## Create the data partitions
  ind <- createDataPartition(y = tVST[, c("Treatment")],  ## Treatment evenly 
                                                          ## distributed
                             p = splitRatio,              ## % into training
                             list = FALSE)                ## don't return a list
  train <- tVST[ind, ]
  test <- tVST[-ind,]
  
  ## Calculate row medians of VST gene counts
  medians <- rowMedians(assay(vst))

  ## Filter the features out of train:
  train <- train[, medians > filterCutoff]  
  print(paste0("After filtering, number of genes in the dataset are: ", ncol(train)))

  splits <- list(train, test)
  return(splits)
}
```

#### 2. `findOverlappingGenes()`
Function that compares the `DESeq` results at a given log-fold change (LFC) with the list of important genes from an ML classifier. This could again be iterated over to automate collation of results for more automated comparisons.

```{r, echo = FALSE}
findOverlappingGenes <- function(lfc, important) {
  ### @lfc = the log-fold change cutoff you'd like to employ on the originall DESeq results
  ### @important = the list, df, or matrix that contains the importance values from the ML classifier; make sure it is already filtered if needed.

  res <- resultsDESeq %>% 
    as.data.frame() %>% 
    filter(abs(log2FoldChange) >= lfc)   # Make sure to filter by the ABSOLUTE VALUE :)
  
  # Move the rownames (genes) back to a column
  res$geneID <- rownames(res)
  # Coerce to a dataframe, if needed
  important <- important %>% 
    as.data.frame() %>% 
    filter()
  # Move the rownames (genes) back to a column, if needed
  if (!"geneID" %in% colnames(important)) {
      important$geneID <- rownames(important)
  }
  #Perform an inner join to find the overlap
  overlap <- inner_join(res, important, by = "geneID")
  
  return(overlap)
}
```

#### 3. `compareConfusion()`
This function compares the results of confusion matrices by printing them in a table side-by-side. You could also use this to graph your results, if desired.

```{r, echo=FALSE}
compareConfusion <- function(confusionList) {
  ## instantiate
  finalDF <- data.frame()
  for(i in 1:length(confusionList)) {
    ## The first one
    if(i == 1) {
      confMat <- confusionList[[i]]   ## grab the first one
      df <- confMat$overall %>% as.data.frame() 
      finalDF <- rownames(df) %>% as.data.frame()
      colnames(finalDF)[1] <- "Metric"
      finalDF$`Confusion Matrix 1`  <- df[, 1]       ## grab the value
    }
    if(i > 1) {
      name <- paste0('Confusion Matrix ', i)
      confMat <- confusionList[[i]]
      df <- confMat$overall %>% as.data.frame()
      finalDF[, name] <- df[, 1]       ## grab the value
    }
  }
  return(finalDF)
}
```

# 2 Tuning the Random Forest Model

One of the reasons that random forest is generally considered to work so well OOB is that it picks up a portion of the predictors for training. At each step, it randomizes the variable selection during each tree split. Thus, it is doing its own form of cross-validation internally as part of the algorithm. This is also even the OOB is considered to be less prone to overfitting as compared to other machine learning algorithms. 

That said, we can explicitly apply an external cross-validation too - **because even models that are more robust to overfitting are STILL capable of being overfitted.** 

More importantly, though, there are certain __hyperparameters__ of random forests that will assist with overfitting much better than cross-validation given the internal split-variable randomization of the algorithm. 

**Note**: One of the other advantages to RF that you may come across is that it is ***more*** robust to multicollinearity. Why? Well, since the algorithm is randomly selecting a certain number of samples via bootstrapping to train on and another random sample of features at each split, then a more diverse set of trees is produced. It tends to lessen correlation among the trees and would have the potential to 'break up' multicollinearity in the process. However, being "more" robust is not the same thing as being perfectly robust! It is worth thinking this through if, for example, we expect similar **kinds** of genes to be expressed in response to chill coma. 

## 2.1 Random Forest Hyperparameters

### 2.1.1 Number of features at any given split.
Each time a split is to be performed, the search for the split variable is limited to a random subset of the $p$ features. In R's packages `randomForest` and `ranger`, this hyperparameter is called $m_{try}$ and is arguably one of the single most important hyperparameters to tune. By default, for regression RF, $m_{try} = \frac{p}{3}$ and for classification RF, $m_{try} = \sqrt{p}$. 

But let's think about that: our OOB random forest, by default, would have run an $m_{try} = \sqrt{12927}$ or $approx$ `r sqrt(12927)`. Thus, $m_{try}$ is often one of the first hyperparameters you should consider tuning for random forest.

### 2.1.2 Number of trees in the forest.
**How many trees should you have in your forest?** Although this isn't technically a hyperparameter, if you do not have enough trees in your forest, you will not be able for your model to **converge**, i.e., stabilize the error rate. Note that one of the cross-validation packages we will use, `train()` from the `caret` package does not allow us to tune __num.trees__. I will be showing you a work-around for that.

Also, note that [Hands-on Machine Learning in R](https://bradleyboehmke.github.io/HOML/random-forest.html) recommends starting with $num.trees = 10 \cdot p$, but the challenge for us is __how massive the computational time would be__. DO NOT ATTEMPT THIS! By these recommendations, you would set an $num.trees =$ `r 10*12927`, which would only crash your computer!

**Instead, I am going to recommend an** $num.trees \leq 2000$ **unless you have wayyyy more computing power than I do at the moment ;).**

### 2.1.3 Complexity of each tree, aka, node split.
Whereas the number of features at any given split ($m_{try}$) allows us to control the learning of the algorithm, the node split allows us to control the complexity of the individual trees in the forest. The default values are typically 1 and 5 for classification and regression, respectively, and while the default can work well on many datasets, noisier datasets or those that require higher values of $m_{try}$ may perform better with **increased** values of node size. Increasing values of node size **decreases** complexity and thereby tree depth. Further, larger node sizes can often speed up computation without large sacrifices to error rates.
  
### 2.1.4 Splitting rule to use during tree construction.
The main rules we will be using are the __gini__ rule and the __extratrees__ rule. We will practice tuning these rules to see what effect, if any, they can have on forest construction and overall accuracy.

#### **Question 1**: Look up these rules and try to explain what they are doing for yourself.

Source:
“Gini Impurity and Entropy in Decision Tree.” Geeks for Geeks, 8 Nov. 2025, www.geeksforgeeks.org/machine-learning/gini-impurity-and-entropy-in-decision-tree-ml/.

Additional source: Referenced Module 6 PPT. from DSE6111

> The Gini rule is also known as Gini impurity. This metric selects the split that reduces impurity the most. When there is a gini impurity of 0, that means that the node is pure and that everything is in the same class. It is used when determining how well the data should be split into branches in a random forest model. According to Geeks for Geeks, “Gini Impurity checks how often a randomly selected sample would be mislabeled if assigned by class probability” (Gini Impurity and Entropy in Decision Tree). This metric is used in algorithms such as CART (Classification and Regression Trees) to determine splits.

Source: “Extratreesclassifier — Scikit-Learn 1.9.0 Documentation.” Scikit Learn, scikit-learn.org/stable/modules/generated/sklearn.ensemble.ExtraTreesClassifier.html. Accessed 27 June 2026.

Source: “How Extra Trees Classification and Regression Algorithm Works.” ESRI, doc.esri.com/en/arcgis-pro/latest/tool-reference/geoai/how-extra-tree-classification-and-regression-works.html. Accessed 27 June 2026.

> The extratrees (also known as extremely randomized trees) rule is based on randomized decision trees. According to ESRI, it is similar to the random forest algorithm, since it creates multiple decision trees. ESRI also noted that it is often faster than random forest. However, each split is selected randomly. This is different from other models, where the optimal splits are calculated. A benefit of extratrees is that it helps reduce overfitting in the model.

### Sampling scheme.
This would allow us to alter whether we are sampling features with or without replacement at each node split. We will not be tuning this at this time.


## 2.2 Pre-processing and feature selection
Two important elements of the data science life cycle that we did not have a chance to get into yet last week in Assignment 1 was __pre-processing__ (which includes transformations, encoding, splitting, and normalizing) and __feature selection__ (which includes reduction of the number of features or dimensionality of the dataset). We will spend more time on both of these topics throughout the course, but for this project it is largely hidden in the function that I wrote for us above, `doSplits()`. 

#### **Question 2**: Take a moment to go look at the function and the documentation I have written. Also try playing with the parameters a bit - how many more or fewer genes do you get when you lower or raise the `filterCutoff`, respectively?

```{r}

splits <- doSplits(vst = vsData, 
                   algorithm = "rf", 
                   splitRatio = 0.8, 
                   filterCutoff = 10) # Tested 1, 2, and 10
train <- splits[[1]]
test <- splits[[2]]

```

> With the original filterCutoff value of 5, there were 7,676 genes in the dataset after filtering. Keeping everything else the same, I then decreased the filterCutoff, testing it with both values 1 and 2. This caused the number of genes in the dataset to increase to 12,927 in both cases. This showed that a lower filterCutoff filters fewer genes out. It was interesting that the number of genes did not change between filterCutoff values 1 and 2. This showed that in this case, the slight change in the filterCutoff value did not impact the number of genes being filtered out.

> Next, I increased the filterCutoff to 10. There were 765 genes in the dataset after filtering. This showed that a higher filterCutoff value filters more genes out.

Notice in the comments that I am only doing feature selection on the training data set. Why? Current research strongly suggests that filtering on the entire dataset has the potential to generate bias, not surprisingly, which in turn can create __data leakage__. The advice of Vabalas et al. (2019) is to only do our feature selection **only on the training model**.

#### **Question 3a**: Do some research on __data leakage__ and try to explain what that means in this context. Why do we care about data leakage?

> Data leakage happens when data that should not be used during model training is used. An example of this is if the test data is used during training. Ensuring data leakage does not happen is important for various reasons. When there is data leakage, the model will initially look strong and will likely have high performing metrics such as accuracy, recall, and precision. However, once the model is run on unseen data, it will not perform nearly as well.

> In the context of Project 3, this is why we only performed feature selection on the training set. When we move onto cross-validation and additional tuning in later sections, that will be performed on the training set. This ensures that the model results are not inflated during training. 

Now, we will set `filterCutoff = 5` and partition data for random forest tuning.
```{r}
splits <- doSplits(vst = vsData, 
                   algorithm = "rf", 
                   splitRatio = 0.8, 
                   filterCutoff = 5)   
train <- splits[[1]]
test <- splits[[2]]
```

#### **Question 3b**: Using the annotation I have provided in the function `doSplits()`, explain the parameters chosen and try to explain ***why*** you think they were chosen, based on what we did in Demo_3. In particular, make sure to explain WHY a filterCutoff of 5 was chosen. (**Hint**: In Demo_3, did we do anything to adjust for genes with low variance in expression?)

> The vst = vsData parameter represents the data used in the split. Algorithm = "rf" means that the algorithm used is a random forest model. The split ratio (splitRatio) = 0.8 represents a training and testing split of 80% and 20% respectively. This is further annotated in the train and test code above.

> In Demo 3, we identified genes with low variance in expression, but did not filter these genes out. In this case, low variance in expression was represented by the chill coma the bumblebees were in. Filtering these genes out would have taken away valuable data from when the bees were in their coma, negatively impacting the benchmarking study. In Project 3, based on the code above, we do want to filter out some of these genes.

> In this project, after filtering with a filterCutoff = 5, there are 7,676 genes in the dataset. I think a value of 5 was chosen because it removes a signficiant number of low expression genes from the dataset, without removing too many. Based on my experimentation of the filterCutoff values in Question 2, I saw that if the filterCutoff value was too low, not enough genes would be filtered out. If the filterCutoff value was too high, too many genes were filtered out. Choosing a value of 5 allows for a sufficient number of genes to remain, without keeping too many, which could create too much noise.

## 2.3 Apply $k$-fold Cross-Validation (CV) to Random Forest

### 2.3.1. Leverage the `trainControl()` and `train()` functions from the `caret` package. 

Recall that when were doing OOB random forest in Demo_3, we didn't use `trainControl()`. That will be a deviation from what we do this time. 

However let's first fit the same RF we fit last week for comparison:

```{r, echo = TRUE, eval=T}
rfOOB <- randomForest::randomForest(
  Treatment ~ ., 
  data = train)

pred.test.rf <- predict(rfOOB, test, type = "response")
confMat <- confusionMatrix(pred.test.rf, test$Treatment)
```

### 2.3.2. Employ a $k$-fold CV where $k = 10$.

Set the control for $k = 10$ and then run the RF:
```{r, eval = T}
# Set the CV arguments
kFoldCtrl <- trainControl(method = "cv",    ## for k-fold CV
                          number = 10)      ## k
```

And fit the Random Forest model:
```{r}
rfCV <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl)    ## added in the 10-fold CV
```

#### **Question 4**: Explain, in your own words, what a 10-fold CV does here. Is it vital to perform with a random forest? Why or why not?

> A 10-fold cross-validation (CV) splits the training data into 10 different folds and has a 10% hold-out at each iteration. This allows the model to be trained 10 times. 1 fold will be used for testing, while the remaining 9 folds will be used for training. This method often provides better model performance metrics. While not vital to perform with a random forest, it is often used to enhance the model and better understand the results. It can be beneficial to use, as it allows us to evaluate the stability of the model and also provides insight on how the model will perform on unseen test data.

```{r, echo = FALSE, warning=FALSE, message=FALSE, eval=T, fig.cap="Figure 1. Visualize the 10-fold CV Performance"}
ggplot(rfCV, highlight = TRUE) +
  ggtitle("Random Forest Performance After 10-fold CV") + 
  theme_bw()
```

You'll notice that by default with the 10-fold CV is that **it is doing automated hyperparameter tuning too!** Also, note the x-axis here. It is actually the $m_{try}$ hyperparameter, the number of randomly selected predictors $P$ at each split. This allows us to **visualize the effect of** $m_{try}$ **on accuracy.**

#### **Question 5**: What combination of hyperparameters gives the highest accuracy in the **training model**? (Use plot above to answer.)

> The following combination of hyperparameters gives the highest accuracy in the training model: the gini splitting rule with approximately 7,800 randomly selected predictors (mtry). This combination gives the highest accuracy of approximately 0.63 (63%). Although both gini and extratrees use approximately the same number of randomly selected predictors, the extratrees splitting rule performs worse, with the highest accuracy at approximately 0.62 (62%).

### 2.3.3. Extract best model.

You can also extract the best model as well to get the precise combination of hyperparameters.

```{r, echo = F}
rfCV$bestTune %>% 
kable(
    format = "html",
    caption = "Table 1. Hyperparameters of the best Random Forest Model using 10-fold CV") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### **Question 6**: Explain, in your own words, what the values of these hyperparameters mean.

> Hyperparameters allow us to tune the model with the goal of improving performance. In this case, there are two splitting rules: gini and extratrees. These determine how the splits are chosen. Here, they are chosen using either Gini impurity, which measures node impurity or extratrees, which adds randomness to the splits.

> Additional hyperparameters include the number of features at any given split (mtry) and the number of trees in the forest (num.trees). The mtry hyperparameter is an important hyperparameter and allows us to try and improve the model by increasing or decreasing the number of features. This can often help reduce overfitting in a dataset. Similarly, adjusting the number of trees in the forest can help stabilize the model.

### 2.3.4. Assess for model stability.

Let's look at the fold-specific results for our best model. This is essential to assess **stability** of the model. 

```{r, echo=FALSE}
rfCV$resample %>% 
kable(
    format = "html",
    caption = "Table 2. Fold-specific accuracy for the 10-fold CV Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### **Question 7**: What is the average, minimum, and maximum accuracy values across the ten folds? What does this tell you about the **stability** of the random forest? Is this concerning? Make sure to discuss reasons why there may be instability, if that's the conclusion you draw.

**Hint**: Think about these results relative to our __sample size__. Would we expect instability in resampling from, among other issues, small sample size alone in this case?

```{r}
# Create accuracy vector
accuracy <- c(0.6666667, 0.6666667, 0.8333333, 0.1666667, 0.5000000, 1.0000000, 0.5000000, 0.3333333, 0.8333333, 0.8571429)

# Calculate average accuracy
average_accuracy <- mean(accuracy)
# Calculate minimum accuracy
minimum_accuracy <- min(accuracy)
# Calculate maximum accuracy
maximum_accuracy <- max(accuracy)

average_accuracy
minimum_accuracy
maximum_accuracy

```

```{r}
# Check training and testing split to answer the question
nrow(train)
nrow(test)
```

> The average accuracy is 0.6357143 (63.57%). The minimum accuracy is 0.1666 (16.66%), and the maximum accuracy is 1 (100%). The train/test split is 60 and 40, respectively. Since this is a 10-fold cross-validation being run on the training set (60 observations), there are approximately 6 observations per fold. This is a very small sample size, which can lead to a high variance in the results. Therefore, the accuracy can fluctuate easily if the data changes even slightly.

> We would expect some instability in resampling from the small sample size in this case. Therefore, this does not necessarily mean there is an issue with the model, but that the model is sensitive. This is describing the big p little n issue that we have discussed in class. This issue makes it harder to obtain useful information from the dataset, since we have such a large number of predictors (p) and a small number of observations (n). 

### 2.3.5. Fit the test data to the best tuned model from the 10-fold CV.
Regardless of the decision you made in Question 7, let's proceed with assessing model performance.

```{r, echo=TRUE}
rfCV <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl, 
               tuneGrid = rfCV$bestTune)   ## Add in the results of the CV 
                                           ## and auto tuning

pred.test.rf <- predict(rfCV, test, type = "raw")  ## type is now 'raw' 
## Store the confusion matrix for later comparison
confMatCV <- confusionMatrix(pred.test.rf, test$Treatment)
```

### 2.3.6. Compare the two confusion matrices.

Lastly, let's compare the results of what we just ran (stored in `confMatCV`) with what we the OOB performance from Demo_3 (stored in `confMat`).

```{r, echo = FALSE}
compareConfusion(confusionList = list(confMat, confMatCV)) %>% 
  kable(
    format = "html",
    caption = "Table 2. Comparing Accuracy - Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### **Question 8**: Let's take a moment to unpack what we are looking at here. What do each of these metrics measure? **Hint**: Make sure to look at the documentation for the `confusionMatrix()` function from `caret`.

Sources:

Ariza-Lopez, Francisco Javier, et al. “Package ‘ConfMatrix.’” CRAN R, 30 May 2026, cran.r-project.org/web/packages/ConfMatrix/ConfMatrix.pdf.

“confusionMatrix: Create a Confusion Matrix.” Rdocumentation, www.rdocumentation.org/packages/caret/versions/7.0-1/topics/confusionMatrix. Accessed 27 June 2026.

“ConfusionTableR.” CRAN R, cran.r-project.org/web/packages/ConfusionTableR/vignettes/ConfusionTableR.html. Accessed 27 June 2026. 

I used the caret::confusionMatrix document through the Help Page in R.

> __Accuracy__: The Accuracy metric measures how correct the model is. This value can be calculated from the values in a confusion matrix. The calulaction is as follows: the sum of true positives and true negatives, divided by the sum of true positives, true negatives, false positives, and false negatives. Confusion Matrix 1 (OOB model) has an accuracy of 0.6428 (64.28%), while Confusion Matrix 2 (10-Fold CV model) has a higher accuracy of 0.7857 (78.57%).

> __Kappa__: The Kappa metric measures the relationship between the observed proportion and the expected proportion (Ariza-Lopez, Francisco Javier, et al.). This provides insight on model performance and can be used in imbalanced data. Confusion Matrix 1 (OOB model) has a kappa of 0.2857 (28.57%), while Confusion Matrix 2 (10-Fold CV model) has a kappa of 0.5714 (57.14%).

> __Accuracy Lower Bound__: The AccuracyLower metric represents the lower bound of the confidence interval for accuracy. As noted in the documentation, the 95% confidence interval is also calculated. Confusion Matrix 1 (OOB model) has an AccuracyLower of 0.3513 (35.13%), while Confusion Matrix 2 (10-Fold CV model) has an AccuracyLower of 0.4920 (49.20%).

> __Accuracy Upper Bound__: The AccuracyUpper metric represents the upper bound of the confidence interval for accuracy. As noted in the documentation, the 95% confidence interval is also calculated. Confusion Matrix 1 (OOB model) has an AccuracyUpper of 0.8724 (87.24%), while Confusion Matrix 2 (10-Fold CV model) has an AccuracyUpper of 0.9534 (95.34%).

> __Accuracy Null__: The AccuracyNull metric represents the "no information rate", which is the largest class percentage in the data. Therefore, the AccuracyNull provides a baseline value for model performance. Both Confusion Matrix 1 (OOB model) and Confusion Matrix 2 (10-Fold CV model) have an AccuracyNull of 0.50 (50%).

> __Accuracy p-value__: The AccuracyPValue metric represents the one-sided test that is used to determine if the accuracy is better than the "no information rate". Confusion Matrix 1 (OOB model) has an AccuracyPValue of 0.2119 (21.19%), while Confusion Matrix 2 (10-Fold CV model) has an AccuracyPValue of 0.0286 (2.86%). Since this is a p-value test, the significance level of 0.05 is used to determine whether the answer is statistically significant. Therefore, we fail to reject the null hypothesis for the OOB model and reject the null hypothesis for the 10-Fold CV model.

> __McNemar p-value__: The McnemarPValue metric represents the p-value that is calculated from the McNemar's test. As mentioned in the documentation, this tests for symmetry of rows and columns in a two-dimensional table. Confusion Matrix 1 (OOB model) has a McnemarPValue of 1 (100%), while Confusion Matrix 2 (10-Fold CV model) has a McnemarPValue of 0.2482 (24.82%). Again, since this is a p-value test, the significance level of 0.05 is used to determine whether the answer is statistically significant. Since both of the p-values are > 0.05, they are not statistically significant. Therefore, we fail to reject the null hypothesis.


#### **Question 9**: What was the impact, if any, of the pre-processing we performed here? What evidence do you have? Lastly, what was the biggest change, other than tuning, that we made as compared to the OOB RF?

> One of the pre-processing steps was using the filterCutoff to filter out genes. This was a form of feature reduction, as the filterCutoff value of 5 that we used reduced the number of predictors from over 12,000 to 7,676. This helped reduce noise and overfitting in the dataset.

> The biggest change, other than tuning, that we made was switching from out-of-bag (OOB) to a 10-fold cross-validation. While the OOB method provided a single result, the 10-fold cross-validation method provided us specific accuracy and kappa values for each of the folds. This allowed us to obtain fold-specific accuracy, which helps us better understand model performance. The results in Table 2 show that by switching to the 10-fold cross-validation model, some of the performance metrics improved. This highlights that tuning can improve model performance.

#### **Question 10**: What are your conclusions, at this stage, about the performance of the 10-fold CV, tuning, and applying the filtering cutoff relative to the OOB performance from Demo_3?

> The performance of the 10-fold CV, tuning, and the application of the filtering cutoff allowed us to get a better understanding of model stability and accuracy. In Demo_3, the OOB performance on the training data had very high performance, including an accuracy and kappa of 1. This highlights a potential overfitting issue with the data, as results this strong are unlikely. It was also noted that the McnemarPValue was NaN.

> The average accuracy of the 10-fold CV was 0.6357 (63.57%). Although this is still a lower accuracy than we would want for a model, it shows a much more realistic performance. We had a minimum accuracy of 0.1666 (16.66%) and a maximum accuracy of 1 (100%). The accuracy of 1 was located in Fold07.

> Overall, these results are showing that we continue to deal with the big p little n situation, which can impact the results of cross-validation, as we have a large number of predictors and a small number of observations. We saw this in the accuracy results, which had a lot of variability.

## 2.4 Further Tuning: Searching

Hyperparameter tuning is akin to looking for a needle in a haystack. Each model we fit will have its own set of unique hyperparameters, and thus they are not something that you can "guess" or "estimate". Thus, the only approach to narrow our search is to evaluate a wide range of values for each hyperparameter (kind of like setting fire to the haystack, computationally speaking). We can allow packages like `caret` to automatically search for us (as we just did), but if the goal is to generate the most exhaustive search of hyperparameters - and in a way that is agnostic to human error - we need to do a __search__. Searches come in several flavors, but we will be executing a __grid__ search vs. a __random__ search for our study because both are quite simple to execute with the packages we already have.

### 2.4.1 Grid Search
As the name suggests, grid searches take a matrix of values and will iteratively and linearly search through the grid until all combinations have been tested. This is also sometimes referred to as a 'Cartesian grid' search. 

##### How many searches to do?
You will always be limited by time \& computational resources. As much as I would like this to be exhaustive, for our purposes we will be keeping this smaller and more manageable since we are not running it on a computing cluster or in parallel.

##### A Note about Randomness
You may have already noticed that you get different results from your teammates or even if you run it yourself over and over. Why? By default, `caret` is setting a random seed for each split, and so you could have a different set of seeds than someone else (or even versus when you run it again). Wait, that's not terribly helpful, is it? 

It is and it isn't. You might imagine times when you just want to be able to run a model, check the fit, and move on. Other times, like this time, we need something highly reproducible. `caret` will allow us to set our seeds so that we can force it to be more reproducible. Once we have some seeds to add, we update the `kFoldctrl` object.

**NOTE**: When you strike out on your own, you may get a 'bad seeds' error. This is often because you need to increase the size of your sample space.
```{r, echo = FALSE}
# Make a set of k = 10 seeds for reproducibility
seeds <- vector(mode = "list", length = 11)
for(i in 1:10) {
  seeds[[i]]<- sample.int(n=1000, 54)   # Increase 54 if you have a larger grid! 
}
# For the last model
seeds[[11]]<-sample.int(1000, 1)
```

```{r, echo = FALSE}
# Set the CV arguments
kFoldCtrl <- trainControl(method = "cv",    # for k-fold CV
                          number = 10,      # k
                          seeds = seeds)    # sets the seeds, one for each split

```


#### 2.4.1.1. Set the search grid.
```{r}
searchGrid <- expand.grid(
  mtry = floor(ncol(train) * c(.05, .15, .25, .35, .45)),
  splitrule = c("gini", "extratrees"),
  min.node.size = c(1, 3, 5, 10) 
)
```

#### **Question 11**: How many different combinations are we going to search with this grid?

> In the above code chunk, there are 5 mtry values: .05, .15, .25, .35, and .45. There are 2 splitrule values: gini and extratrees. Lastly, there are 4 min.node.size values: 1, 3, 5, and 10.

```{r}
# Calculate the number of different combinations we will search with this grid
5 * 2 * 4
```

> We will search 40 different combinations with this grid.

#### 2.4.1.2. Run the RF, this time adding in the search grid.

Be patient. This may take a few minutes to run.

```{r}
rfTuned <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl, 
               tuneGrid = searchGrid # Add in the search grid
)
```

```{r, echo = FALSE, fig.cap="Figure 2. Visualize the Grid Search Performance"}
ggplot(rfTuned, highlight = TRUE) +
  ggtitle("Random Forest Performance Grid Search Tuning") + 
  theme_bw()
```

#### **Question 12**: What combination of hyperparameters gives the highest accuracy in the **training model**? (Use plot above to answer.)

> The gini splitrule provides the highest accuracy in the training model. There are approximately 1,200 randomly selected predictors (mtry) and the minimal node size is 5. With this combination of hyperparameters, the model achieves an accuracy of approximately 0.74 (74%). The plots also show that the gini splitrule outperforms extratrees.

#### 2.4.1.3. Take a look at the hyperparameters of the best model.

```{r}
rfTuned$bestTune %>% 
kable(
    format = "html",
    caption = "Table 3. Results of the Grid Search on Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### 2.4.1.4. Fit the best model's hyperparameters to the test data.

```{r}
rfTuned <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl, 
               tuneGrid = rfTuned$bestTune)   ## Use the best model
```

#### 2.4.1.5. Make the new predictions to assess accuracy.

```{r}
pred.test.rf <- predict(rfTuned, test, type = "raw")  ## type is now 'raw' 
# Store the confusion matrix
confMatTuned <- confusionMatrix(pred.test.rf, test$Treatment)
```

#### 2.4.1.6. Compare the three confusion matrices.

```{r}
compareConfusion(confusionList = list(confMat, confMatCV, confMatTuned)) %>% 
    kable(
    format = "html",
    caption = "Table 4. Comparing Accuracy - Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### **Question 13**: What impact did hyperparameter tuning with the grid search have on model performance compared to both the OOB and 10-fold CV with auto-tuning?

> Table 4 shows that the hyperparameter tuning with grid search had a slight impact on model performance when compared to both the OOB and 10-fold CV with auto-tuning. The hyperparameter tuning with grid search did not have the highest accuracy. Instead, it had the second highest, of 0.7142 (71.42%). Overall, it performed worse than the 10-Fold Cross-Validation model, but better than the OOB model. This shows that tuning the model generally yields better results than the OOB model. However, the cross-validation model still outperformed the grid search.


**Note on methodology**:
As I mentioned above, there is no parameter for the number of trees that we could tune in the search grid. Why not? It's because we used the `train()` function from the `caret` package. `train()` is what is called a __wrapper function__, which means that we may be somewhat stuck using what's available to us in the function. But what if we want to tune hyperparameters that are available through that wrapper function? We would need to switch our tactics. We could either write a __custom function or loop__ to do it, or we could modify the function in `caret`. Wait, we can do that?! Yep!

#### **Question 14**: Use the `modelLookup` function to help you figure out what hyperparameters are available for the `rf` vs. `ranger` methods for performing random forest algorithm with `caret`. What are the key differences you observce in the hypers you can tune?

```{r}
# Uncommented to run
 modelLookup("ranger")
 modelLookup("rf")
```

> The modelLookup function showed that there are different hyperparameters that can be tuned for ranger and rf. The ranger method allows the following parameters to be tuned: mtry (number of randomly selected predictors), splitrule (in this project, gini or extratrees), and min.node.size (minimum number of observations in a node). The rf method only allows mtry to be tuned and is very limited.

### 2.4.2 Grid Search Using a Customized `caret` Method

#### 2.4.2.1. Write a customized `caret` method for random forest using the `ranger` method.

```{r, echo = FALSE}
makeCustomMethod <- function(paramList, methodName) {
  # Create a custom list, for a classification model and the ranger method. This list does not use loops.
  custom <- list(type = "Classification", 
                 library = "ranger", 
                 loop = NULL)
  # Create a dataframe that defines the hyperparameters that can be tuned, assign each of them to a class, and label them
  custom$parameters <- data.frame(parameter = paramList, 
                                  class = rep("numeric", length(paramList)), 
                                  label = paramList)
    # Create a function for the grid search
    custom$grid <- function(x, y, len = NULL, search = "grid") {}
    # Fit the model using the hyperparameter values in the grid search function
    custom$fit <- function(x, y, wts, param, lev, last, weights, classProbs) {
        # Fit the random forest model and specify mtry and num.trees values
        randomForest(x, y, 
               mtry = param$mtry,    ## Need to update to make dynamic
               num.trees = param$num.trees)
  }
  # Create a function for model predictions
  custom$predict <- function(modelFit, newdata, preProc = NULL, submodels = NULL)
  # Create the predictions from the model
  predict(modelFit, newdata)

  # Create a function for model probabilities
  custom$prob <- function(modelFit, newdata, preProc = NULL, submodels = NULL)
  # Create the probabilities from the model
  predict(modelFit, newdata, type = "prob")

  # Sort the results of the hyperparameter tuning by the first column
  custom$sort <- function(x) x[order(x[,1]),]

  # Return the class levels for the classification model
  custom$levels <- function(x) x$classes
return(custom)
}
```

#### **Question 15**: Add comments above, as best you can, to our custom function. In your own words, what exactly does this function do? Why are we writing a customized function? (**Hint**: I told you above!)

> The customized function above creates a caret function for the random forest model using the ranger method. This was written because as we saw in Question 14, not all methods have the same hyperparameters that can be tuned. Therefore, by writing a customized function, we can use it to tune the hyperparameters we want, not just the default options. Comments were added above.

#### 2.4.2.2. Run the customized method.
**NOTE**: This takes about 12 min to run on 16gb of RAM. This is exactly why we are doing caching for this assignment. Otherwise, your report will take forever to re-knit. 

```{r}
## Run the custom method setting the hyperparameters we want to use
customRF <- makeCustomMethod(paramList = c("mtry", 
                                           "num.trees", 
                                           "min.node.size",
                                           "splitrule"),
                             methodName = "ranger")

## Now expand the search grid
searchGrid <- expand.grid(.mtry = floor(ncol(train) * c(.05, .15, .25)),
                        .num.trees = c(50, 200), 
                        .min.node.size = c(1, 5, 10, 20),
                        .splitrule = "gini"
                        )
## Now run the RF with the custom tuned hypers and the new search grid
customTuned <- train(y = train$Treatment, x = train, 
                method = customRF, 
                metric = "Accuracy", 
                tuneGrid = searchGrid, 
                trControl = kFoldCtrl)
```

#### **Question 16**: Did you catch that I am setting `.splitrule = gini` explicitly? Why do you think I made this choice?

> I think you set .splitrule = gini because so far, each time we have compared results, the gini method has outperformed the extratrees method. This occured in both cross-validation and grid search tuning. Since it is the better performing splitrule, it makes sense that we would want to use it in the custom function. Additionally, unlike extratrees, it does not perform randomized splits.

```{r, echo=FALSE, fig.cap = "Figure 3. Visualize the Customized Grid Search Performance"}
ggplot(customTuned, highlight = TRUE) +
  ggtitle("Customized RF Performance Grid Search Tuning") + 
  theme_bw()
```

#### **Question 17**: What combination of hyperparameters gives the highest accuracy in the **training model**? (Use plot above to answer.) How does it compare to previous results?

> The following combination of hyperparameters gives the highest accuracy in the training model: num.trees = 50, mtry = 1,919, and min.node.size = 10. This combination of hyperaparameters reaches an accuracy of approximately 1 (100%). This shows an almost perfect accuracy, which is unlikely.

> For all three models (OOB, 10-Fold CV, and Grid Search), the gini rule provided the best results. For OOB, the accuracy was 0.6428 (64.28%). For 10-Fold Cross-Validation, the accuracy was 0.7857 (78.57%). For the grid search, the accuracy was 0.7142 (71.42%).

#### 2.4.2.3. Take a look at the hyperparameters of the best model.
```{r}
customTuned$bestTune %>% 
kable(
    format = "html",
    caption = "Table 5. Results of the Customized Grid Search on Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### 2.4.2.4. Refit the best model's hyperparameters to the test data set.
```{r, warning = FALSE, message=FALSE}
newTuned <- train(Treatment ~.,  
               data = train,
               method = customRF,
               trControl = kFoldCtrl, 
               tuneGrid = customTuned$bestTune)
```

#### 2.4.2.5. Make the new predictions to assess accuracy.
```{r}
pred.test.rf <- predict(newTuned, test, type = "raw")  
# Store the confusion matrix
confMatTuned2 <- confusionMatrix(pred.test.rf, test$Treatment)
```

#### 2.4.2.6. Compare the four confusion matrices.
```{r, echo = FALSE}
compareConfusion(confusionList = list(confMat, confMatCV, confMatTuned, confMatTuned2)) %>% 
    kable(
    format = "html",
    caption = "Table 6. Comparing Accuracy - Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### **Question 18**: How well did hyperparameter tuning with the number of trees and the new grid search perform? Are we seeing sufficient improvement to continue down this path? 

**Hint**: Make sure to check model stability again as well by looking at the performance across folds, even though I didn't do that for you. 

> The hyperparameter tuning with the number of trees and the new grid search (Confusion Matrix 4) showed improvement over the OOB model (Confusion Matrix 1) and the original grid search model (Confusion Matrix 3). It did not perform better than the 10-Fold CV model (Confusion Matrix 2). In fact, they had identical performance metrics. Both Confusion Matrix 2 and Confusion Matrix 4 had an accuracy of 0.7857 (78.57%).


```{r}
# Check model stability across folds
customTuned$resample
```

> The above code chunk checks the model stability again by looking at performance across folds. It provides an interesting result: across all 10 folds, both the accuracy and kappa are 1. This would indicate near-perfect model performance, which is unlikely and is not shown in our current performance metrics. These results come back to the big p little n scenario we have with this data, since there are many predictors, but not many observations.

> These results are not improvement enough to make it worth continuing down this path. It is noteworthy that this latest model took the longest to run, yet it did not have any improved metrics from the 10-Fold CV model. In a production environment, this would increase both time and cost. Therefore, since it isn't yielding improved results or potential promise, it is not something to continue working on.

### 2.4.3 Random Search

A random search is where we randomly choose hyperparameters and, based on an initial set of random values, iteratively search until we get closer and closer to the best fit. Now, at this point we might also consider performing a random search in addition to the grid search; however, as we can see that we are generally STILL seriously under-performing, let's table that idea and put our efforts somewhere we will hopefully see a bigger impact: __Nested Cross-Validation__.

## 2.5 Nested Cross-Validation

As we have already discussed in lecture and Demo_3, the norm in machine learning is to have a lot more observations ($n$) than the number of features ($P$). Although it is tempting to think the same thing here, remember that the columns here are not features but rather samples, giving us a $P >> N$ problem. 

Whenever $P >> N$ as here, our typical approaches for cross-validation must change. We have several options, but here we will employ a __nested cross-validation__ approach. Other options would include a hold out or leave-one-out, although with some caveats.

**You will find that this method is not employed as much as it should be yet in gene expression analysis.** Why? Conceptually, this may be a bit hard to understand - and ML methods are already really different from what is done with the conventional methods like `DESeq2`. Let's go over what nested CV is to try to unpack this a little. 

### 2.5.1 Nested CV: An Overview

In biological and biomedical data, the sample size is frequently limited but, as we see here, we have `r ncol(train)-3` __genes__ serving as features. This means the number of predictors are much, much larger than the observations ($P >> n$). When we did the 10-fold CV above, each fold contained 10% of the data. However, when you have only a few observations (here, $n = 74$!) that's a very small number to be withheld in each fold for validation, leading to the overfitting situation we continue to see **even with CV**. Enter **nested cross-validation (CV)**: it maximizes use of the whole dataset for testing accuracy while maintaining the split between training and testing, thereby allowing us to get around the issue I just mentioned.

### 2.5.2 The `nestedcv` package 
Although we can also execute this with custom functions or loops, it's usually going to be far more expedient to leverage a pre-existing package. Here, we will use the `nestedcv` package. 

`nestedcv` partitions the dataset into outer and inner folds, with the default being 10-fold outer CV and 10-fold inner CV, also called __10 x 10 folds__ CV. The inner-fold cross validation tunes the optimal hyperparameters for whatever model(s) are run. The model is then fitted on the whole inner-fold and tested against the left-out data from the outer-fold. This is repeated across all outer-folds. Then, the test predictions from the outer-folds are compared against the true results for the test outer-folds. The combined results give a measure of accuracy for the whole dataset as you are accustomed, e.g., either RMSE for regression or AUC for classification.

Lastly, a final round of cross-validation is performed on the whole dataset to determine hyperparameters to fit the final model to the whole data, which can be used for prediction with external data (which could be yet another test set you have withheld, for example, as we've been doing).

**Whew!**

Shall we try it?!?

### 2.5.3 Running a Nested CV

#### 2.5.3.1. Set a search grid and train the model using the `nestedcv.train()` function.

Note that the setup looks a little reminiscent of our customized `caret` function. That's because the authors of this package are doing something similar, although they have improved the computational speed. We could do that too - **if** we wanted to write a package. Otherwise, we can just let ours run and take a coffee/tea break. :)

**IMPORTANT!**: We are not able to use $m_{try}$ as large as we are used to because we're doing inner- and outer-folds. We lowered $m_{try}$ here as a result. If you are doing this on different data and you get an error, that's something to look into!

**Note**: This took 14 mins on 16gb RAM, depending on what else I was asking my computer to do at the same time.
```{r}
searchGrid <- expand.grid(.mtry = floor(ncol(train) * c(0.01, 0.05, 0.10)),
                        .min.node.size = c(1, 5, 10),
                        .splitrule = "gini"                        
                        )
ncv <- nestcv.train(y = train$Treatment, x = train,
                    method = 'ranger',
                    tuneGrid = searchGrid, 
                    savePredictions = "final")
```

```{r, echo = FALSE, fig.cap = "Figure 4. Visualize the Nested CV Performance on RF + Grid Search"}
ggplot(ncv$outer_result[[1]]$fit) +
  scale_x_log10() +
  ggtitle("Results of Nested CV with hyperparameter tuning") +
  theme_bw()
```

#### **Question 19**: Notice that the y-axis is now **log-Loss** and not **accuracy**. You may need to do some digging, but can you figure out why? Citing sources can be helpful to ensure you're on the right track!

Source: https://www.cs.tufts.edu/cs/135/2023f/slides/day09_logistic_regression.pdf

> According to the Tufts powerpoint cited above, logLoss is also known as binary cross entropy. It measures how well a model predicts the probabilities of a label, with better performance represented by a lower logLoss value. I think the y-axis is now logLoss instead of accuracy because this plot is showing the logLoss values that come from the predicted probabilities as opposed to the accuracy values. The logLoss is computed by the predicted probabilities that are affected by the combination of hyperparameters, such as mtry and minimal node size (in this plot, 1, 5, and 10). The goal of this plot is to find the combination of hyperparameters that minimizes logLoss. Previously, we were looking for the combination of hyperparameters that maximized accuracy.

#### 2.5.3.2. Plot the Receiver-operator curves (ROC) and precision-recall curves for both the inner- and outer- loops.

**Note**: You're going to see something really weird here...

```{r, echo=FALSE}
# Plot ROC and Precision-Recall curves
op <- par(mfrow = c(1, 2))

## Outer CV ROC
plot(ncv$roc, 
     main = "Outer-folds ROC", 
     col = 'blue')
legend("bottomright", 
       legend = paste0("AUC = ", signif(pROC::auc(ncv$roc), 3)), 
       bty = 'n')

## Inner CV ROC
inroc <- innercv_roc(ncv)
plot(inroc, 
     main = "Inner-folds ROC", 
     col = 'red')
legend("bottomright", 
       legend = paste0("AUC = ", signif(pROC::auc(inroc), 3)), 
       bty = 'n')
```

#### 2.5.3.3. Get final accuracy by fitting to the test data.

```{r}
preds <- predict(ncv, 
                 newdata = test)
confusionMatrix(preds, test$Treatment)
```

#### **Question 20**: The ROC and precision-recall curves are suggesting very high accuracy. Woohoo!!... Right?! But is the confusion matrix telling us? Unpack the evidence more carefully, and decide whether model performance is really as good as it seems at first blush. 

**Hint**: Make sure to mention why/when ROC curves can be misleading.

> The Outer-folds ROC curve has an AUC of 0.911 (91.1%), while the Inner-folds ROC curve has an AUC of 0.918 (91.8%). Both of the AUC values are considered excellent and that the model would be a high performing and accurate model. However, the confusion matrix statistics provides insight into why this is misleading.

> There are multiple reasons ROC curves can be misleading. One is when there is a class imbalance. I checked this, and since the prevalence is 0.5, the dataset is balanced, so in this case, this is not the cause of the ROC curve being misleading.

> The model has a balanced accuracy of 0.8571 (85.71%). This exceeds the no information rate of 0.5 (50%), which is the baseline compraison. However, an accuracy of 85% is considered decent model performance. We continue to come back to the big p little n issue, which highlights that while we have many predictors, we do not have many observations, which can impact our model results, and make it difficult to get useful information from them.

> The above code mentions that the positive class is Control. The confusion results are as follows:
True Positives (TP): 5
False Positives (FP): 0
False Negatives (FN): 2
True Negatives (TN): 7
There are two False Negative values, indicating that some values are still being incorrectly classified. In this case, the model is identifying the Cold class, but is not always able to identify the Control class (positive class). Since we have the big p little n problem, even a slight change in these values can drastically impact the model.

>This shows that although the ROC and AUC are high, they may be misleading for our model, especially based on the results seen in the confusion matrix.

#### **Question 21**: Using help from generative AI or search engines, figure out how to assess the **stability** of the model across the nested folds. What conclusions can you draw? Write a 1-2 sentence conclusion for your stakeholders about whether or not you think the random forest is tunable and what you'd recommend next.

> There are various ways we can assess the stability of the model across the nested folds. Earlier, in Question 18, we checked the model stability across the folds. This code output the accuracy and kappa values for each of the 10 folds. One of the biggest takeaways from this code was that all 10 folds had identical accuracy and kappa values of 1. These are unlikely results that should be viewed with caution.

> Due to this, I do not think the Random Forest model is tunable. Although it did show some minor improvements, it was clear that our performance metrics were unlikely to improve. Since we have a big p little n problem, I would recommend trying a regularization model next. This may lead to better results through the use of penalties.


#### 2.5.3.4. Grab the important features.

Even if you decided that model performance is still poor and that you'd not recommend continuing with the RF, let's practice grabbing the important features so you have code to do it.

```{r, echo=FALSE}
nested_importantRF <- rf_filter(ncv, 
                               y = train$Treatment, 
                               x = train[, -ncol(train)], # drop Treatment
                               type ="full") %>% as.data.frame()
names(nested_importantRF) <- "MeanDecreaseGini"
```

```{r, echo=FALSE}
nested_importantRF %>% 
  arrange(desc(abs(MeanDecreaseGini))) %>% 
  top_n(10) %>% 
kable(
    format = "html",
    caption = "Table 7. Important Features from the Nested CV Grid Search on Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### 2.5.3.5. Compare our gene results to the ones out of `DESeq2`.

Remember, the ultimate goal was to see which ML method was getting us close to results from `DESeq2`, the respected, conventional method.

Let's filter the nested CV random forest importance:
```{r, echo = FALSE}
importantRF_filt <- nested_importantRF %>% 
  as.data.frame() %>% 
  filter(MeanDecreaseGini > 0)
```

Now let's compare that to what we got out of `DESeq2`. 

**Note** that I am setting a log-fold change filter to 1. In other words... any gene! Am I going to get any credible results?

```{r, echo = FALSE}
findOverlappingGenes(lfc = 1, importantRF_filt) %>% 
  arrange(desc(abs(MeanDecreaseGini))) %>% 
  kable(
    format = "html",
    caption = "Table 8. Overlapping Features identified from DESeq and Nested CV Random Forest") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

**IMPORTANT!** I have had this list reshuffle on me. I believe the issue is with the `rf_filter()` function we're using, and is something to consider for reproducibility!

#### 2.5.3.6. Find out what our top gene from nested CV random forest does.
Although the gene list has shuffled a bit from run to run, I did want to give you an example of one gene that I keep seeing come up again and again and _one way_ we might try to explore what these genes are and what they do. Because, at the end of the day, having a list of genes is rather dissatisfying if we want to know the underlying biology and mechanism of chill coma.

One of the genes that came up as the top important gene for me repeatedly was I went to [NCBI](https://www.ncbi.nlm.nih.gov), and just searched for __LOC100747964__. The gene should come up from _Bombus impatiens_, common Eastern bumblebee and it does. [UDP-N-acetylhexosamine pyrophosphorylase](https://www.ncbi.nlm.nih.gov/gene/?term=LOC100747964) seems to be involved in protein glycosylation and glycolipids, which is just a long way of saying it looks like it's a gene that is differentially regulated during metabolism. 

A second gene that came up frequently for me was __LOC100748162__, which is also a gene from _Bombus impatiens_ and [encodes pancreatic lipase-related protein 2](https://www.ncbi.nlm.nih.gov/gene/?term=LOC100748162), another metabolism gene. So, it may suggest one of the physiological effects that happens to _B. impatiens_ when it goes into chill-coma is an alteration in metabolism, which makes sense.

#### **Question 22**: Do you think we should we have any confidence in these results? Why or why not? Think about this from the perspective of what you would tell/advise your stakeholder.

> Source: https://www.ncbi.nlm.nih.gov/gene/?term=LOC100748162
> Source: https://www.ncbi.nlm.nih.gov/gene/?term=LOC100749486


> The two genes that showed up as overlapping features were LOC100748162 and LOC100749486.

> Gene LOC100748162 is a gene from Bombus impatiens that encodes pancreatic lipase-related protein 2. This was one of the genes that Professor Geist also saw in her model. As she mentioned above, it is a metabolism gene and in the context of this project, it could indicate that an alteration in metabolism can happen when the Bombus impatiens go into a chill coma.

> Gene LOC100749486 is a gene from Bombus impatiens that encodes abaecin. Abaecin is a naturally occurring peptide that is an important part of their immunity. This indicates that the immunity of Bombus impatiens may be affected when they go into a chill coma.

> Both genes LOC100748162 and LOC100749486 have positive log2FoldChange values, indicating that they are upregulated. Gene LOC100749486 is the most upregulated. Neither padj values for either of these genes are statistically significant.

> I think we can have limited confidence in these results; however, I don't think we should be overly confident in them. Although we have strong AUC values (0.952 and 0.91), we are using a very small dataset, which can cause the model to overfit. This leads to misleading results, which often look better than they actually are. Additionally, as mentioned previously, we have the big p little n problem, which can cause overfitting. Overall, the model provides some valuable results, such as the two genes that overlap between models. However, extensive trust in the model should be used with caution. I would not recommend this to be deployed in any production environments.


# 3 Takeaways & Next Steps

Hopefully the ROC and precision-recall curves from the nested CV clued you in to the fact that we have somewhat resolved the overfitting of our random forest, although we may still want to consider other options. One issue to consider with regard to RF is that of __OVERTRAINING__ - random forests are somewhat prone to it, despite all the great things about the algorithm. Secondly, we know this was a time-series of data, with different expression results more obvious at some time points than others (Verble et al. 2023). Thus, we would want to perform additional analyses before we feel too confident in these results. But we can see a positive impact of the nested CV.

## 3.1 Enter Phase 2 of this Project:

As we move forward with our benchmarking study, you have two choices. Choose your own adventure!

* **Path 1** - Potentially easier and more satisfying.
    
    Perform $k$-fold & nested CV Random Forest, as well as hyperparameter tuning, on a different VST from the 30-minute experiment only.

    If you decide to choose this adventure, you will use the VST data for the 30-min time point expression data only. You will also be given the `DESeq` results so you can use the `findOverlappingGenes()` function again to look at performance. Your task will be to take this vst, and work through all the same steps as in this first-half of the project. **Make sure to look at the important features at the end and comment on whether you have any confidence in your results!**

* **Path 2** - Potentially more challenging, and possibly dissatisfying.

    Ready to do go farther? Try performing $k$-fold & nested CV, as well as hyperparameter tuning, on SVM.

    If you choose this adventure, you will have the option to work through this same vst OR the 30-minute VST and your challenge is to apply SVM, including tuning the hyperparameters of __kernel__, __gamma__, and __cost__. Currently, we cannot get the important features out of nested cv SVM easily (alas - it needs another custom function!), so for this project you only need to engineer a best SVM model, discuss why you think it's the best model, and what your next recommendations are.

# 4 Final Comments / Advice

* Make sure to use the template I provide on GitHub for the project, rather than adding on to this document. It will just get too long! **If you can't upload the knitted file, just submit a note on Canvas that it wouldn't knit and that yoour RMD is on on your Team GitHub.* I expect some folks may have difficulty knitting. Be patient, too. It might take a half-hour or longer.

* **Don't forget!** Answer questions scattered throughout the Random Forest CV and Tuning Sections to ensure your understanding (you may copy these answers, without code, to your project document.)

**Files you need for the 30-min data will be found in the `data_30min` directory**:

1. `vst_30_min.Rdata` = the 30-min equivalent of the `vsData` object with the VST normalized read counts, formatted and ready to use as in this walk-through.

2. `vst_30_min.txt` = a text file version, which you likely do not need

3. `DESeq_on_30min_Data.Rmd` = an RMD file that was used to generate the 30-min data, plus some exploratory analysis. For the results see #4.

4. `DESeq_on_30_Data.html` = knitted file with some EDA and results to get you better acquainted to the data. Take a look! :)

5. `DESeqResults_30_min.Rdata` = the 30-min equivalent of the `DESeqResults` object that you can use with this walk-through.

**You will absolutely need to take a look at #4 to get you started. You will also want to use #1 and #5 for your analysis.**

# 5 Phase 2 of Project: Path 1

I have chosen to complete Path 1 for Phase 2 of the project:
Perform $k$-fold & nested CV Random Forest, as well as hyperparameter tuning, on a different VST from the 30-minute experiment only.

If you decide to choose this adventure, you will use the VST data for the 30-min time point expression data only. You will also be given the `DESeq` results so you can use the `findOverlappingGenes()` function again to look at performance. Your task will be to take this vst, and work through all the same steps as in this first-half of the project. **Make sure to look at the important features at the end and comment on whether you have any confidence in your results!**

## 5.1 Load in the data:

```{r, echo = TRUE, message=FALSE, warning=FALSE}
# Load in 30m data needed for Phase 2 of project
load("/Users/elizabethgreenan/Documents/DSE6330/Merrimack_DSE6630/Team_Directories/lambda/Module_Projects/Project_3/data_30min/vst_30_min.Rdata")

# Load in DESeq2 results
load("/Users/elizabethgreenan/Documents/DSE6330/Merrimack_DSE6630/Team_Directories/lambda/Module_Projects/Project_3/data_30min/DESeqResults_30_min.Rdata")

```

> NOTE: I chose to overwrite the variables from Phase 1. Therefore, no variables were renamed in this phase.

## 5.2 Pre-processing and feature selection

Set `filterCutoff = 5` and partition data for random forest tuning.
```{r}
splits <- doSplits(vst = vsData, 
                   algorithm = "rf", 
                   splitRatio = 0.8, 
                   filterCutoff = 5)   
train <- splits[[1]]
test <- splits[[2]]
```

> After filtering, there are 8,553 genes in the dataset. Phase 1 had 7,676 genes after filtering. This indicates that the 30 minute data retains more genes than the original dataset while using the same filterCutoff value and splitRatio. Additionally, although both datasets have the big p little n problem, this problem is much more pronounced in the 30 minute dataset. Therefore, this can cause the results to vary since there are so few samples.


## 5.3 Apply $k$-fold Cross-Validation (CV) to Random Forest

### 5.3.1. Leverage the `trainControl()` and `train()` functions from the `caret` package. 

Fit the same RF we fit last week for comparison:

```{r, echo = TRUE, eval=T}
rfOOB <- randomForest::randomForest(
  Treatment ~ ., 
  data = train)

pred.test.rf <- predict(rfOOB, test, type = "response")
confMat <- confusionMatrix(pred.test.rf, test$Treatment)
```

### 5.3.2. Employ a $k$-fold CV where $k = 10$.

Employ a $k$-fold CV where $k = 10$.

Set the control for $k = 10$ and then run the RF:
```{r, eval = T}
# Set the CV arguments
kFoldCtrl <- trainControl(method = "cv",    ## for k-fold CV
                          number = 10)      ## k
```

And fit the Random Forest model:
```{r}
rfCV <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl)    ## added in the 10-fold CV
```

```{r, echo = FALSE, warning=FALSE, message=FALSE, eval=T, fig.cap="Figure 1. Visualize the 10-fold CV Performance for 30 Min Data"}
ggplot(rfCV, highlight = TRUE) +
  ggtitle("Random Forest Performance After 10-fold CV for 30 Min Data") + 
  theme_bw()
```

> The plot above shows the Random Forest Performance After 10-fold CV for 30 Min Data. This shows that the gini rule outperformed the extratrees rule, just like in Phase 1 of the project. The combination of hyperparameters that gives the highest accuracy are: the gini splitting rule with approximately 1 randomly selected predictors (mtry). This combination results in an accuracy of approximately 0.67 (67%). From the plot alone, it is difficult to tell the exact mtry value, so when running the code below, I was able to determine the mtry value is 2. The big p little n problem is also impacting these results. Since the training set has 14 observations and the test set has 2 observations, any performance metrics will be highly impacted.

### 5.3.3. Extract best model.

You can also extract the best model as well to get the precise combination of hyperparameters.

```{r, echo = F}
rfCV$bestTune %>% 
kable(
    format = "html",
    caption = "Table 1. Hyperparameters of the best Random Forest Model using 10-fold CV for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

> This table shows that the hyperparameters of the best model are mtry = 2, splitrule = gini, and min.node.size = 1.

### 5.3.4. Assess for model stability.

Let's look at the fold-specific results for our best model. This is essential to assess **stability** of the model. 

```{r, echo=FALSE}
rfCV$resample %>% 
kable(
    format = "html",
    caption = "Table 2. Fold-specific accuracy for the 10-fold CV Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

> This shows a much different result than Phase 1 of the project. Here, we can see that not all of the folds have an accuracy and kappa of 1 like they did earlier. Instead, there are two accuracy values: 0 and 1. There are four kappa values: -1, 0, 1, and NA.

> Although this is set up for 10-fold, there are some missing values, resulting in only 9 folds being shown in the table. There was a warning message when running the code that said "There were missing values in resampled performance measures". I think since there is such a small sample size, that the accuracy was not able to be computed accurately for one of the folds.


```{r}

# Create accuracy vector with only 9 values

accuracy <- c(0, 1, 1, 1, 0, 1, 1, 1, 0)

# Calculate average accuracy
average_accuracy <- mean(accuracy)
# Calculate minimum accuracy
minimum_accuracy <- min(accuracy)
# Calculate maximum accuracy
maximum_accuracy <- max(accuracy)

average_accuracy
minimum_accuracy
maximum_accuracy

```

> The average accuracy is 0.6666 (66.66%). The minimum accuracy is 0. The maximum accuracy is 1 (100%).

> From Phase 1, the average accuracy was 63.57%, the minimum accuracy was 16.66%, and the maximum accuracy was 100%. Interestingly, the average accuracies were not very different, with the 30 minute average accuracy only slightly lower than the original dataset.


```{r}
# Check training and testing split to answer the question
nrow(train)
nrow(test)
```

> The above code shows that the training set has 14 observations and the test set has 2 observations. This is a much smaller number of observations than the original dataset, emphasizing the big p little n problem.

### 5.3.5. Fit the test data to the best tuned model from the 10-fold CV.

```{r, echo=TRUE}
rfCV <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl, 
               tuneGrid = rfCV$bestTune)   ## Add in the results of the CV 
                                           ## and auto tuning

pred.test.rf <- predict(rfCV, test, type = "raw")  ## type is now 'raw' 
## Store the confusion matrix for later comparison
confMatCV <- confusionMatrix(pred.test.rf, test$Treatment)
```

### 5.3.6. Compare the two confusion matrices.

Lastly, let's compare the results of what we just ran (stored in `confMatCV`) with what we the OOB performance from Demo_3 (stored in `confMat`).

```{r, echo = FALSE}
compareConfusion(confusionList = list(confMat, confMatCV)) %>% 
  kable(
    format = "html",
    caption = "Table 3. Comparing Accuracy - Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

> The above table shows identical results between Confusion Matrix 1 (OOB) and Confusion Matrix 2 (10-Fold CV). This is likely due to the very small sample size in the 30 minute dataset. Therefore, the models are having a difficult time generalizing, due to such small variance in the number of samples. These results must be interpreted with caution.

## 5.4 Further Tuning: Searching

**NOTE**: When you strike out on your own, you may get a 'bad seeds' error. This is often because you need to increase the size of your sample space.
```{r, echo = FALSE}
# Make a set of k = 10 seeds for reproducibility
seeds <- vector(mode = "list", length = 11)
for(i in 1:10) {
  seeds[[i]]<- sample.int(n=1000, 54)   # Increase 54 if you have a larger grid! 
}
# For the last model
seeds[[11]]<-sample.int(1000, 1)
```

```{r, echo = FALSE}
# Set the CV arguments
kFoldCtrl <- trainControl(method = "cv",    # for k-fold CV
                          number = 10,      # k
                          seeds = seeds)    # sets the seeds, one for each split

```

#### 5.4.1.1. Set the search grid.
```{r}
searchGrid <- expand.grid(
  mtry = floor(ncol(train) * c(.05, .15, .25, .35, .45)),
  splitrule = c("gini", "extratrees"),
  min.node.size = c(1, 3, 5, 10) 
)
```

```{r}
# Calculate the number of different combinations we will search with this grid
5 * 2 * 4
```

> Just like in Phase 1, we will seach 40 different combinations with this grid.

#### 5.4.1.2. Run the RF, this time adding in the search grid.

Be patient. This may take a few minutes to run.

```{r}
rfTuned <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl, 
               tuneGrid = searchGrid # Add in the search grid
)
```

```{r, echo = FALSE, fig.cap="Figure 2. Visualize the Grid Search Performance for 30 Min Data"}
ggplot(rfTuned, highlight = TRUE) +
  ggtitle("Random Forest Performance Grid Search Tuning for 30 Min Data") + 
  theme_bw()
```

> The combination of hyperparameters that gives the highest accuracy is: gini splitting rule, mtry = ~500, and a minimal node size of 3. This gives an accuracy of approximately 0.77 (77%). When compared to Phase 1, once again, gini outperforms extratrees.

> When looking at the table below, we can see that mtry = 427 and splitrule = gini. However, the min.node.size changed from 3 to 1. I think this is again likely due to the small sample size causing results to have some variance.

#### 5.4.1.3. Take a look at the hyperparameters of the best model.

```{r}
rfTuned$bestTune %>% 
kable(
    format = "html",
    caption = "Table 3. Results of the Grid Search on Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### 5.4.1.4. Fit the best model's hyperparameters to the test data.

```{r}
rfTuned <- train(Treatment ~.,  
               data = train,
               method = "ranger",
               trControl = kFoldCtrl, 
               tuneGrid = rfTuned$bestTune)   ## Use the best model
```

#### 5.4.1.5. Make the new predictions to assess accuracy.

```{r}
pred.test.rf <- predict(rfTuned, test, type = "raw")  ## type is now 'raw' 
# Store the confusion matrix
confMatTuned <- confusionMatrix(pred.test.rf, test$Treatment)
```

#### 5.4.1.6. Compare the three confusion matrices.

```{r}
compareConfusion(confusionList = list(confMat, confMatCV, confMatTuned)) %>% 
    kable(
    format = "html",
    caption = "Table 4. Comparing Accuracy - Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

> Once again, the results from Confusion Matrix 1 (OOB), Confusion Matrix 2 (10-Fold CV), and Confusion Matrix 3 (Grid Search) output identical metrics. These results are likely due to the big p little n problem. The accuracy for all three models is 0.5 and the kappa for all three models is 0. These results indicate that the models are not performing any better than random guessing.

### 5.4.2 Grid Search Using a Customized `caret` Method

#### 5.4.2.1. Write a customized `caret` method for random forest using the `ranger` method.

```{r, echo = FALSE}
makeCustomMethod <- function(paramList, methodName) {
  # Create a custom list, for a classification model and the ranger method. This list does not use loops.
  custom <- list(type = "Classification", 
                 library = "ranger", 
                 loop = NULL)
  # Create a dataframe that defines the hyperparameters that can be tuned, assign each of them to a class, and label them
  custom$parameters <- data.frame(parameter = paramList, 
                                  class = rep("numeric", length(paramList)), 
                                  label = paramList)
    # Create a function for the grid search
    custom$grid <- function(x, y, len = NULL, search = "grid") {}
    # Fit the model using the hyperparameter values in the grid search function
    custom$fit <- function(x, y, wts, param, lev, last, weights, classProbs) {
        # Fit the random forest model and specify mtry and num.trees values
        randomForest(x, y, 
               mtry = param$mtry,    ## Need to update to make dynamic
               num.trees = param$num.trees)
  }
  # Create a function for model predictions
  custom$predict <- function(modelFit, newdata, preProc = NULL, submodels = NULL)
  # Create the predictions from the model
  predict(modelFit, newdata)

  # Create a function for model probabilities
  custom$prob <- function(modelFit, newdata, preProc = NULL, submodels = NULL)
  # Create the probabilities from the model
  predict(modelFit, newdata, type = "prob")

  # Sort the results of the hyperparameter tuning by the first column
  custom$sort <- function(x) x[order(x[,1]),]

  # Return the class levels for the classification model
  custom$levels <- function(x) x$classes
return(custom)
}
```

#### 5.4.2.2. Run the customized method.

**NOTE**: This takes about 12 min to run on 16gb of RAM. This is exactly why we are doing caching for this assignment. Otherwise, your report will take forever to re-knit. 

```{r}
## Run the custom method setting the hyperparameters we want to use
customRF <- makeCustomMethod(paramList = c("mtry", 
                                           "num.trees", 
                                           "min.node.size",
                                           "splitrule"),
                             methodName = "ranger")

## Now expand the search grid
searchGrid <- expand.grid(.mtry = floor(ncol(train) * c(.05, .15, .25)),
                        .num.trees = c(50, 200), 
                        .min.node.size = c(1, 5, 10, 20),
                        .splitrule = "gini"
                        )
## Now run the RF with the custom tuned hypers and the new search grid
customTuned <- train(y = train$Treatment, x = train, 
                method = customRF, 
                metric = "Accuracy", 
                tuneGrid = searchGrid, 
                trControl = kFoldCtrl)
```

```{r, echo=FALSE, fig.cap = "Figure 3. Visualize the Customized Grid Search Performance for 30 Min Data"}
ggplot(customTuned, highlight = TRUE) +
  ggtitle("Customized RF Performance Grid Search Tuning for 30 Min Data") + 
  theme_bw()
```

> The following combination of hyperparameters gives the highest accuracy: num.trees = 50, mtry = 427, and min.node.size = 3. These hyperparameters produce an accuracy of approximately 0.81 (81%). For all three models (OOB, 10-Fold CV, and Grid Search), the gini rule provided the best results.


#### 5.4.2.3. Take a look at the hyperparameters of the best model.
```{r}
customTuned$bestTune %>% 
kable(
    format = "html",
    caption = "Table 5. Results of the Customized Grid Search on Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### 5.4.2.4. Refit the best model's hyperparameters to the test data set.
```{r, warning = FALSE, message=FALSE}
newTuned <- train(Treatment ~.,  
               data = train,
               method = customRF,
               trControl = kFoldCtrl, 
               tuneGrid = customTuned$bestTune)
```

#### 5.4.2.5. Make the new predictions to assess accuracy.
```{r}
pred.test.rf <- predict(newTuned, test, type = "raw")  
# Store the confusion matrix
confMatTuned2 <- confusionMatrix(pred.test.rf, test$Treatment)
```

#### 5.4.2.6. Compare the four confusion matrices.
```{r, echo = FALSE}
compareConfusion(confusionList = list(confMat, confMatCV, confMatTuned, confMatTuned2)) %>% 
    kable(
    format = "html",
    caption = "Table 6. Comparing Accuracy - Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

> The hyperparameter tuning with the number of trees and the new grid search (Confusion Matrix 4) had identical metrics to the OOB model (Confusion Matrix 1), the 10-Fold CV model (Confusion Matrix 2), and the original grid search model (Confusion Matrix 3). All four models have an accuracy of 0.50 (50%) and a kappa of 0, once again indicating that these models perform as well as randomly guessing.


```{r}
# Check model stability across folds
customTuned$resample
```

> I checked the model stability across folds, which provided different results than Phase 1. Here, not all of the accuracy values are 1. There are three folds with an accuracy of 0.50. Additionally, there are three kappa values: 0, 1, and NA.

> These results do not provide any imrpovement in results. Therefore, it is not worth continuing down this path.

### 5.5.3 Running a Nested CV

#### 5.5.3.1. Set a search grid and train the model using the `nestedcv.train()` function.

Note that the setup looks a little reminiscent of our customized `caret` function. That's because the authors of this package are doing something similar, although they have improved the computational speed. We could do that too - **if** we wanted to write a package. Otherwise, we can just let ours run and take a coffee/tea break. :)

**IMPORTANT!**: We are not able to use $m_{try}$ as large as we are used to because we're doing inner- and outer-folds. We lowered $m_{try}$ here as a result. If you are doing this on different data and you get an error, that's something to look into!

**Note**: This took 14 mins on 16gb RAM, depending on what else I was asking my computer to do at the same time.

```{r}
searchGrid <- expand.grid(.mtry = floor(ncol(train) * c(0.01, 0.05, 0.10)),
                        .min.node.size = c(1, 5, 10),
                        .splitrule = "gini"                        
                        )
ncv <- nestcv.train(y = train$Treatment, x = train,
                    method = 'ranger',
                    tuneGrid = searchGrid, 
                    savePredictions = "final")
```

```{r, echo = FALSE, fig.cap = "Figure 4. Visualize the Nested CV Performance on RF + Grid Search for 30 Min Data}
ggplot(ncv$outer_result[[1]]$fit) +
  scale_x_log10() +
  ggtitle("Results of Nested CV with hyperparameter tuning for 30 Min Data") +
  theme_bw()
```

#### 5.5.3.2. Plot the Receiver-operator curves (ROC) and precision-recall curves for both the inner- and outer- loops.

**Note**: You're going to see something really weird here...

```{r, echo=FALSE}
# Plot ROC and Precision-Recall curves
op <- par(mfrow = c(1, 2))

## Outer CV ROC
plot(ncv$roc, 
     main = "Outer-folds ROC", 
     col = 'blue')
legend("bottomright", 
       legend = paste0("AUC = ", signif(pROC::auc(ncv$roc), 3)), 
       bty = 'n')

## Inner CV ROC
inroc <- innercv_roc(ncv)
plot(inroc, 
     main = "Inner-folds ROC", 
     col = 'red')
legend("bottomright", 
       legend = paste0("AUC = ", signif(pROC::auc(inroc), 3)), 
       bty = 'n')
```

#### 5.5.3.3. Get final accuracy by fitting to the test data.

```{r}
preds <- predict(ncv, 
                 newdata = test)
confusionMatrix(preds, test$Treatment)
```

> The Outer-folds ROC curve has an AUC of 0.714 (71.4%), while the Inner-folds ROC curve has an AUC of 0.739 (73.9%). These are average AUC values that indicate moderate model performance. These are much more realistic values than we saw in Phase 1 of the project.

> The model has a balanced accuracy of 0.5 (50%). This matches the no information rate, which is the baseline comparison.The big p little n issue continues to be an issue for many of our metrics, highlighting that while we have many predictors, we do not have many observations. This can impact our model results and make it difficult to get useful information from them.

> The above code mentions that the positive class is Control. The confusion results are as follows:
True Positives (TP): 0
False Positives (FP): 0
False Negatives (FN): 1
True Negatives (TN): 1
There is one False Negative and one True Negative value. This indicates that the model only correctly identifies one instance of the negative class (Cold). However, there are zero True Positives or False Positives. This indicates that the model is not identifying the positive class (Control) at all.

#### 5.5.3.4. Grab the important features.

Even if you decided that model performance is still poor and that you'd not recommend continuing with the RF, let's practice grabbing the important features so you have code to do it.

```{r, echo=FALSE}
nested_importantRF <- rf_filter(ncv, 
                               y = train$Treatment, 
                               x = train[, -ncol(train)], # drop Treatment
                               type ="full") %>% as.data.frame()
names(nested_importantRF) <- "MeanDecreaseGini"
```

```{r, echo=FALSE}
nested_importantRF %>% 
  arrange(desc(abs(MeanDecreaseGini))) %>% 
  top_n(10) %>% 
kable(
    format = "html",
    caption = "Table 7. Important Features from the Nested CV Grid Search on Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

> The above table shows the important features from the Nested CV Grid Search on Random Forest for the 30 minute data. There are 13 features here, with MeanDecreaseGini values ranging from 1.41 to 2.8. Although these variables look promising, the big p little n issue still continues. Therefore, since we have such a small number of samples, these results do not generalize well. As we saw earlier, we had trouble with resampling folds due to the small number of samples. This leads to the conclusion that although these results look ok, they should not be trusted on their own, due to the small number of samples. We continue to have minimal confidence in the results, especially since in the case of the 30 minute data, the number of observations were so small.

#### 5.5.3.5. Compare our gene results to the ones out of `DESeq2`.

Remember, the ultimate goal was to see which ML method was getting us close to results from `DESeq2`, the respected, conventional method.

Let's filter the nested CV random forest importance:
```{r, echo = FALSE}
importantRF_filt <- nested_importantRF %>% 
  as.data.frame() %>% 
  filter(MeanDecreaseGini > 0)
```

Now let's compare that to what we got out of `DESeq2`. 

**Note** that I am setting a log-fold change filter to 1. In other words... any gene! Am I going to get any credible results?

```{r, echo = FALSE}
findOverlappingGenes(lfc = 1, importantRF_filt) %>% 
  arrange(desc(abs(MeanDecreaseGini))) %>% 
  kable(
    format = "html",
    caption = "Table 8. Overlapping Features identified from DESeq and Nested CV Random Forest for 30 Min Data") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F))
```

#### 5.5.3.6. Find out what our top gene from nested CV random forest does.

Source: https://www.ncbi.nlm.nih.gov/gene/?term=LOC100747885
Source: https://www.ncbi.nlm.nih.gov/gene/?term=LOC112213068
Source: https://pmc.ncbi.nlm.nih.gov/articles/PMC10701545/

> The above table shows that there are 14 genes that overlap between DESeq and the Nested CV Random Forest model. As mentioned in Demo 3, a negative log2FoldChange represents the down regulation of a gene, which is the gene getting turned off. A postiive log2FoldChange represents the up regulation of a gene, which is the gene getting turned on. Here, gene LOC100747885 is the most down regulated, with a log2FoldChange of -4.084690. Gene LOC100748162 is also down regulated, with a log2FoldChange of -3.578099. Conversely, gene LOC112213068 is the most up regulated, with a log2FoldChange of 1.518687.

> The padj (adjusted p-value) for the genes mentioned above are 0.0133650, 0.0125250, and 0.0191387. Since these are < 0.05, they are statistically significant. There are a total of 5 genes that are statistically significant in this table, while the remaining 9 are not.

> The most down regulated gene, LOC100747885, is a phospholipase A1 gene. The function of this gene is that it enables lipase activity.

> The most up regulated gene, LOC112213068, is uncharacterized. According to the National Libary of Medicine, an uncharacterized gene is one whose function has rarely been identified. Therefore, the effect the chill coma has on this gene is unknown.

> The two overlapping features from Phase 1 were LOC100749486 and LOC100748162. Gene LOC100748162 shows up as an overlapping feature in both Phase 1 and Phase 2 of the project.

# References
