---
title: "Demo 4: Patient Privacy & Data Integrity"
subtitle: "Merrimack College DSE6630: Healthcare & Life Sciences Analytics"
author: "Vicky (Xin) Guo"
date: "16 June 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_crop: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,    
                      cache.comments = TRUE,
                      size = 13)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# clean, set up, and load
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
rm(list = ls(all = TRUE))

pacman::p_load(here,
               tidyverse, 
               ggplot2, 
               kableExtra,
               forcats,
               jsonlite, 
               lubridate,
               sqldf, 
               SQLite, 
               RSQLite,
               RMySQL,
               DBI)

## load the data files
load("patient-data.Rdata")
## a little tom-foolery to make a fake, local instance of a database to practice queries
con <- dbConnect(RSQLite::SQLite(), ":memory:")
dbWriteTable(con, "patient_info", patient_info)
dbWriteTable(con, "patient_disability", patient_disability)
dbWriteTable(con, "patient_clinical", patient_clinical)
dbWriteTable(con, "patient_insurance", patient_insurance)
dbWriteTable(con, "patient_binge_drinking", patient_binge_drinking)

## now unload the datafiles to enhance the simulation so it looks just like connecting to an external database!
rm(patient_info)
rm(patient_disability)
rm(patient_clinical)
rm(patient_insurance)
rm(patient_binge_drinking)
```

# 1 Introduction

You have just been brought onto a new project that is using an __mHealth__ (mobile health) application (**"iQOL"**, where QOL stands for "Quality of Life") to track the health of patients with and without disabilities. The goal of this mHealth application is to better understand how functional and quality of life differs day-to-day among patients with and without a disability.

As a first pass, you ask the back-end (database) development team to provide you with some descriptive data from the first 100,000 users of the mHealth application. You want to make a report to stakeholders about who is using the application, the nature of the (dis) ability, geographic breakdown, and descriptive statistics about other information gathered by the mHealth application when a patient first signs up for the mHealth Tracker. 

The app's back-end developer sends you several tables from the database in an Rdata file, with the following descriptions of fields:

```{r, echo=FALSE}
dict <- tribble(
  ~ Variable, ~ Description,
  "PT_KEY", "Patient unique identifier",
  "INS_KEY", "Insurance unique identifier",
  "FIRST_NAME", "Patient first name",
  "SURNAME",  "Patient surname",
  "DOB",  "Date-of-birth",
  "AGE_DAYS",   "Age in days at time of database pull",
  "AGE_YEARS",  "Age in years at time of database pull",
  "EMAIL_ADDRESS", "Patient email address",
  "STREET",   "Patient mailing address",
  "CITY",  "City of mailing address",
  "STATE",  "State of mailing address",
  "ZIPCODE", "Zipcode of mailing address",
  "TELEPHONE", "Patient primary phone number",
  "ADULT_CHILD", "Flag if user is under 18 (we're developing a separate app for teens)",
  "FLU_INNOC", "Patient states they received their influenza vaccine this year",
  "COVID_INNOC", "Patient states they received their COVID-19 vaccine this year",
  "TET_INNOC", "Patient states they received their tetanus vaccine in past 10 years",
  "DATE_LAST_SEEN", "Patient reported last date seen by primary care physician",
  "NEW_PATIENT", "New user to the application",
  "PUBLIC_INS", "Patient reports receiving public insurance (Medicaid, Medicare, VA Benefits)",
  "DISABILITY_STATUS", "Patient-reported disability status, per federal guidelines",
  "EMPLOYMT_STATUS", "Patient-reported employment status for past 6 months",
  "EDUC_LEVEL", "Patient-reported highest educational attainment",
  "BMI_CATEGORY", "Body mass index categorization; BMI < 18.5 Underweight, 18.5 < BMI < 25 Normal, 25 <= BMI < 30 Overweight, BMI >= 30 Obese",
  "SMOKING_STATUS", "Patient is a former or current smoker, or has never smoked",
  "BINGE_DRINKING", "Patient reports at least one episode of binge drinking in past 6 months"
)

kable(
    dict,
    format = "html",
    col.names = c("", ""),
    caption = "Table 1. Data dictionary of available fields in the mHealth database pull.") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

## 1.1 Database connections

If your developer had given you `.db` files, or if you were accessing a server, you would have needed to walk through two steps, which we will simulate here, in order to load the tables into R for analysis. 

**Importantly**, the general purpose of doing this work in R will generally be for **research/analysis only**. However, the advantage to writing it in SQL is two-fold:

1. You can use your SQL queries on the server itself to do the same work, say, for a report you need to generate regularly.

2. If you DO need to update tables in the database, you can actually do that from R!!

**A word of caution**, however: if you've been given access to the __production__ databases, do not ever alter or update the tables unless you've been given explicit permission to do so. 

### 1.1.1 Making connections

```{r, echo=TRUE, eval=FALSE}
## This is how you would set a connection to an external server

# con = dbConnect(MySQL(),
#                 dbname = 'my_db_name',
#                 host = 'my_host',
#                 port = 3306,   ## The typical port used for SQL and MySQL servers
#                 user = 'username',
#                 password = 'password')

## Can be helpful to establish which flavor of SQL

# options(sqldf.driver = "SQLite")
```

But here, I've already made a connection to our mock database, also stored into `con`. So we will use that.

### 1.1.2 How many tables?

Usually the first thing we want to know if we're unfamiliar with a database is **how many tables are in it**. We are going to make our first SQL query together to get the names of all the tables in the connection, `con`. Notice that I have to use `dbGetQuery()` as my function to set the query, then I have to specify the name of the connection, `con`. The actual query itself is then given as a string in quotations after that.

```{r}
dbGetQuery(con, "SELECT name FROM sqlite_master WHERE type='table'")
```

### 1.1.3 What's in each table?

Notice that, unlike when we work with dataframes directly in R, we can't __see__ the tables. I did this on purpose; I'm hiding them from you so that it's as much like a simulated database experience as possible.

So, if we're flying blind, how can we find out what's in the tables? Simple - we write a query! Since we have five tables, the easiest way to do this is probably to write a `for` loop, where we query each table as we loop. 

Although there are multiple ways we could choose to execute this, perhaps some more clean than others, we are going to use the SQLite statement `PRAGMA`, which is a SQLite-specific extension and can be used to modify an operation or query a SQLite library. We're using it for the second purpose. In other words, we will use `PRAGMA` to allow us to query each table using `dbGetQuery`. Confused yet? It's okay if you are!

1. Make a query using `dbGetQuery` to get all of the names of the tables on the connection `con`. We will store these table names into `tables`. (Note that I didn't call them `table` because that's the name of a base `R` function! Best programming practice will ALWAYS have us not overwrite base objects; but in `R`, there are no protected name spaces, like there are in some other programming languages. So, we as the programmer must be cautious.)

2. Start the `for` loop, looping through each table by its `name`. 

3. Make a second query using `dbGetQuery`, this time setting the connection `con` AND using `PRAGMA` to get the `table_info` of each table. `table_info` returns a set of pre-specified fields on information, but I am only going to print the first two fields for brevity.

And that's it! A little clearer? If not, work through this code trying to change variables to help yourself figure out how it works. It's a **query within a query!**

```{r}
## 1. Make a query to get the list of tables from the connection, which we store in tables
tables <- dbGetQuery(con, "SELECT name FROM sqlite_master WHERE type = 'table' ")

# Loop over each table by name and print the table information 
for (t in tables$name) {
  ## Store the table information after making a second table-specific query about what's in the table using PRAGMA and table_info()
  table_info <- dbGetQuery(con, paste0("PRAGMA table_info(", t, ")"))
  ## Print the name of the table so we know the order 
  print(t)
  ## Show the first two columns from the table_info() of each table in the connection
  print(table_info[, 1:2])
}
```

It might feel like we're starting a bit deep, but don't worry - we can get a lot of information using simpler queries too, including assessing how well our developer has been protecting __patient privacy__.


# 2 Ensuring Patient Privacy

You had already decided, since you're still learning **SQL (Structured Query Language)** to continue using that in R for practice rather than the more familiar `tidyverse`. See, we would have the option to __store our dataframes__ from the connection `con` and manipulate them in all the familiar ways! And we will, in good time. But first, we are going to continue to use `dbGetQuery()` from the `SQLite` package to help us to make our queries.

You noticed as soon as you saw the fields of some of the tables that the developer sent that there's a __big problem!!__ The identities of your mHealth users are __NOT__ being safeguarded correctly. Uh-oh!

##### Just how bad IS the problem?

## 2.1 Checking for duplicate keys

You decide to first check for duplicates because, well, you asked for 100,000 adult patients... did you actually get 100,000? You start in the `patient_info` table because you see a lot of problems there in terms of protecting patient privacy!

```{r}
dbGetQuery(con, "SELECT pt_key, 
  COUNT(*) AS patient_count
  FROM patient_info
  GROUP BY pt_key
  HAVING patient_count > 1")
```

Whew! No problems here. Notice first that, to write this query, we are using the `dbGetQuery()` function in R (which is serving as a __wrapper__) to call our SQL query on our database connection, `con`. Our query is given as a string in quotations.  Also notice that even though the __field__ in our table is `PT_KEY`, in SQL letter case doesn't matter. In fact, upper case is usually reserved for certain __actions__ in SQL: `SELECT`, `COUNT`, `FROM`, `GROUP BY`, and `HAVING`. These are the __functions__. However, you don't even have to use the case for your query to run! Notice, too, that we used the `(*)` wildcard to select ALL patient keys. 

Also, note that even though I put each statement on separate lines (which is how we often write it in a query on a SQL server), we do not have to. We just do that to improve _readability of the code_. 

#### Question 1 [1 point]:

Try to translate into English exactly what this SQL query does.

**Answer：**
This query groups every row in patient_info by patient key, counts how many rows share each key, and keeps only the keys that show up more than once. It returns the patient key and the row count for that key. Since nothing is returned, every PT_KEY is unique — no duplicate patients.


#### Question 2 [1 point]:

You try it: check for duplicate insurance keys. Make sure to query the right table! (**Hint**: You should get 81.)

```{r}
dbGetQuery(con, "SELECT ins_key, 
  COUNT(*) AS insurance_count
  FROM patient_insurance
  GROUP BY ins_key
  HAVING insurance_count > 1")
```

#### Question 3 [1 point]:

Is this a problem? Why or why not? (Hint: Can multiple individuals be on the same insurance policy?)

**Answer:**
No, this isn't a problem. An insurance key identifies a policy, not a person. Multiple people are routinely on the same policy.So INS_KEY is naturally one-to-many: one policy can cover many patients. The duplicates are expected. It would only become a problem if I mistakenly treated INS_KEY as a unique per-patient identifier, or joined on it expecting one row per person.

## 2.2 Dropping fields with sensitive data

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_info") %>% 
head() %>% 
  kable(
    format = "html",
    caption = "Table 2. The patient_info table BEFORE removing any sensitive information. ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

Let's start by dropping ALL of the ultra sensitive data from the `patient_info` table: date-of-birth, age_days, name fields, street address, telephone, & email address from `patient_info`. 

In SQL, there is a function called `DROP COLUMN` that unfortunately is not supported by any of the packages we are using in R. Instead, we will have to use the `SELECT` function to choose the fields we want to keep. 

But note that it is not nearly as straightforward as if we were doing it in R. We will have to go through a series of steps to do it correctly; NOT ONLY THAT, __but it is making permanent changes to the database connection!__ PROCEED WITH CAUTION in the real world!

This is why there are often two versions of the database, one that is in __production__ and one that is used for __research/analysis__. Thus, we can think of this as we are making corrections to the analytical database. But we will also need to make sure to __give our suggestions__ back to the developer!!

* 1. Make a copy of the `patient_info` table, called `temp`, from just the columns we want to __keep__.

* 2. Now, this is blank. We have to actually `INSERT INTO` the new table, `temp`, the columns we want!

* 3. Remove, or `DROP TABLE` the original `patient_info` table

* 4. `RENAME` the table `temp` back to the original name, `patient_info`, by using the `ALTER TABLE` statement.

Notice too that to execute these table-level functions I am now using `dbExecute()` from `SQLite` rather than `dbGetQuery()` because these are not queries. These are executive statements.

```{r, results='hide'}
# Step 1
dbExecute(con, "CREATE TABLE temp AS 
  SELECT pt_key, age_years, city, state, zipcode,educ_level
  FROM patient_info")

# Step 2
dbExecute(con, "INSERT INTO temp (pt_key, age_years, city, state, zipcode,educ_level)
  SELECT  pt_key, age_years, city, state, zipcode,educ_level
  FROM patient_info")

# Step 3
dbExecute(con, "DROP TABLE patient_info")

# Step 4
dbExecute(con, "ALTER TABLE temp RENAME TO patient_info")
```

**Important**: We have permanently replaced the original `patient_info` table the developer gave us!!!

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_info") %>% 
head() %>% 
  kable(
    format = "html",
    caption = "Table 3. The patient_info table AFTER removing any sensitive information. ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

#### Question 4 [1 point]:
Your turn!

Remove all the sensitive information remaining in any of the other tables. 

```{r}

# Step 1
dbExecute(con, "CREATE TABLE temp AS 
  SELECT pt_key, flu_innoc, tet_innoc, covid_innoc, bmi_category, smoking_status
  FROM patient_clinical")

# Step 2
dbExecute(con, "DROP TABLE patient_clinical")

# Step 3
dbExecute(con, "ALTER TABLE temp RENAME TO patient_clinical")


dbGetQuery(con, "SELECT * FROM patient_clinical") %>%
  head() %>%
  kable(
    format = "html",
    caption = "Table 4. The patient_clinical table AFTER removing sensitive information.") %>%
    kable_styling(bootstrap_options = c("hover"))

```

Not sure what to do? Load the R script and run it instead to proceed.

```{r}
source("q4_answer.R")
```

Make sure to demonstrate that the sensitive information is gone using the follow code that employs the `head()` function after removing the comments and making it executable. 

```{r, echo = TRUE}
dbGetQuery(con, "select * from patient_clinical") %>% ## fill in the name of the table you removed sensitive info from
head() %>% 
  kable(
     format = "html",
     caption = "Table 4. The table AFTER removing sensitive information. ") %>%
     kable_styling(bootstrap_options = c("hover"),
)
```


#### Question 5 [1 point]:
**Have your noticed that I am having to first extract the table as a dataframe using `dbGetQuery()` before I can pass it through `head()`?**

Why do you think I have to do this?

**Answer:**
Because head() and kable() are R functions — they operate on R objects in R's memory. The tables don't live in R; they live inside the SQLite database, which is effectively a separate system sitting behind the con connection. dbGetQuery() runs the query over that connection and pulls the result back into R as a dataframe.

## 2.3 Preserving integrity while removing anything that can identify a patient.

When looking at the `ZIPCODE` column of the `patient_info` table, you start to worry about some of the entries. Some zipcodes have the four-digit US postal service identifiers, which can really narrow down a patient's identity to a much smaller geographic region. This makes you uneasy (and me too!). So, we decide to preserve zipcode (a much larger geographic area) while dropping the four-digits at the end when they exist. 

You know you could do this easily with `gsub()` as we've used before on a dataframe - but we aren't currently working with a dataframe! So, you decide to try doing it the "SQL-way" first to see if you can do it.

* 1. Create a `temp` table. This time, we are making it from everything in the `patient_info` table and we are specificying it as a `TEMPORARY TABLE` with our `CREATE` statement. This is to save us the two steps we had to do to make the `temp` table earlier.

```{r, results='hide'}
dbExecute(con, "CREATE TEMPORARY TABLE temp AS
  SELECT * FROM patient_info")
```

* 2. Now use the `UPDATE` statement with the `SUBSTR()` (substring) function to just keep the first 5 characters of the ZIP code. Note that we are doing this on the `temp` table and not `patient_info`.

```{r, results='show'}
dbExecute(con, 
          "UPDATE temp SET zipcode = SUBSTR(zipcode, 1, 5)")
```

Uh-oh... 200,000? Does that mean we __duplicated rows?__

```{r, results='show'}
dbGetQuery(con, 
           "SELECT COUNT(*) AS row_count FROM temp")
```

Noooooo! What happened? Weren't we just trying to lop off the "-xxxx" off the end of a 5-digit ZIP code? 

So, the problem is that SQL is not smart enough to know what to do when data match in more than one column. It errs on the side of caution, preserving the data - which inherently isn't a bad thing at all, just a bit annoying. Although one way to avoid it is to ensure that you're mapping any many-to-many relationships (if they exist), but sometimes it is easier to simply remove duplicated rows.

##### Enter your first major lesson in writing SQL:
Always, always, ALWAYS alter a **TEMPORARY** version of the table when you're working in your database connection. Because undo buttons do not exist (although back-end developers will write in safeguards and, chances are, you wouldn't be allowed to alter the production database unless you have a really sloppy back-end dev).

* 3. Okay, let's tidy up! Notice that what I am doing is using the `DISTINCT` statement to remove all duplicated rows from the temp table. Not only that, rather than having to use another `ALTER` statement, I am choosing to just overwrite `patient_info` so then we're done.

```{r, results='hide'}
## Drop the original
dbExecute(con, "DROP TABLE patient_info")

## Remove the duplicates from the temp and make a new patient_info
dbExecute(con, "CREATE TABLE patient_info 
  AS SELECT DISTINCT *
  FROM temp")

## Drop the temporary
dbExecute(con, "DROP TABLE temp")
```

```{r, echo = FALSE}
dbGetQuery(con, "select * from patient_info") %>% 
head(10) %>%
  kable(
    format = "html",
    caption = "Table 5. Did we fix the ZIP codes? ") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

Yay, we did it! But let's pause to reflect a moment: Didn't this feel like A LOT more work than if we'd just made the dataframe in R and then used the one-liner `gsub()`? Arguably, I think so - but it depends on our ultimate goal. If the goal is to `UPDATE` the database, we would need to do it this way and not just `gsub()` it!

# 3 Descriptives

Next, since we're working in SQL and flying 'blind', we would also like to try to get some sense for what these data contain, if there are any issues we need to take back to the developer for integrity as well as privacy concerns. We start by going through and looking at the tables descriptively.

## 3.1 Geographic representation

```{r}
dbGetQuery(con, 
      "SELECT COUNT(DISTINCT state) AS num_states FROM patient_info")
```

##### Does 89 states seem a little odd to you?
It does to me! Let's see what they are:

```{r, echo = FALSE}
dbGetQuery(con, 
      "SELECT DISTINCT state FROM patient_info") %>% 
  arrange(STATE) %>% 
  head(12) %>% 
  kable(
    format = "html",
    caption = "Table 6. What's going on with the states??!?") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```
#### Question 6 [1 point]:

What is your best guess about what you think has happened? Do you need to iunclude this in your report back to the developer?

**Answer:**
There are only 50 US states, plus DC and a few territories — nowhere near 89. The distinct values reveal the cause: the STATE field contains non-US subdivisions. Alongside real US codes (AK, AL, CA…) there are Canadian provinces (AB = Alberta, BC = British Columbia) and Mexican states (AG = Aguascalientes, CH = Chihuahua), plus a few codes that don't map to anything obvious. So the pull was never restricted to the US, and STATE isn't being validated against a known list of values. Yes — this goes in the report, both because it breaks the spec (I asked for US-only) and because it signals there's no domain constraint enforced on STATE at the source.

#### Question 7 [1 point]:

Confirm your suspicions by writing a query to print all the ZIP codes in **AB**. Use the `SELECT`, `DISTINCT`, and `WHERE` statements. For example, in SQL, `SELECT zipcodes FROM patient_info WHERE state = 'AL';` would return all of the ZIP codes in the state of Alabama. Look up one of the ZIP codes - does it confirm your suspicions about what happened with the data pull?

```{r}
dbGetQuery(con, "SELECT DISTINCT zipcode FROM patient_info WHERE state = 'AB'")

```

**Answer:**
Querying the ZIP codes where state = 'AB' returns alphanumeric values starting with T (e.g. T2E 8…, T5Z 0…). These are Canadian postal codes, not US ZIP codes. the T prefix is Alberta — which confirms the pull included Canadian users. They also appear truncated to 5 characters, because the earlier SUBSTR(zipcode, 1, 5) step (written for US 5-digit ZIPs) ran against every row and clipped the 6-character Canadian codes, leaving them malformed.

#### Question 8 [2 points]:

Decipher, into English, what you think this SQL query is doing.

```{r}
dbGetQuery(con, 
  "SELECT COUNT(DISTINCT pt_key) AS patient_count FROM patient_info WHERE state = 'AL' AND city = 'WETUMPKA' ")
```

**Answer:**
This is counts the number of disctinct patients in patient_info whol live in City WETUMPKA, Alabama. 

Now write your own! Pick your hometown or your homestate and find out how many patients in the `patient_info` table are from there.

```{r}
dbGetQuery(con, 
  "SELECT COUNT(DISTINCT pt_key) AS patient_count FROM patient_info WHERE state = 'MA' AND city = 'WORCESTER'")
```

## 3.2 Age

SQL does not have a lot of the sophisticated statistics we have grown accustomed to having access to in R, but we can execute the basics like max, min, and mean. Let's find the average, max, and minimum age in years of our patients in the `patient_info` table.

```{r}
dbGetQuery(con, "SELECT 
    MAX(age_years) AS max_age,
    MIN(age_years) AS min_age,
    AVG(age_years) AS mean_age
  FROM patient_info")
```
Pretty easy, right? This wasn't so bad!

#### Question 9 [2 points]:

Can you find the max, min, and mean age of all patients who reside in Massachusetts? (**MA**). 

**Hint**: Use the `WHERE` statement! You should get a mean of 41.90843, max of 96.6, and min of 20.5.

```{r}
dbGetQuery(con, "SELECT 
    MAX(age_years) AS max_age,
    MIN(age_years) AS min_age,
    AVG(age_years) AS mean_age
  FROM patient_info WHERE state = 'MA' ")
```

## 3.2 Disability Status

This is our primary interest, right? It would be nice to know the counts of folks in each of the disability categories, which we can return in SQL just as we could in R on a dataframe using the `table()` function or `dplyr`'s `count()` or the `n()` of `tidyverse`.

Notice we need to switch to the `patient_disability` table this time. We introduce the `GROUP BY` statement, which works very much like it does in `dplyr` and `tidyverse`! 

```{r}
dbGetQuery(con, 
  "SELECT disability_status,
  COUNT(*) AS patient_count
  FROM patient_disability
  GROUP BY disability_status")
```

#### Question 10 [2 points]:

Now do the same for `EMPLOYMT_STATUS` and `EDUC_LEVEL`, changing the tables as appropriate.

```{r}
# Your code here for employment status
dbGetQuery(con, 
  "SELECT employmt_status,
  COUNT(*) AS patient_count
  FROM patient_disability
  GROUP BY employmt_status")

```

```{r}
# Your code here for education level
dbGetQuery(con, "SELECT name FROM pragma_table_info('patient_info')")
dbGetQuery(con, "SELECT name FROM pragma_table_info('patient_disability')")

# Due to we drop EDUC_LEVEL columns earily, nowe we cannot get that columns in patient_info table

dbGetQuery(con, 
  "SELECT educ_level,
  COUNT(*) AS patient_count
  FROM patient_info
  GROUP BY educ_level")
```

# 4 Joining data tables in SQL

So much of what you may need to do will involve joining, or merging, tables in some way. You may do this on the fly to produce unique tabulations of data. You may make a new, temporary table that you export to a dataframe to perform analysis upon. We will be doing it for the second purpose.

## 4.1 Left joins

By now in your data science career, the idea of left vs. right, inner vs. full joins are probably starting to feel pretty comfortable. Good! One could argue that **the table join was invented in SQL**, so what better place to practice them?

A `left join` will keep all of the data on the left but only the data that match from the right table. If any data are missing in the right table, they will be dropped. 

Let's perform a `left join` of the `patient_info` and `patient_disability` tables first.
```{r}
dbExecute(con, 
  "CREATE TABLE merged AS SELECT *
  FROM patient_info
  LEFT JOIN patient_disability
  ON patient_info.pt_key = patient_disability.pt_key")
```

Notice the syntax: `ON patient_info.pt_key = patient_disability.pt_key`. We name the left-hand table, then say what key we want to use; then we set that equal to the matching key in the right-hand table.

#### Question 11 [1 point]:

Write a quick query to report back how many rows successfully merged.

```{r, results='show'}
dbGetQuery(con, "SELECT COUNT(*) AS info_count FROM patient_info")
dbGetQuery(con, "SELECT COUNT(*) AS merged_count FROM merged")

```

Any problems to report back to the back-end developer?
**Answer:**

The merged table has 100,000 rows, exactly matching the 100,000 in patient_info. Since a LEFT JOIN preserves every left-hand row, an identical count means there was no fan-out (no duplicate pt_keys in patient_disability) and every patient found a match. No problems to report from this join.


## 4.2 Inner join with aliasing

Next, let's practice an inner join this time using what's called `aliasing` - because old-timey SQL programmers HATED writing out table names! They'd just call them 'a', 'b', or whatever else they felt like. Here, we will use `m` for the `merged` table and `c` for the `patient_clinical` table.

```{r, results = 'hide'}
dbExecute(con, 
  "CREATE TABLE merged2 AS
  SELECT 
    m.pt_key,
    m.ins_key,
    m.age_years,
    m.city,
    m.state,
    m.zipcode,
    m.disability_status, 
    m.employmt_status,
    c.flu_innoc,
    c.tet_innoc,
    c.covid_innoc,
    c.bmi_category,
    c.smoking_status
  FROM merged AS m
  INNER JOIN 
  patient_clinical AS c
  ON m.pt_key = c.pt_key")

## Now remove any duplicates:

## Drop the original
dbExecute(con, "DROP TABLE merged")

## Remove the duplicates from the newly merged file and make it the new merged
dbExecute(con, 
  "CREATE TABLE merged 
  AS SELECT DISTINCT *
  FROM merged2")

## Drop merged2 to tidy up
dbExecute(con, "DROP TABLE merged2")
```

And how did we do?
```{r}
dbGetQuery(con,
           "SELECT COUNT(*) AS row_count FROM merged")
```

Excellent! It looks like we haven't lost any rows in the merging process. Let's see where we are currently at.

```{r, echo = FALSE}
dbGetQuery(con, "SELECT * FROM merged") %>% 
  head() %>% 
  kable(
    format = "html",
    caption = "Table 7. Our beautifully merged table...") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

#### Question 12 [1 point]:

Write either a `LEFT JOIN` or `INNER JOIN` to merge the `merged` table with the `patient_insurance` table. Make sure to join on TWO keys; you will find the `AND` statement helpful for this. Try using aliases!

```{r, results='show'}
dbExecute(con,
  "CREATE TABLE merged2 AS
  SELECT
    m.*,
    i.public_ins
  FROM merged AS m
  INNER JOIN patient_insurance AS i
  ON m.pt_key = i.pt_key
  AND m.ins_key = i.ins_key")

dbExecute(con, "DROP TABLE merged")
dbExecute(con, "CREATE TABLE merged AS SELECT DISTINCT * FROM merged2")
dbExecute(con, "DROP TABLE merged2")

#Validate the result
dbGetQuery(con, "SELECT COUNT(*) AS row_count FROM merged")

```

Not sure what to do? Load the R script and run it instead to proceed.

```{r}
source("q12_answer.R")
```


#### Question 13 [1 point]:

Any problems to report back to the back-end developer?

**Answer:**
The inner join on both pt_key and ins_key returned 100,000 rows — the same count as before. Since an inner join drops any row that doesn't match on both keys, an unchanged count means every patient had a consistent insurance record: the pt_key/ins_key pairing in merged agrees with patient_insurance for all 100,000 patients. No problems to report from this join.


## 4.3 Joining when there are duplicate keys

There is another table, `patient_binge_drinking`, that was pulled from the backend but there seems to be a problem with it. Notice that there is no `pt_key`!

```{r, echo = FALSE}
dbGetQuery(con, "SELECT * FROM patient_binge_drinking") %>% 
  head() %>% 
  kable(
    format = "html",
    caption = "Table 8. The patient binge-drinking table as pulled from the database.") %>%
    kable_styling(bootstrap_options = c("hover"),
)
```

#### Question 14 [1 point]:

Why is it problematic for merging if the patient key is missing and only the insurance key is available?

**Answer:**
Because INS_KEY identifies a policy, not a person, and INS_KEY is one to many. With no pt_key, the only thing I can join on is ins_key, which isn't unique per patient. That creates a many-to-many join: a single insurance key in this table can match several patients in merged, so the binge-drinking answer gets copied onto every person on that policy. One member's "Yes" would be attributed to their whole family.  So the data can't be merged at the patient level without either fabricating or corrupting the attribution.


There are really a couple of ways we could deal with this issue. First, and probably most practical, would be to ask our back-end developer to make sure the tables are linked up in the database. Something seems off! (**Hint**: Make sure to give them this recommendation!).

But since we are also under a tight timeline to stakeholders, we are going to do our best to 'plow through' as quickly as we can with what we have. After all, if we remove the duplicated insurance keys entirely, we only lose 81 observations total - not so bad out of 100,000 for a first pass, right?

Let's do it!

## 4.3.1 Using `tidyverse` when we do not want to make any permanent changes to the connection.

I have mentioned this several times, but we are definitely at a point where we do **NOT** want to be sending any changes back to the database. We don't want to permanently delete any data, for one thing!

So, from this point forward, we will lean on our trusty friend from the past few weeks, `dplyr` and `tidyverse`.

* 1. Extract the `merged` table and store it as a dataframe. 

```{r}
merged <- dbGetQuery(con,
                     "SELECT * FROM merged")
```

* 2. Remove the duplicate keys from the `patient_binge_drinking` table. 

```{r}
dbExecute(con, 
  "CREATE TEMPORARY TABLE temp AS
  SELECT DISTINCT ins_key, binge_drinking
  FROM patient_binge_drinking")
```

* 3. Verify that the duplicates were removed.

```{r}
dbGetQuery(con, "SELECT COUNT(*) AS patient_count FROM temp")
```

* 4. Now, extract it to a dataframe we will call `binge`.

```{r}
binge <- dbGetQuery(con, "SELECT* FROM temp")
```


* 5. Join `binge` with the bigger `merged` table we already have. Note that I am doing an `inner_join` - think about why this is! Also note that we are merging on the `ins_key`s rather than the patient keys as before!
**Case matters again!!!**

```{r}
final <- inner_join(merged, binge, by = "INS_KEY", relationship = "many-to-many") %>% 
  distinct(INS_KEY, .keep_all = TRUE)
```

* 6. Verify that our join was successful.

```{r}
str(final)
```

# 4 Aggregation to protect patient privacy

One very important way that we can protect patient privacy is by aggregating across zip codes, regions, states, or any other grouping variable. Given that we often have very small sample sizes (e.g., 12 people with disabilities in Wetumpka, Alabama) it is ostensible that we STILL would not be doing a sufficiently good job at protecting their identity. What can we do?

#### Question 14 [5 points]:

Think about at least two ways we might aggregate the data here to protect patient privacy and conform with HIPAA regulations. Make sure to describe what you would do in detail and why.

**Answer:**
There have few ideas I would consider to help: 
1. Generalize geography upward. Truncating ZIP+4 to 5 digits which the demo already did isn't enough. So I'd roll location up to 3-digit ZIP (with the 000 substitution for sparse areas) or to state, and drop CITY entirely.Remove the town, and you can't single out a person by where they live.
2. Suppress small cells (k-anonymity).Even after aggregating, a cross-tab can still produce thin cells, disability × employment × a small county might come down to one or two people.CMS commonly uses n < 11 as the suppression threshold, so I'd suppress any cell with 10 or fewer records.
3. Bin ages and apply the 90+ rule. Report age in bands (18–29, 30–44, …) instead of exact values, and collapse everyone over 89 into a single "90 or older" category. why 90+: exact ages at the high end are identifying because so few people are that old, and note we actually saw a max of 96.6 in MA. Pulling in our own result makes it concrete.

What I learned, The unifying principle worth stating explicitly for the points: aggregation only protects privacy when the resulting groups are large. 

# 5 Recommendations to the back-end development team

#### Question 15 [10 points]:

Draft a letter or brief report detailing the ways in which you would recommend that the database connection be improved to ensure patient privacy and data integrity. 

* Did you get the data you 'asked' for? (100,000 adult users from the United States)
* Were the data of high quality?
* Were there any problems with keys?
* Should you have had access to the production database? Why or why not?
* Are we in danger of accidentally committing a HIPAA violation?

Make sure to be detailed and specific. You are responsible for safe-guarding our patients' data - just as they are! The job doesn't fall to just one team. So be kind but firm in your recommendations.

**Answer:**

Let me walk through what I found.
1. Did we get what we asked for? Not quite. The first thing is that the pull isn't actually US-only. I'd asked for 100,000 adults in the United States, but the STATE field came back with 89 different values and when I looked closely, there are Canadian provinces in there (AB, BC) and Mexican states too (AG, CH), mixed right in with the real US codes. So somewhere the filter isn't catching country. The headcount of 100,000 looks right, and the ages seem to be adults (the youngest in Massachusetts was 20.5), but I'd want a clean re-pull limited to the 50 states plus DC before I trust the geography.
2. Were the data of high quality? When we trimmed the ZIP codes down to five characters, that logic assumed everyone was American — US ZIPs are five digits. But Canadian postal codes are six characters and alphanumeric, so the trim sliced them in half and left garbage behind (like T2E 8). So the international contamination didn't just break the spec; it quietly corrupted a downstream cleaning step that assumed everyone was American.
3. Were there problems with keys?Two. First, 81 insurance keys are shared across multiple patients, that one's expected (family policies), but it needs to be documented so nobody downstream treats INS_KEY as a per-person identifier. Second, and more serious: the patient_binge_drinking table has no PT_KEY at all, only INS_KEY. Because a policy can cover several people, there's no reliable way to attribute those records to individuals, and joining on INS_KEY alone fans out into a many-to-many mess. To get through this first pass I had to de-duplicate on the insurance key, which dropped 81 real patients (100,000 down to 99,919). That's a workaround, not a fix,  restore PT_KEY to that table at the source so it can be joined cleanly.
4. Should I have had access to the production database? No,We should seperate analysis database with production database. Analysts shouldn't be able to touch production, and shouldn't be receiving direct identifiers in the first place. The standard setup is a separate, read-only, de-identified analytical replica, kept apart from the live production database. That's what I'd ask for going forward.

5. Are we at risk of a HIPAA violation? Yes, as things stand. De-identification really has two layers, and this pull failed both. The first is removing the direct identifiers — names, DOB, email, phone, street, full ZIP. Those are PHI under Safe Harbor, and sending them to an analyst who only needs aggregates violates the minimum-necessary principle and would be a reportable breach if it ever leaked. The second layer is generalizing what's left: even after the obvious fields are gone, exact ages over 89 (we have a 96.6) and small-town counts can still re-identify someone by linking against public records. So both the obvious identifiers and the quasi-identifiers need handling before anything is released.


What I'd recommend:
1. Re-pull restricted to US states + DC, adults only.
2. Strip the direct identifiers at the source, before any extract reaches an analyst.
3. Stand up a read-only, de-identified analytical database for our work, separate from production database.
4. Restore PT_KEY to patient_binge_drinking so it joins cleanly without dropping patients.
5. Add validation on STATE and ZIPCODE so non-US and malformed values can't get in.



