---
title: 'Project 1: Biomedical & Clinical Informatics'
author: "Vicky (Xin) Guo"
date: "20 May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_crop: false
  pdf_document:
    toc: true
  word_document:
    toc: true
subtitle: 'Merrimack College DSE6630: Healthcare & Life Sciences Analytics'
bibliography: references.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      error = FALSE,
                      cache = TRUE,
                      cache.comments = TRUE,
                      size = 13)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Turn off scientific notation
options(scipen=999)

# Set seed
set.seed(98501)

# Clean, set up, and load
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
rm(list = ls(all = TRUE))

pacman::p_load(tidyverse,
               janitor,
               naniar,
               stringr,
               ggplot2, 
               RColorBrewer,
               gridExtra,
               gtsummary,
               factoextra,
               ggrepel,
               caret,
               missForest,
               mice,
               car,
               stargazer,
               glmnet,
               forcats,
               snakecase,
               randomForest  ## add this package
)
```

# 1 Introduction

We will pick up where we left off in Demo 1 to prepare you for analysis. Although this is not going to be as exhaustive as we would like to do if we were doing a complete project, our goal is to see this project go from __messy data__ $\rightarrow$ __prediction & interpretation__. After all, we need to give something to the patient advocacy watchdog that they can deploy!

## 1.1 Our Analysis Plan

Our goal is to see this analysis through to generate a predictive model that the advocacy group can use, and hopefully it will be a decent one. To ensure that it stands the best chance of being a good predictive model, our analysis plan includes all the key steps in the data science life cycle/project flow, including: __cleaning__, __exploration__, __feature selection__, __pre-processing__, __unsupervised machine learning exploration__, and __supervised modeling__. Whew! 

Our specific plan is to perform two big analyses in the hopes of having a nice deliverable to present to our client. The first will be a __segmentation analysis__, or clustering analysis, which will attempt to group our hospitals together on the basis of their attributes and their pneumonia-related readmissions. Thie will be done using a combination of PCA and $k$-means. Our second analysis will be to perform supervised, predictive analysis using an elastic net, which conveniently combines a penalized regression with variable selection so we can point our clients down a path of hospital attributes to focus on, as well as a model they can use to make future predictions.

## 1.2 Your objective.

Your objective is to work through this Project 1. You will have 31 Questions to answer as you work through this, just as with Project 1. Then, you will be able to choose your own adventure again to perform your own __clinical informatics__ analysis:

### Adventure 1. 
__[Minimal coding, focus is on understanding cleaning, pre-processing, and the ML analyses we performed here.]__
Choose ONE other condition (anything other than pneumonia) and run through the analyses again, using our same target variable. You will update your question, hypotheses, and predictions, but will be able to re-use all of the other code with very minor modifications. Credit will be based more on interpretation and not on coding. See more information on Adventure 1 [here](#adventure1).

### Adventure 2. 
__[More coding, but not as much, with a focus on the steps we've undertaken here and extending it to a new question & condition.]__
Choose a slightly more complicated analysis to undertake, for example, focusing on surgical interventions (`HIP-KNEE` & `CABG`) or heart-related conditions (`HF`, `AMI`, & `CABG`). The advantage of this analysis is that instead of looking at a single condition, you will be looking at the __average__ hospital reamdissions rate for the conditions you select. This might also enable us to say something far more informative to a subset of patients (e.g., those considering undergoing surgical interventions). See more information on Adventure 2 [here](#adventure2).

### Adventure 3. 
__[Most coding, but no cleaning or pre-processing, with an emphasis entirely on the machine learning analyses.]__
Continue with the pneumonia-related data that you have here, completely cleaned and pre-processed, and do two more unsupervised OR supervised analyses. For example, you could do another kNN using the $k$ we identify in our Segmentation Analysis here, or you could do a random forest to account for any possible non-linearity. You could even choose to do a Ridge and LASSO regression and compare those analyses to what we get out of an Elastic Net. Any of these are fine, but you must __explain your decision__ in your submission. See more information on Adventure 3 [here](#adventure3).

## 1.3 A Fork in the Road! {#section1.3}

This is the first of your choices in your adventure, and what you choose here could either lead to success or... __sudden death__ (dum, dum, dummmm...). Choose wisely, young data scientists!

## Option 1.3.1 
**Load data from the end of Demo 1, which includes the ordinal and frequency encoding you did (plus a little bonus cleaning to help).** 

Load my version of the fully cleaned, encoded, & ready-to-proceed dataset from the end of Demo 1. These data are not yet ready for analysis (!) but we would be picking up with that here.

```{r}
## Uncomment if this is your choice:
load("FY2024_data_files/pneumoniaAnalyzeFullyEncoded2024.Rdata")
dat2Analyze <- pneumoniaAnalyzeFullyEncoded
```

## Option 1.3.2 
**Load data from Demo 1 without ANY encoding, but all other tidying the same.**

```{r}
## Comment out if this is NOT your choice:
load("FY2024_data_files/pneumoniaAnalyzeNoEncoding2024.Rdata")
# dat2Analyze <- pneumoniaAnalyzeNoEncoding
```

#### **Question 1**: [1 point]
Which dataset do you choose and why?

**Answer:**
I chose Option 1.3.1, the fully cleaned and encoded dataset. The downstream analyses in this project (PCA, k-means clustering, and the regularized regressions via glmnet) all require numeric inputs, and several of the variables in this dataset — particularly the ComparedToNational_ columns and the payment/value-of-care categories — are inherently ordinal (e.g., Worse < Same < Better than the national rate). Using the pre-encoded version preserves that rank information, which would be lost if I let caret one-hot encode the factors automatically. Option 1.3.2 would have required me to redo the ordinal encoding before any of the analyses could run, which is duplicate work since the instructor has already provided a clean encoded version. That's the reason using already encoded dataset would be better choice. 


## 1.4 Helper code: Using `R` like an object-oriented language {#helperCode}

Many of you still may be early enough in your programming journeys that you haven't yet learned how to fully leverage the __objected-oriented__ nature of programming languages like `Python`. `R` is not _technically_ an object-oriented language in the same way, but we can hack it to function like one. That's exactly what we are going to do here by loading a __helper function__ directly from the `source` code. 

We went through all the trouble in the Demo to perform encoding and now we get to use it! You will use the `source()` function to load a helper function that I have written, using my answer to the ordinal encoding question of the Demo, so that you can do all your **ordinal** encoding in a single line of code. Neat, huh?

**BUT WHY** are we doing the encoding now if we just chose to bring in non-encoded data? Because we can do __ordinal encoding__ before or after data partitioning with impunity. It does not introduce data leakage to do so here, so we can do it in either order. This means that we can do it now simply for convenience.

**NOTE**: If you chose Option 1, the fully-encoded dataset, you can still run this chunk because there is a conditional in here that will not run your dataset since it is already encoded! But still make sure you understand the idea of `source` code because we will keep using it throughout the course!

```{r, results='hide'}
## This calls the code in the associated .R file
source(file = "doOrdinalEncoding.R")

## This checks to see if you loaded the already encoded dataset
if(!exists("pneumoniaAnalyzeFullyEncoded")) {
    ## This sets a list of PATTERNS to match for the columns to ordinal encode
    encodeList <- c("ComparedToNational_", 
                    "PaymentCategory", 
                    "ValueOfCareCategory",
                    "Score_Emergency department volume")

    ## This runs the function, which takes 3 arguments
    ## Open the source code to see what each argument does!
    dat2Analyze <- doOrdinalEncoding(df = dat2Analyze,
                                     encodeList = encodeList,
                                     quiet = FALSE)
}

dat2Analyze
```

# 2 Exploratory Data Analysis: Which of the possible target (outcome) variables should we use? (Continued)
**NOTE**: What follows is my (long-winded) answer to **Question 19** from the Demo.

### 2.1 Overview of our options

I am going to quickly break down the differences between the possible targets so we can make sure we really understand **what** they are measuring to allow us to make our most-informed selection of a target. Note: More information was found [in this PDF](https://qualitynet.cms.gov/files/5d0d3ae5764be766b0104c37?filename=Pneumo_ReadmMeasMethod.pdf). **This is really important for helping us choose the right one!**

Notice that I am also including a variable we didn't look at previously called `ComparedToNational_Hospital return days for pneumonia patients`. This is because this is our best assessment of how the hospital is performing with regard to pneumonia readmissions relative to a national average. However, even though I'm including it in the table, I think it's better to think of it as a **solid predictor**. It's telling us how __length of stay (LOS)__ in the hospital after pneumonia readmission stacks up compared to the national average. After a hospital readmission, it's generally better to have _fewer_ return days rather than more because it suggests better quality of care, more timely follow-up, and a greater chance of the patient recovering well at home. Since these are readmissions, having _more_ days than average could also suggest that the hospital missed the severity or didn't properly plan for the patient's recovery at home and released them too early. 

You may be wondering - could we use it as a target? Sure - but it's not for the question we were posed by our stakeholder! Note that, if we did, we would need to perform a multinomial classification analysis. So, instead, let's use this category to help us contextualize our hospitals as we choose between `Predicted`, `Observed`, and `Expected` readmission rates, as well as `Excess Readmission Ratio`. 

```{r, echo = FALSE, collapse=TRUE}
dict <- tribble(
~`Possible Target`, ~`Description`,
"ComparedToNational_Hospital return days for pneumonia patients", "Hospital return days measures the number of days patients spent back in the hospital (in the emergency department, under
observation, or in an inpatient unit) within 30 days after they were first treated and released for pneumonia. Reported as compared to the national average, such that 'below average' is better and 'above average' is worse.", 
"ExpectedReadmissionRate", "The expected number of readmissions in each hospital is estimated using its patient mix and an average hospital-specific intercept. It is thus indirectly standardized to other hospitals with similar case and patient mixes.",
"PredictedReadmissionRate", "The number of readmissions within 30 days predicted based on the hospital’s performance with its observed case mix. The predicted number of readmissions is estimated using a hospital-specific intercept, and is intended to reflect the annual expected performance of the hospital given its historical case and patient mix and performance.",
"ExcessReadmissionRatio", "The ratio of the predicted readmission rate to the expected readmission rate, based on an average hospital with similar patients. Performance is compared against a ratio of one, such that below one is better and above one is worse in terms of readmission rates.",
"observed_readmission_rate", "Our calculation of the observed number of pneumonia-related readmissions within 30 days found by dividing the number of readmissions observed for the hospital during the given period by the number of discharges for that same period, and multiplied by 100 to put it on the same scale as the Predicted and Expected Readmission Rates. It reflects a true observation as reported by the hospital during that period, but is not adjusted for case mixes or prior information for that hospital. Thus, this is a crude, unadjusted value."
)

dict %>% 
  kable(
    format = "html",
    caption = "Table 1. List of possible target variables from the pneumonia-related hospital readmission data. Which one to choose?") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
  )
```

### 2.2 Summary statistics of the possible targets

Now, I know I didn't ask you to do it, but I will actually begin my assessment of which target is most appropriate by constructing a quick summary table. In my assessment of what is an appropriate target, 

```{r, echo = FALSE}
sliceData <- dat2Analyze %>% 
  select(`ComparedToNational_Hospital return days for pneumonia patients`, 
         PredictedReadmissionRate,
         observed_readmission_rate, 
         ExcessReadmissionRatio,
         ExpectedReadmissionRate) %>% 
  rename(`National Comparison Return Hospital Days` = `ComparedToNational_Hospital return days for pneumonia patients`,
         `Predicted Readmission Rate` = PredictedReadmissionRate, 
         `Observed Readmission Rate` = observed_readmission_rate, 
         `ExcessReadmissionRatio` = ExcessReadmissionRatio,
         `Expected Readmission Rate` = ExpectedReadmissionRate) %>% 
  mutate(`National Comparison Return Hospital Days` = ifelse(is.na(`National Comparison Return Hospital Days`), "Unknown", 
                                                      ifelse(`National Comparison Return Hospital Days` == 1, "Better than average",
                                                      ifelse(`National Comparison Return Hospital Days` == 0, "Same as average", "Worse than average")))) %>% 
  mutate(`National Comparison Return Hospital Days` = factor(`National Comparison Return Hospital Days`, levels = c( "Better than average", "Same as average", "Worse than average", "Unknown")))

sliceData %>% 
  tbl_summary(statistic = list(
      all_continuous() ~ "{mean} ({sd}), {median}",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ), missing_text = "(Missing)") %>% 
  modify_caption("**Table 2. Possible target measure summaries**") %>%
  bold_labels()
```

We can see right away that we lose nearly **600 more data points** if we opt to use our calculated `observed_readmission_ratio` (NOTE: nearly 500 if using the 2025 FY data), which is going to be a by-product of the fact that the `Predicted` and `Expected` readmission ratios are using prior information and information about patient mixes to generate these values. Our `observed_readmission_ratio` is crude and unadjusted any way - which could present problems when it comes to true comparability. **So, off the bat it feels as if the constructed target is possibly not a good choice despite its simplicity for stakeholders to understand.** 

### 2.3 Comparing the four possible metrics to`ComparedToNational_Hospital return days for pneumonia patients`

Although `ComparedToNational_Hospital return days for pneumonia patients` is very compelling, it's better as a **predictor** rather than a target because it gets at an aspect of hospital performance: once readmitted due to pneumonia, how long (as compared to the national average) are patients spending in the hospital? This means it makes a really nice predictor to use to _compare our four possible choices for pneumonia-related hospital readmissions_. We are looking for a clear separation in our ideal target between hospitals that have lower readmission rates AND also tend to do better than the national average at healing their readmitted patients!

So, this makes a really nice variable I could choose to **facet** plots, which means creating sub-plots that show up as panels. The thing is, if I know I want to **facet** plots, the first thing I have to do is reshape the data I want from *wide* into *long* format using, for example, the `pivot_longer()` function (the complement of `pivot_wider()` that we used in the Demo).

I don't have to store it as a separate data frame (i.e., I can just pipe it directly into `ggplot()`!) but here I am choosing to so we can more easily make sure it's reshaped correctly along the way. Thus, I am making a long-format version of the possible target variables called `longTargets` so that I can better decide which target to choose.

```{r, echo = FALSE}
## Make the long-format version of just the possible target variables, with some cleanup of the names, etc. for nicer graphs
longTargets <- dat2Analyze %>%
  ## Make a shorter name
  rename(NatlComparisonHospitalDays = `ComparedToNational_Hospital return days for pneumonia patients`,
         `Predicted Readmission Rate` = PredictedReadmissionRate,
         `Excess Readmission Ratio` = ExcessReadmissionRatio,
         `Observed Readmission Rate` = observed_readmission_rate, 
         `Expected Readmission Rate` = ExpectedReadmissionRate) %>% 
  ## Flexible handling depending on which dataset students select; recodes into better, same, and worse
  mutate(NatlComparisonHospitalDays = case_when(
    (NatlComparisonHospitalDays == 1  | NatlComparisonHospitalDays == "Fewer Days Than Average per 100 Discharges") ~ "Better", 
    (NatlComparisonHospitalDays == -1 | NatlComparisonHospitalDays == "More Days Than Average per 100 Discharges") ~ "Worse", 
    (NatlComparisonHospitalDays == 0  | NatlComparisonHospitalDays == "Average Days per 100 Discharges") ~ "Same", 
    .default = "Unknown"),
  NatlComparisonHospitalDays = factor(NatlComparisonHospitalDays, levels = c("Better", "Same", "Worse", "Unknown"))) %>% 
  ## Grab just the five features we want so we can compare them
  select(contains(c("readmission", "NatlComparisonHospitalDays"))) %>% 
  ## Pivot longer
  pivot_longer(-c(5), names_to = "Variable", values_to = "Value")
```

I mentioned `violin` and `density` plots specifically in Question 19 of Demo 1, as well as faceting. I am going to show you two ways to explore these data with these types of plots and you can decide which one(s) you find most informative here. You also may have thought of a better way to display the data too!

#### 2.3.1 Example with `geom_violin()`:
Here I am leaving the `NA` data (which I relabelled as `Unknown`) to get a sense for the degree of missingness.

```{r, fig.cap = "Figure 1. Readmission Rates vs. readmitted hospital LOS", fig.width=8, fig.height=6, echo = FALSE}
longTargets %>% 
  ggplot(aes(x = Value, y = NatlComparisonHospitalDays, fill = NatlComparisonHospitalDays)) +
  geom_violin(alpha = 0.7) +
  labs(title = "Which target to use?",
        subtitle = "By readmitted LOS in hospital compared to national average",
        x = "",
        y = "Rate or Score",
        fill = "Pneumonia-related hospital days (compared to national average):") +
  scale_fill_manual(values = c("gold", "cadetblue", "maroon", "darkgreen"),
                    labels = c("Better", "Same", "Worse", "Unknown")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size=9),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        title = element_text(size = 12, color = "maroon")) + 
  facet_wrap(~Variable, 
             scales = "free",
             ncol = 2)
```

#### **Question 2A**: [0.5 points]
If you have never encountered [violin plots](https://en.wikipedia.org/wiki/Violin_plot) before, they are honestly a wonderful way to assess the **symmetry** and **skew** of a distribution while more easily enabling comparisons between groups. They're effectively a hybrid between a histogram and a boxplot, and we can even opt to overlay the boxplot quartiles in the plots as well, which is particularly handy!

**Your task** is to do just that by adding `geom_boxplot(width = 0.1)` (yes, just that one line!) in the appropriate place. Also change `include = FALSE` to `include = TRUE` in the R chunk header so that it shows up when you knit!

> Make sure to comment your line to call attention to it:

```{r, fig.cap = "Figure 2. Readmission Rates vs. readmitted hospital LOS", fig.width=8, fig.height=6, echo = FALSE, include=TRUE}
longTargets %>% 
  ggplot(aes(x = Value, y = NatlComparisonHospitalDays, fill = NatlComparisonHospitalDays)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1) +
  labs(title = "Which target to use?",
       subtitle = "By readmitted LOS in hospital compared to national average",
        x = "",
        y = "Rate or Score",
        fill = "Pneumonia-related hospital days (compared to national average):") +
  scale_fill_manual(values = c("gold", "cadetblue", "maroon", "darkgreen"),
                    labels = c("Better", "Same", "Worse", "Unknown")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size=9),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        title = element_text(size = 12, color = "maroon")) + 
  facet_wrap(~Variable, 
             scales = "free",
             ncol = 2)
```

#### **Question 2B**: [0.5 points]
Now that you can compare these groups more easily, what we're looking for is a balance between **clear difference** between hospital ordinality relative to the national average (i.e., "Are better hospitals really better in pneumonia readmission rates?").

Which target(s) appear the most distinct to you?

**Answer: **
1. The Predicted Readmission Rate is by far the most distinct target. Its medians follow a clear ordered pattern — Better hospitals sit around ~15, Same around ~16, and Worse around ~17–18 — with the boxes showing only modest overlap between adjacent categories. This is exactly the "Better hospitals really are better" signal we want a target to carry.
2. The Observed Readmission Rate also shows ordered separation in the medians (Better < Same < Worse), but the boxes are much wider and there are many outliers extending into the 30s. The signal is there, but it's noisier because Observed isn't risk-adjusted for patient case mix.
3. The Excess Readmission Ratio and Expected Readmission Rate are essentially indistinct. ERR squashes all categories near 1.0 with heavily overlapping boxes, and Expected shows nearly identical medians (~16) across Better, Same, and Worse — which makes sense because Expected reflects patient case mix rather than hospital performance, so it shouldn't track hospital quality categories.

In Summary, Predicted Readmission Rate is therefore the strongest candidate as a target because it provides the cleanest ordinal separation between hospital performance categories.


#### 2.3.2 Example with `geom_density()`:
We could also choose to overlay density plots instead of violin or boxplots, as this can sometimes help us see subtler separation of groups than we can see otherwise. This time, I am also going to ignore the "Unknown" (i.e., missing) class of hospitals to prevent things from getting too messy!

```{r, "Figure 2. Readmission Rates vs. readmitted hospital LOS", warning=FALSE, message=FALSE, echo=FALSE}
longTargets %>% 
  filter(NatlComparisonHospitalDays != "Unknown") %>% 
  ggplot(aes(x = Value, fill = NatlComparisonHospitalDays)) +
  geom_density(alpha = 0.45) +
  labs(title = "Which target to use?",
       subtitle = "By readmitted LOS in hospital compared to national average",
        x = "",
        y = "Rate or Score",
        fill = "Pneumonia-related hospital days (compared to national average):") +
  scale_fill_manual(values = c("gold", "cadetblue", "maroon", "darkgreen"),
                    labels = c("Better", "Same", "Worse", "Unknown")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size=9),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        title = element_text(size = 12, color = "maroon")) + 
  facet_wrap(~Variable, 
             scales = "free",
             ncol = 2)
```

#### **Question 3**: [0.5 points]
Does this view confirm or change your decision from Question 3?
**Answer:**
This view confirms my decision from Question 2. The density plot makes the comparison even clearer: in the Predicted Readmission Rate panel, the three category distributions have visibly separated peaks (Better ~14–15, Same ~16, Worse ~17–18) with relatively narrow spreads. The Expected Readmission Rate panel shows all three distributions stacked essentially on top of each other, reinforcing that Expected reflects patient case mix rather than hospital quality. Observed Readmission Rate still shows ordered peak separation, but with much wider tails extending past 30, indicating more within-category noise from unadjusted patient-mix variation. The Excess Readmission Ratio shows only modest separation tightly clustered around 1.0. Predicted Readmission Rate remains the clearest, cleanest target for predicting hospital readmission performance.


### 2.4 What is the relationship of pneumonia-related hospital readmissions to the hospital performance metrics?

Something else we could do to help us make a decision is compare across predictors to look for the most striking relationships among the proposed target (readmission rate), how well compared to national average hospitals are performing AFTER readmitting pneumonia patients (how long after readmission they are in the hospital), and relationships among other hospital-performance predictors, like Medicaid spending or pneumonia-related deaths. 

So, to do this, I'm going to make another long-format dataset and use this to make a bunch of faceted scatterplots where I put one of the readmission rates (let's started with `Predicted Readmissions`) and look for strength in relationships. Note that doing it this way, through plotting, is a **heuristic way** of choosing a target variable and it isn't the only way to do it!

```{r, echo = FALSE}
longScore <- pneumoniaAnalyzeNoEncoding %>%
  rename(NatlComparisonHospitalDays = `ComparedToNational_Hospital return days for pneumonia patients`) %>% 
  select(contains(c("Score", "NatlComparisonHospitalDays", "readmission"))) %>% 
  ## Flexible handling depending on which dataset students select; changes to better, same, worse
  mutate(NatlComparisonHospitalDays = case_when(
    (NatlComparisonHospitalDays == 1  | NatlComparisonHospitalDays == "Fewer Days Than Average per 100 Discharges") ~ "Better", 
    (NatlComparisonHospitalDays == -1 | NatlComparisonHospitalDays == "More Days Than Average per 100 Discharges") ~ "Worse", 
    (NatlComparisonHospitalDays == 0  | NatlComparisonHospitalDays == "Average Days per 100 Discharges") ~ "Same", 
    .default = NA),
  NatlComparisonHospitalDays = factor(NatlComparisonHospitalDays, levels = c("Better", "Same", "Worse", "Unknown")),
  `Score_Emergency department volume` = as.numeric(case_when(`Score_Emergency department volume` == "very high" ~ "5",
                                                         `Score_Emergency department volume` == "high" ~ "4",
                                                         `Score_Emergency department volume` == "medium" ~ "3", 
                                                         `Score_Emergency department volume` == "low"~ "2", 
                                                         `Score_Emergency department volume` == "very low" ~ "1",
                                                        .default = `Score_Emergency department volume`))) %>% 
  pivot_longer(1:12, names_to = "Variable", values_to = "Value") %>% 
  drop_na() %>% 
  mutate(Variable = gsub("Score_", "", Variable)) %>% 
  mutate(Variable = case_when(
    Variable == "Average (median) time patients spent in the emergency department before leaving from the visit A lower number of minutes is better"
                ~"Median time (minutes) patients spent in ED",
    Variable == "CMS Medicare PSI 90: Patient safety and adverse events composite" ~ "Composite patient safety", 
    Variable == "Abdomen CT Use of Contrast Material" ~ "Abdomen CT w/ contrast", 
    Variable == "Perioperative pulmonary embolism or deep vein thrombosis rate" ~ "Perioperative pulmonary embolism/DVT",                     
    Variable == "Left before being seen" ~ "% left ED before being seen", 
    .default = Variable))

```

#### **Question 4**: [1 point]

Change the __y-axis__ out for _each of the four readmission rates. `Predicted Readmission Rates` has been started for you. Re-run the chunk and compare the results. **FOR FULL CREDIT** you must leave the graph on the choice you feel most strongly about AND update the title of the graph with which of the four possible readmission rates you chose. Explain below why you chose this one. 

**Answer:**
I chose Predicted Readmission Rate as the y-axis after trying all four. The comparison made the choice clear:

1. Expected Readmission Rate showed essentially no separation between Better, Same, and Worse hospitals — all three lines overlapped almost completely in every panel. This is consistent with Expected being a case-mix benchmark rather than a hospital-quality metric.
2. Excess Readmission Ratio carried similar information to Predicted but compressed into a 0.8–1.4 band, making slopes hard to read visually. Its very narrow variance (SD = 0.06) squashes everything.
3. Observed Readmission Rate showed clean vertical separation between performance categories, but the slopes within each category were largely parallel and noisy — reflecting that Observed mixes hospital-quality signal with patient case-mix noise.
4. Predicted Readmission Rate was the only target that produced both clear category separation AND interpretable, differentiated slopes across predictors. The lines fan out and cross rather than running parallel, indicating that the relationship between predictors and readmissions actually differs by hospital performance category — that's modelable signal.


**HINT 1**: What does higher medicare spending per patient tell you about the underlying population of a hospital?

**HINT 2**: Venous thrombolisms and deep vein thromboses (DVTs) are often "sneakier" conditions that have "silent" symptoms until the patient is in crisis. What would having higher scores at providing prophylaxis for these conditions potentially suggest about a hospital's performance?


```{r, fig.width=7, fig.height=8, echo = FALSE, message=FALSE, fig.cap="Figure 4. Readmission rate by hospital performance and readmitted patient LOS"}
longScore %>%  
       #    `Predicted Readmission Rate` = PredictedReadmissionRate,
       #  `Excess Readmission Ratio` = ExcessReadmissionRatio,
       #  `Observed Readmission Rate` = observed_readmission_rate, 
       #  `Expected Readmission Rate` = ExpectedReadmissionRate)
  ggplot(aes(y = PredictedReadmissionRate,  ## <-- CHANGE THIS!
             x = Value)) +
    geom_point(alpha = 0.25) + 
    geom_smooth(aes(fill = as.factor(NatlComparisonHospitalDays), color = as.factor(NatlComparisonHospitalDays)), 
                method = "lm", 
                se = T, 
                show.legend = T) + 
    ## CHANGE THE TITLE FOR FULL CREDIT!
    labs(title = "Relationship of Predicted Readmission rate in this pneumonia-related hospital \nreadmissions to each performance metric",
        x = "Rate or Score",
        y = "Predicted Pneumonia-related Readmissions",
        color = "Pneumonia-related Readmission Compared to National Average") +
  scale_fill_manual(values = c("gold", "cadetblue", "maroon")) + 
  scale_color_manual(values = c("gold", "cadetblue", "maroon"),
                    labels = c("Better", "Same", "Worse")) + 
  guides(fill = "none", color = guide_legend(override.aes = list(fill=NA))) +    ## To override the filled SE boxes from geom_smooth() in the legend
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size=8),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        title = element_text(size = 12, color = "maroon")) + 
  facet_wrap(~Variable, 
             scales = "free_x",
             ncol = 2)
```

**HINT 3**: Here's how I would interpret the __relationship between PREDICTED pneumonia readmissions and pneumonia death rates, looking at whether hospitals performed better, same, or worse in terms of length-of-stay after readmission__. What I notice immediately is a moderate slope for hospitals with worse readmitted lengths of stay (LOS) [maroon line] compared to same [blue line] and especially better [yellow line]. This relationship is negative, implying that as the pneumonia-related death rate increases, readmissions __goes down__. Oh! So what this implies is that hospitals with higher deaths from pneumonia tend to have lower readmissions - and this is especially true for the ones who are have longer (worse than average) readmitted patient stays. In other words - these hospitals are __ probably releasing their pneumonia patients too soon from their first hospitalization__, leading to higher readmissions, longer readmitted stays, and even higher pneumonia-related deaths! 

### 2.5 What is the relationship of pneumonia-related hospital readmissions to the patient satisfaction metrics?

Now let's do a similar analysis but for the `HcahpsLinear...` variables. The __HCAHPS (Hospital Consumer Assessment of Healthcare Providers and Systems)__ is a standardized survey that measures adult inpatient perceptions of hospital care. It's a national, publicly reported survey mandated by the Centers for Medicare & Medicaid Services (CMS). Because it focuses on aspects of hospital care that are important to patients, it is often used to improve patient satisfaction and hospital quality. 

```{r, echo = FALSE}
longHCAHPS <- pneumoniaAnalyzeNoEncoding %>%
  rename(NatlComparisonHospitalDays = `ComparedToNational_Hospital return days for pneumonia patients`) %>% 
  select(contains(c("HcahpsLinear", "NatlComparisonHospitalDays", "readmission"))) %>% 
  ## Flexible handling depending on which dataset students select; changes to better, same, worse
  mutate(NatlComparisonHospitalDays = case_when(
    (NatlComparisonHospitalDays == 1  | NatlComparisonHospitalDays == "Fewer Days Than Average per 100 Discharges") ~ "Better", 
    (NatlComparisonHospitalDays == -1 | NatlComparisonHospitalDays == "More Days Than Average per 100 Discharges") ~ "Worse", 
    (NatlComparisonHospitalDays == 0  | NatlComparisonHospitalDays == "Average Days per 100 Discharges") ~ "Same", 
    .default = NA),
  NatlComparisonHospitalDays = factor(NatlComparisonHospitalDays, levels = c("Better", "Same", "Worse", "Unknown"))) %>% 
  pivot_longer(1:10, names_to = "Variable", values_to = "Value") %>% 
  drop_na() %>% 
  mutate(Variable = gsub("HcahpsLinearMeanValue_", "", Variable))
```


#### **Question 5**: [1 point]

Again, you task is to change the __y-axis__ out for _each of the four readmission rates. `Predicted Readmission Rates` has been started for you. Re-run the chunk and compare the results. **FOR FULL CREDIT** you must leave the graph on the choice you feel most strongly about AND update the title of the graph with which of the four possible readmission rates you chose. Explain below why you chose this one. 

**Answer:**
I picked Predicted Readmission Rate again. Same pattern as Q4 — Expected showed almost no separation between the three performance categories (all three lines stacked), Excess Ratio compressed everything into a tiny 0.8–1.4 band that made the slopes hard to read, and Observed gave good separation but with wider, noisier distributions because it isn't risk-adjusted. Predicted hit the right balance: clear vertical separation between Better/Same/Worse and slope differences I could actually interpret.

The story is also cleaner here than in Q4. Nearly every HCAHPS panel shows a negative slope — higher patient satisfaction tracks with lower predicted readmissions across the board. The steepest negative slopes are in the communication-related metrics:

1. Doctor communication — strong downward slope, especially for Worse (maroon)
2. Nurse communication — same pattern, strong negative slope
3. Communication about medicines — clear negative slope
4. Discharge information — clear negative slope
5. Overall hospital rating — moderate negative slope, steepest for Worse

This matches HINT 1 — when doctors and nurses communicate well, patients understand their meds, recognize warning signs, and actually follow discharge instructions, so they don't end up back in the hospital. Worse-LOS hospitals show the steepest improvement here because they have the most room to gain.

The flattest panels are Cleanliness and Quietness, which matches HINT 2. Pneumonia readmissions happen weeks after discharge, driven by patient health and follow-up care — not by whether the hospital was clean or quiet during the original stay. Cleanliness would affect hospital-acquired infections, but those show up immediately, not as 30-day pneumonia readmissions.

The Overall hospital rating panel matches HINT 3 exactly: steepest negative slope for Worse-LOS hospitals, moderate for Same, gentlest for Better. Hospitals already doing well don't have much room to improve, but a high overall rating substantially helps hospitals already struggling — possibly reflecting the "we got it wrong the first time, let's make sure we get it right the second time" effect where a longer readmitted LOS at a high-quality hospital actually represents better care.


**HINT 1**: What does higher nurse and doctor communication mean for improved patient outcomes, including readmission?

**HINT 2**: Why might cleanliness not have much effect here even though pneumonia is a communicable disease?

```{r, fig.width=7, fig.height=8, echo = FALSE, message=FALSE, fig.cap="Figure 5. Readmission rate by patient satisfaction and readmitted patient LOS"}
longHCAHPS %>%
       #    `Predicted Readmission Rate` = PredictedReadmissionRate,
       #  `Excess Readmission Ratio` = ExcessReadmissionRatio,
       #  `Observed Readmission Rate` = observed_readmission_rate, 
       #  `Expected Readmission Rate` = ExpectedReadmissionRate)
  ggplot(aes(y = PredictedReadmissionRate,  ## <-- CHANGE THIS!
             x = Value)) +
    geom_point(alpha = 0.25) + 
    geom_smooth(aes(fill = as.factor(NatlComparisonHospitalDays), color = as.factor(NatlComparisonHospitalDays)), 
                method = "lm", 
                se = T, 
                show.legend = T) + 
    ## CHANGE THE TITLE FOR FULL CREDIT!
    labs(title = "Relationship of Predicted in this  pneumonia-related hospital \nreadmissions to each patient satisfaction metric",
        x = "Rate or Score",
        y = "Predicted Pneumonia-related Readmissions",
        color = "Pneumonia-related Readmission Compared to National Average") +
  scale_fill_manual(values = c("gold", "cadetblue", "maroon")) + 
  scale_color_manual(values = c("gold", "cadetblue", "maroon"),
                    labels = c("Better", "Same", "Worse")) + 
  guides(fill = "none", color = guide_legend(override.aes = list(fill=NA))) +    ## To override the filled SE boxes from geom_smooth() in the legend
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size=8),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        title = element_text(size = 12, color = "maroon")) + 
  facet_wrap(~Variable, 
             scales = "free_x",
             ncol = 2)
```

**HINT 3**: Here's how I would interpret the __relationship between PREDICTED pneumonia readmissions and overall hospital rating, looking at whether hospitals performed better, same, or worse in terms of length-of-stay after readmission__. What I notice immediately is a pretty substantial negative slope for hospitals with worse readmitted lengths of stay (LOS) [maroon line], moderate negative slope for hospitals with the same readmitted LOS [blue line], and slightly decreasing slope for hospitals with better readmitted LOS better [yellow line]. So what this implies is that hospitals with high ratings tend to have lower pneumonia-related readmissions, even among the hospitals with worse readmitted length-of-stays. This relationships is strongest when hospitals have longer LOS; maybe having a long readmission LOS stay **actually implies a form of quality**, like "Well, we got it wrong the first time, let's make darn sure we get it right the second?" But hospitals that were already getting it pretty right don't have as steep of slope because they were already doing well at readmissions to begin with, hence the almost flat line.


#### **Question 6**: [2 points]
What relationships do you find most compelling from these scatterplots and why? Also, name one outstanding question you have at this point, whether it is about the data, the data collection, or something more speculative.

**Answer:**
The two relationships I found most compelling were Medicare spending per patient (Q4) and the communication-related satisfaction metrics (Q5). Medicare spending had the cleanest positive slope across all three LOS categories, steepest for Worse hospitals — higher spending reflects sicker patients, and struggling hospitals are most sensitive to that complexity. That tells me patient acuity is a structural driver of readmissions hospitals can't easily change.
The communication metrics (doctor, nurse, medicines, discharge information) had the steepest negative slopes with Predicted readmissions, especially for Worse hospitals. Communication is one of the few hospital-level factors that's both modifiable AND clearly tied to outcomes after discharge — it's the most actionable signal in the data. The contrast with cleanliness and quietness (which were flat) reinforces this: what patients leave understanding matters more than the stay itself.

Worse-LOS hospitals consistently sit above the others and respond more steeply to nearly every predictor. Is that because they have more room to improve, or because they're structurally different in ways we're not capturing — rural vs. urban, teaching status, nurse-to-patient ratios? The current data shows *that* Worse hospitals are different, not *why*. Knowing the answer would change whether we recommend process interventions (training, communication protocols) or structural ones (staffing, resources).

Although there is certainly more EDA we could do, we are going to wrap that up here for the sake of moving on - but, as always, I encourage you to think about or even explore what else you would like to know about your data that we did not do!

### 2.5 Final remarks on target selection.

Going forward, I have selected `PredictedReadmissionRate` as our target. In a more thorough analysis, we would likely want to consider:

1. Circling back to our stakeholder to try to refine our question or target using domain knowledge

2. Perform experiments with our models to find the best predictive model since that is our ultimate goal. 

However, given that we do not have the option or time for that, let me explain my choice. It's the **best balance** of distinguishing hospitals with better and worse readmitted hospital stays, while still providing an __adjusted, hospital-specific__ measure. Adjusted values will generally always be a better choice than the crude, even if our crude measure captures something interesting about the data, as it did here, or if it's easier for our stakholder to comprehend. Our crude measure, `observed_readmission_rate` did the __best job__ of recapitulating the readmitted hospital LOS (`ComparedToNational_Hospital return days for pneumonia patients`), but if that's really our focus, wouldn't it then be better to switch from a readmission rate to that as our target? Probably. 

If you still feel unsure about what is the best choice, you're not alone - so do I! This is where, because we are working with an __imagined stakeholder__, we must make the best decision we can given the question we defined in the Demo. I would defend my choice of `Predicted Readmission Rate` for three key reasons:

* It has been adjusted to be hospital-specific, using a hospital-specific intercept and historic data about the hospitals' performances

* We don't lose nearly 600 data points to missingness.

* Our imagined stakeholder, and the question we set, was about __readmission rate__ so even though __readmitted return hospital days__ might also be a similarly viable candidate, it doesn't reflect our imagined stakeholder's request. Note that the `ExpectedReadmissionRate` is standardized for similar hospitals, and may not truly reflect a given hospital; the `ExcessReadmissionRatio`, while informative, was the least likely to reflect the distinctiveness in hospital return days, which is still of interest to us. 

But remember that what matters most is that we are __matching our question__. 

#### **Question 7**: [1 point]

Do you agree with my choice of target? Why or why not? (You are free to disagree!)

**Answer:**
I agree with the choice of PredictedReadmissionRate. Across the EDA — the violin and density plots, both sets of relationship scatterplots, and the discussion we have been through — Predicted consistently came out as the best balance we should use for this use case. Let's go a little bit deeper: 

1. Hospital-specific adjustment matters, for example I could see in the observed plots where the wide, parallel slopes reflected unadjusted variability rather than hospital-quality signal. 
2. Using PredictedReadmissionRate whichmatching the stakeholder's question (readmission rate, not LOS or comparison category), compare other direction.
3. The ~600 data points we save vs. Observed is meaningful given how much of the data is already missing.


### 2.6 Bias and Limitation

#### **Question 8**: [2 points]
All studies have bias (introduced or systemic error) and limitations (failure to fully explain something). We could go more deeply into these concepts, but for now I want you to take a stab and just brainstorming either ways we might have bias OR the limitations of using this target variable. You do not need to answer both, but you're welcome to do so. You can read more about bias [here](https://www.ncbi.nlm.nih.gov/books/NBK574513/) or see some deeper explanations of limitations [here](https://www.aje.com/arc/how-to-write-limitations-of-the-study/). You do not need to write a lot; just 2-3 sentences. I am simply asking you to pause and reflect on our choices here before we proceed. 

**Answer:** 

There are two issues worth flagging. 
First, bias: PredictedReadmissionRate inherits whatever's baked into CMS's risk-adjustment model. If CMS didn't adjust for variables like socioeconomic status or neighborhood resources, hospitals serving disadvantaged populations could look worse than they really are — meaning we'd be modeling structural inequity as if it were a hospital-quality signal. 
Second, limitations: we're modeling a modeled value, not a raw outcome, which makes our predictions only as good as CMS's adjustment and harder to communicate to stakeholders who care about actual patient experience rather than statistical estimates.

# 3 Pre-processing and Feature Selection

The time has finally come to officially move __back__ into pre-processing and feature selection so that we can finally build some models. I want you to take note of the order in which I'm doing things here; the flow can change from project to project, but generally it's a good idea to get your dataset to a state of __manually curated features__ before making your data partition.

## 3.1 Further feature selection and refinement.

### 3.1.1 Dropping the targets we aren't planning to use. [Manual Method]

It's time to say goodbye to the features we know would be far too redundant or non-informative given our target of `PredictedReadmissionRate`. We don't need any of the `Sample_` columns, for example; these are merely sample sizes. Useful if we were doing standardization, but not necessary for us right now. We also are dropping the other possible target features that are too redundant with `PredictedReadmissionRate`.

#### **Question 9A**: [0.5 points]
Add your comments to the code as indicated to explain what I was doing at each of the steps.

> Your answer in the comments.

```{r}
dat2Analyze <- dat2Analyze %>% 
  ## Drop the other readmission target candidates (redundant with PredictedReadmissionRate)
  ## plus all Sample_ and NumberOfPatients columns ( sample sizes, not predictive of readmissions)
  ## and NumberSurveysCompleted (also a sample-size variable from the HCAHPS surveys)
  select(-ExpectedReadmissionRate, 
         -ExcessReadmissionRatio,
         -observed_readmission_rate, 
         -contains(c("Sample_", "NumberOfPatients")),
         -NumberSurveysCompleted) %>% 
  ## Rename a few long/unwieldy column names to shorter, more readable versions
  
  rename(`Median time (minutes) patients spent in ED` = 
         `Score_Average (median) time patients spent in the emergency department before leaving from the visit A lower number of minutes is better`,
         `Composite patient safety` = `Score_CMS Medicare PSI 90: Patient safety and adverse events composite`,
         `ComparedToNational_Composite patient safety` = `ComparedToNational_CMS Medicare PSI 90: Patient safety and adverse events composite`) %>% 
  ## Move PredictedReadmissionRate to the first column position for convenience
  ## (using everything() to keep all other columns in their existing order)
  select(PredictedReadmissionRate, everything())

## Strip common prefixes (Score_, HcahpsLinearMeanValue_, _Payment) from all column names
## to make them shorter and cleaner without losing meaning
colnames(dat2Analyze) <- gsub('Score_|HcahpsLinearMeanValue_|_Payment', '', colnames(dat2Analyze))
```


### 3.1.2 Dropping any features with near-zero variance. [Automated Method]

Next, we are going to take advantage of the `nearZeroVar()` function that is part of the `caret` package. We will drop any of the columns with near-zero variance. This is a way of automating (and thus standardizing) the process for choosing to drop non-informative columns due to zero or nearly zero variance.

#### **Question 9B**: [0.5 points]
If you are not familiar yet with the concept of __zero or near-zero variance__, do a little digging. Based on what you read, try to explain for your naive advocacy group stakeholders, why it's important to remove the features with little variance.

**Answer:**
A feature with "near-zero variance" is one where almost every hospital in the dataset has the same value — there's essentially no variation in it. For example, if 95%+ of hospitals scored exactly the same on a particular metric, that metric can't help us tell good hospitals apart from bad ones, because they all look identical on that measure.

For our advocacy group stakeholders, the practical reason to drop these features is simple: a variable that doesn't *vary* can't *explain* anything. If we want to identify which hospital characteristics are associated with higher or lower readmission rates, we need characteristics that genuinely differ across hospitals. Including features where everyone looks the same just adds noise to the analysis without giving us any new insight, and it can actually destabilize our statistical models by making the math less reliable. Dropping these features keeps the analysis focused on the variables that actually distinguish hospitals from one another.

In our dataset, three variables were dropped on this basis: two "ComparedToNational_" categorical variables for surgical complications (postoperative respiratory failure and perioperative DVT), where most hospitals don't perform enough surgeries for CMS to compute a meaningful comparison, and PaymentCategory for pneumonia patients, where most hospitals fall into the "as expected" payment tier with very few outliers in either direction.

```{r, echo = F}
## Identify the columns with near-zero variance
zero_var_df <- dat2Analyze[, nearZeroVar(dat2Analyze)]

## Print the columns
data.frame(names(zero_var_df)) %>% 
  rename(`Variables Dropped` = names.zero_var_df.) %>% 
  kable(
    format = "html",
    caption = "Table 3. Columns dropped due to near-zero variance.") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
)

## Drop them from the dataset:
dat2Analyze <- dat2Analyze[, -nearZeroVar(dat2Analyze)]
```

**Take note** of the number of features here.

#### **Question 9C**: [0.5 points]
You may or may not have noticed that I moved `FacilityID` to our rownames in the datasets you loaded today. Why did I choose to do that? Why do I want to make sure unique identifiers, like `FacilityID`, are out of the dataframe before I either (a) run `nearZeroVar()` or (b) proceed with the rest of the analysis?

**Answer:** 
Moving FacilityID to rownames keeps the hospital labels accessible without exposing them to the modeling pipeline. FacilityID is a unique identifier — every hospital has a different value — which means it has maximum variance and would never get flagged by nearZeroVar(). That's the opposite of what we'd want: as a label, it carries no predictive signal, but every algorithm we run downstream (regression, PCA, k-means, random forest) would happily treat it as a feature and try to find patterns between hospital ID numbers and readmission rates. That would be meaningless at best and could distort results at worst.

The same logic applies to any unique identifier — names, addresses, phone numbers, etc. They're useful for tracking which row is which, but they aren't predictors. Storing them as rownames preserves the lookup capability while keeping the dataframe "clean" of non-informative columns before any feature-selection, dimension-reduction, or modeling step runs.

Lesson I learned from here: 

This is actually a generalizable rule worth internalizing: identifiers belong as labels (rownames, indexes), not as data. 


#### **Question 9D**: [1 point]
Do some research on, or lean into your previous knowledge, about the **curse of dimensionality** or the **big-P, little-N (n >> P) problem**. Using whatever you find or remember, can you justify why we likely do NOT have a **big-P, little-N (n >> P) problem** here? Make sure to explain. 

**Answer:**
The "big-P, little-N" problem (more accurately written P >> N) describes a situation where you have many more features than observations — for example, genomics datasets where there are tens of thousands of gene expression measurements but only a few hundred patients. When P >> N, the curse of dimensionality kicks in: the data becomes sparse in high-dimensional space, distance-based methods break down, OLS can't even be uniquely solved because the design matrix isn't invertible, and standard statistical theory (which assumes N >> P) no longer applies. You're forced into specialized methods like regularized regression or dimension reduction just to get a model to converge.

Our dataset is firmly in the opposite regime. After dropping hospitals with missing PredictedReadmissionRate, we'll have around 2,700 observations and only 30-60 features — a ratio of roughly 50 observations per feature. That's well within the range where standard statistical theory applies, OLS works without issue, and the feature space is well-populated. We're using Elastic Net for its feature-selection benefits, not because the math forces us to.

## 3.2 Assess missingness & devise an imputation plan

We are going to deal with imputation in a rather quick and elegant way that requires minimal effort from us using a model-based approach with `missForest`. You may recall from lecture that I said we are choosing a model-based approached _without uncertainty_ because our ultimate goal is to create a deployable model for our stakeholder.

#### **Question 10**: [1 point]
Write some code that will quickly tally the number of columns in the dataset that contain missing values. You will find `colSums()` and `is.na()` functions likely useful here. **What percent of the features have at least some level of missingness?**

```{r}
## Count missing values per column
missing_per_col <- colSums(is.na(dat2Analyze))

## How many columns have at least one missing value?
cols_with_missing <- sum(missing_per_col > 0)

## Total columns
total_cols <- ncol(dat2Analyze)

## Percentage
pct_missing <- (cols_with_missing / total_cols) * 100

## Print the results
cat("Columns with missing values:", cols_with_missing, "out of", total_cols, "\n")
cat("Percentage of features with missingness:", round(pct_missing, 1), "%\n")

```

**Which column(s) are not missing any data?**

**Answer:**
Only one column in dat2Analyze is fully complete: State. Every other feature (29 of 30, or 96.7%) has at least some level of missingness. This makes sense — every hospital in the CMS dataset has a known location, so State has no missing values by nature. All the quality metrics, satisfaction scores, and outcome measures depend on whether hospitals had enough cases or chose to report, so those have at least some hospitals missing values.

**How many complete observations do we have in the dataset?** You may find `nrow()` with `complete.cases()` or `drop_na()` useful!
```{r}
## Count rows with no missing values across any column
nrow(dat2Analyze[complete.cases(dat2Analyze), ])

## Alternative using drop_na() — same result
nrow(drop_na(dat2Analyze))

## For percentage context
n_complete <- nrow(drop_na(dat2Analyze))
n_total <- nrow(dat2Analyze)
cat("Complete observations:", n_complete, "out of", n_total, 
    "(", round(n_complete / n_total * 100, 1), "%)\n")

```


### 3.2.1 Use the `md.pairs()` function from the `mice` package to assess the extent of missing values before imputing. 

When we are missing a lot of data, as we are here, it can sometimes be useful to predetermine whether the _patterns of missing correlate with our target_. Why? We care if missing values correlate with the target because this may indicate that the missingness itself carries information. Thus, ignoring it could lead to biased models or low predictive power!

In fact, missingness can itself __become a predictor__. If a missing value is associated with the outcome, it may carry predictive signal. For example, in our data, hospitals with high rates of missingness in `Left before being seen` highly associated with pneumonia readmission might indicate a worrying pattern. Perhaps hospitals are negligently failing to report how many of their patients leave without being seen because of larger, systemic issues that ALSO happen to impact overall quality and patient care. Thus, missingness could itself become a predictor! 

To do, I will leverage the `md.pairs()` function from the `mice` package which calculates the pairwise correlations of missingness between variables and allows us to visualize those patterns using a clustered heatmap. When I do this, I will drop `State` from the matrix because it will create errors if I retain it. (**Why?**)

```{r}
## Drop State and analyze the pairwise missingness using the mice package
miceMatrix <- dat2Analyze %>% select(-State) %>% md.pairs()
```

What `md.mice()` does is create two pairwise correlation matrices, `mr` and `mm`, where:

* `mr` = __Missing vs. Responded__
  - Pairwise matrix showing how often the row has a missing value in one variable and an observed (i.e., non-missing or "responded") value in the other
* `mm` = __Missing vs. Missing__
  - Pairwise matrix showing how often the row has a missing value in one variable and missing value in the other variable

Let's take a look at the first 6 rows and 4 columns of the `mr` matrix:

```{r, echo=FALSE}
miceMatrix$mr[1:6, 1:4] %>% 
  kable(format = "html",
    caption = "Table 4. Missing vs. Reponded Matrix Slice") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F))
```

Any given cell in a matrix is always of the form $[i, j]$, where $i$ is the row and $j$ is the column - just as you are used to with dataframes! So, in these correlation matrices of our pairwise variables, cell $[1,1]$ is `PredictedReadmissionRate` vs. itself, so the count is 0 - as is every value on the diagonal, like with all correlation matrices. For cell $[1,2]$ `PredictedReadmissionRate` vs. , there are 42 observations where variable `PredictedReadmissionRate` ($i$) is missing but `MRSA Bacteremia` ($j$) was observed. Thus, we can **extract the number of cases where variable $i$ is missing and variable $j$ is observed using** `miceMatrixmr[i, j]`.

Now, let's take a look at the first 6 rows and 4 columns of the `mm` matrix now:

```{r, echo=FALSE}
miceMatrix$mm[1:6, 1:4] %>% 
  kable(format = "html",
    caption = "Table 5. Missing vs. Missing Matrix Slice") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F))
```

Similarly, `miceMatrix$mm[i, j]` gives us the number of cases where both variables $i$ and $j$ were co-missing.

#### **Question 11A**: [0.5 points]
Explain what you think the number in cell $[6,1]$ means. Out of all the observations in the dataset, does this feel like a high degree of co-missingness?

**Answer:**
The value in cell [6, 1] is 1,788, meaning that 1,788 hospitals are missing BOTH `Postoperative respiratory failure rate` AND `PredictedReadmissionRate`. That's a high degree of co-missingness from two perspectives: it represents 37% of the entire 4,816-hospital dataset, and 85.5% of the 2,090 hospitals that are missing the target (1788 / 2090).

This much overlap suggests the two variables aren't missing independently — they're likely missing for a shared underlying reason. The most plausible explanation is hospital size and case volume: CMS suppresses both risk-adjusted readmission rates and surgical complication rates for hospitals with too few cases to compute statistically reliable values. So both variables tend to be missing at the same (smaller, lower-volume) hospitals.


Lastly, we can calculate a **proportion of missingness vs. response** between any two variables $i$ and $j$ such that:

$Proportion_{i,j} = \frac{mr_{i,j}}{mr_{i,j} + mm_{i,j}}$

In non-mathy speak, it's the fraction of the times that variable $j$ is observed among rows where $i$ is missing. So, if you want to impute variable $i$ - let's say it's `MRSA Bacteremia` - then you will be interested in the variables typically **observed** (present in the dataset) when $i$ is missing. Those are the variables that will be able to help us predict, or fill in, $i$. 

 - If this ratio is close to 1, $j$ is nearly always available when $i$ is missing, so $j$ might be useful for imputing $i$.
 - If this ratio is close to 0, $j$ is usually missing when $i$ is missing, so it’s not helpful for imputing $i$. You can also use this to tell you about **patterns of missingness with regard to your target variable when you target is** $i$. 

```{r, fig.cap="Figure 5. Correlation between Missing and Response", echo = FALSE, fig.width = 9, fig.height = 9}
## Calculate proportion from pairwise pattern of missingness
p <- as.matrix(miceMatrix$mr/(miceMatrix$mr+miceMatrix$mm))

## Set a color palette
prettyPurples <- colorRampPalette(brewer.pal(8, "PuRd"))(8)

## Make a heatmap
heatmap(p, 
        col = prettyPurples, 
        cexRow = 0.5, 
        cexCol = 0.5, 
        margins = c(13, 13), 
        Colv = NA)
```

It may feel a little counterintuitive, but here __darker colors__ indicate a higher correlation between missingness and response (proportions nearer to 1). In other words, these are the variables that are going to be more informative during imputation for a given variable. 

But what if we wanted to know about patterns of missingness with regard to our target, `PredictedReadmissionRate`? Well, we could get that too!

```{r, fig.cap="Figure 6. Correlation between Missing and Missing for Target", echo = FALSE, fig.width = 7, fig.height = 7}
## Calculate proportion from pairwise pattern of missingness
p <- as.matrix(miceMatrix$mm/(miceMatrix$mr+miceMatrix$mm))

## Set a color palette
prettyPurples <- colorRampPalette(brewer.pal(10, "PuRd"))(30)

## Make a heatmap
p[1, ] %>% 
  data.frame(Proportion = .) %>% 
  mutate(Variable = rownames(p),
         Variable = fct_reorder(Variable, Proportion)) %>% 
  filter(Variable != "PredictedReadmissionRate") %>% 
  ggplot(aes(x = Variable , y = Proportion)) +
  geom_bar(aes(fill = Variable), stat = "identity", color = "navy") +
  theme_minimal() +
  labs(title = "Highest Correlations of Missingness \nwith Predicted Readmissions",
       x = "Variables",
       y = "Proportion of Co-missingness") +
  scale_fill_manual(values = (prettyPurples)) +
  coord_flip() +   
  theme_classic() +
  theme(legend.position = "none")
```


#### **Question 11B**: [0.5 points]

What columns seem to be most correlated in terms of missingness to our target, `PredictedReadmissionRate`? Does this give you any cause for concern? Feel free to speculate as to why you might see this.

**Answer:**
The most co-missing variables with PredictedReadmissionRate are MRSA Bacteremia (~98%), followed by Postoperative respiratory failure, Medicare spending per patient, Perioperative PE/DVT, and Composite patient safety (all ~84–85%). The shared pattern is hospital size: CMS suppresses risk-adjusted metrics for hospitals with too few cases, and the same small hospitals lacking enough pneumonia cases also lack the surgical and infection cases to compute those other measures.

The main concern this raises is more practical. When we drop hospitals missing the target, we'll systematically remove smaller hospitals from the analysis, meaning our final recommendations will apply mainly to mid-sized and large facilities, not the small or specialty hospitals that get filtered out.

## 3.3 Drop rows that are missing from the target variable. 

We will not be able to use those rows at all moving forward, sadly. That is going to reduce our sample size pretty sizably, from $N = 4,816$ to only $N = 2,726$ (or from $N = 4,775$ to only $N = 2,731$ .

```{r, echo = FALSE}
dat2Analyze <- dat2Analyze %>% 
  filter(!is.na(PredictedReadmissionRate)) 

paste0(c("Rows: ", "Columns: "), dim(dat2Analyze))
```


#### **Question 12**: [0.5 points]
Thinking back to your answer to Question 9D, do you feel any differently given these dimensions of the dataset?

**Answer:**
We have 2,726 hospitals and 29 predictors, That works out to roughly 94 observations per predictor. That's actually a better ratio than I estimated in Q9D and remains well clear of the P >> N regime. The newer concern isn't dimensionality but sample bias: the 2,090 hospitals we dropped were disproportionately smaller and lower-volume, so our remaining sample is skewed toward mid-sized and larger facilities. That doesn't affect the modeling itself, but it limits the generalizability of our final recommendations.

## 3.4 Do the splits! 

At this stage, we are almost ready to __impute__ - but not until we have done our data partitioning first! In fact, splitting our data into training and testing partitions first is really important to prevent __data leakage__.

#### **Question 13**: [1 point]
Explain to your stakeholder, in simpler language, (1) why we partition data into testing and training sets and (2) what data leakage is. You do not need to go into great detail, but it should explain enough that they understand the basic ideas.

**Answer:**
If we trained a model on all our hospitals and then evaluated it on those same hospitals, we'd be cheating — the model has already seen the answers. To get an honest measure of how well it performs on new data, we split our hospitals into a training set (the data the model learns from, usually 70-80%) and a testing set (the remaining hospitals, set aside as a "fresh" sample). We build the model on training data, ask it to predict outcomes for the testing data, and compare those predictions to the real values. That comparison tells us how the model will perform in the real world on hospitals it hasn't seen before.

Data leakage is when information from the testing set accidentally influences the training process. The model effectively gets a peek at the answers before the "test," which makes it look more accurate than it really is. A common way this happens in our project: if we filled in missing values using statistics calculated from the full dataset (train + test combined), the test set's information would bleed into the training data, and our evaluation would be biased. The simple rule is to treat the testing set as "future" data — never let it influence any step of preparing or training the model.


You may recall that I mentioned that the optimal ratio is has been argued to be $\sqrt{p} : 1$, where $p$ is the number of parameters (which may or may not equal your predictors depending on your planned analysis). Note that you do not have to split only a single time; you could make different splits for different analyses, if needed. However, we are going to use a single split here for convenience. 

Now, wouldn't it be nice to have a handy function that could calculate the optimal split ratio for us? So, let's write one! This is something you can use with ANY analysis going forward!

#### **Question 14A**: [1.5 points]
There are three parts to this question to get it to all come together correctly. 

1. Comment the function to explain what it does (I've defined the arguments for you)

2. Run the function (fill in the blank)

3. Fill in the missing piece in the data partitioning chunk below

```{r}
calcSplitRatio <- function(p = NA, df) {
  ## @p  = the number of parameters. by default, if none are provided, the number of columns (predictors) in the dataset are used
  ## @df = the dataframe that will be used for the analysis
  
  ## If p was not supplied, calculate it from the dataframe instead
  if(is.na(p)) {
    p <- ncol(df) -1  ## Subtract 1 to exclude the target/outcome column, leaving only predictors
  }
  
  ## Calculate the optimal number of rows to allocate to the TEST set using the sqrt(p):1 rule
  test_N <- (1/sqrt(p))*nrow(dat2Analyze)
  
  ## Convert that test row count into a proportion of the full dataset (rounded to 2 decimal places)
  test_prop <- round((1/sqrt(p))*nrow(dat2Analyze)/nrow(dat2Analyze), 2)
  
  ## The training proportion is simply whatever isn't going into the test set
  train_prop <- 1-test_prop
  
   ## Print summary of the recommended split ratio
  print(paste0("The ideal split ratio is ", train_prop, ":", test_prop, " training:testing"))
  
   ## Return the training proportion so it can be used downstream (e.g., as the p argument in createDataPartition())
  return(train_prop)
}

## Fill in the blanks to run:
train_prop <- calcSplitRatio(df = dat2Analyze)
```

**Hint**: One important note here is that if the number of parameters are not provided (which is true for the starter code I gave you to run it - it does NOT pass any parameters to that argument!), then by default our function is designed to take the number of columns of the dataframe and subtract one for the target variable. Neat, huh?

Now, uncomment this code and fill it out to run it. 
```{r}
ind <- createDataPartition(dat2Analyze$PredictedReadmissionRate,   ## Put the target here!
                            p = train_prop,           ## Use the object you just made!
                            list = FALSE)
 
train <- dat2Analyze[ind, ]
test <- dat2Analyze[-c(ind), ]

## Check the dimensions of your splits
dim(train)
dim(test)

## Confirm the split proportion
nrow(train) / nrow(dat2Analyze)  ## Should be about 0.82
```

#### **Question 14B**: [0.5 points]
What split ratio did you get? Is it close to the canonical 80-20 split? Why do you think that happened?

**Answer:** 
I got 0.81:0.19 training:testing, which is essentially the canonical 80-20 split with a small nudge toward training. That's exactly what the math predicts: the sqrt(p):1 rule with p = 29 predictors gives sqrt(29) ≈ 5.39, so 1/5.39 ≈ 0.19 for the test proportion. The classic 80-20 rule of thumb actually became popular because it works well for the typical 20-50 predictor range that most analyses sit in — our dataset fits squarely in that range, so the formula confirms it. If we had only a handful of predictors or several hundred, the optimal would shift away from 80-20 in meaningful ways.

__Did you get stuck somewhere?__ If so, load the `test` and `train` data in case you struggled with Question 18 so you can proceed:

```{r}
load(file = "FY2024_data_files/pneumoniaTrain.Rdata")
load(file = "FY2024_data_files/pneumoniaTest.Rdata")
```


## 3.5 Impute missing variables using `missForest`.

We are going to take advantage of the `missForest()` function from the `missForest` package. In a nutshell, what `missForest` does is it will fit either a regression- or classification-based random forest model and use the OOB ("out-of-bag") results to predict and fill in `NA`s. The advantage is that this process is non-linear, done in just a few lines of code, and can handle mixed data-types (although you need to make sure that all your numerical types are of the same numerical type, and that all characters or factors are of the same type).

#### 3.5.1 Running `missForest`

The way I have chosen to do this is to separate the data so that I can turn the encoded categories back into categories temporarily, then stitch them back together into a temporary dataframe so that I can run `missForest`. Notice that I temporarily drop the `State` and target features; I drop `State` to spare ourselves a little computational time and I drop `PredictedReadmissionRate` to prevent __data leakage__. **Be patient!!**. Since this is randomForest, it could take a minute or two to run. **NOTE:** _If your machine fails to run this, there are data you can load. Some computers may struggle to run this._

```{r, include = TRUE}

# Remove encoded categorical variables for imputation, plus the target and state (which have no missing)
data_sans_encoded = train %>% 
  select(-PredictedReadmissionRate,
         -State,
         -`Emergency department volume`, 
         -contains("ComparedToNational"))

data_cat = train %>% 
  select(`Emergency department volume`, 
         contains("ComparedToNational")) %>% 
  mutate_all(as.factor)

# Stitch them back together:
temp <- cbind(data_sans_encoded, data_cat)

# Impute missing values using missForest
imputedTrain <- missForest(temp, 
                          variablewise = TRUE,
                          verbose = FALSE)


## Count the columns in the temp dataframe
dim(temp)
```

**NOTE**: Did you catch that I not only _removed_ the target before imputation, but I dropped `State` too? This is because we know `State` is/will be frequency encoded (depending on our path), and it could create imputation issues as a result. This leaves me with **29 columns to perform imputation on**. 

#### 3.5.2 Load this if it will not run for you
```{r}
load(file = "FY2024_data_files/imputedTrain.Rdata")
```

#### 3.5.3 Quick summary of the results

**NOTE:** If yours didn't run, make sure to look at the table output in the knitted HTML I provided you. Also, change the `include=FALSE` in the chunk header if you were not able to get `missForest` to run.

```{r, echo = FALSE, include = TRUE}
# Show a quick summary of the results
data.frame(imputedTrain$OOBerror, names(temp)) %>% 
  rename(Variable = `names.temp.`,
         `OOB Error` = imputedTrain.OOBerror) %>% 
## Print the columns
  kable(digits = 2,
    format = "html",
    caption = "Table 6. missForest OOB Error Rates for the imputed variables, training dataset") %>%
    pack_rows("MSE", 1, 22) %>%
    pack_rows("PCF", 23, 28) %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
)
```

#### **Question 15**: [1 point]
What does `MSE` (Mean Squared Error) and `PCF` (Proportion of Falsely Classified) indicate here? You may find looking at the `missForest` [Vignette](https://stat.ethz.ch/education/semesters/ss2012/ams/paper/missForest_1.2.pdf) helpful. Which variable(s) had the highest OOB error?

**Answer:**
MSE (Mean Squared Error) measures imputation error for numeric variables — the average squared difference between the imputed value and the true value, calculated using the random forest's out-of-bag samples. Lower MSE means the imputed values are closer to the truth, but MSE is reported in the original variable's units squared, so it isn't comparable across variables on different scales.

PCF (Proportion of Falsely Classified) measures imputation error for categorical (factor) variables. It's the fraction of out-of-bag observations whose imputed category was wrong, bounded between 0 and 1 — 0 means every imputation matched the true category, and 1 means every imputation was wrong. Unlike MSE, PCF is directly comparable across variables.

In our table, the highest raw OOB error belongs to Payment for pneumonia patients (MSE ≈ 933,239), but that's misleading because Payment is measured in dollars in the thousands — squaring inflates the number dramatically. The square root gives an RMSE of about $960, which is high but not catastrophic given the variable's range. The more concerning result is Emergency department volume (PCF = 0.48), where 48% of imputed values were misclassified — well above the typical 0.30 threshold for trustworthy categorical imputation. ComparedToNational_Hospital return days also has a high PCF (0.36), but it's clinically tied to the outcome so we may want to retain it despite the warning. Without normalizing MSE values by each variable's standard deviation, raw OOB error values can be misleading on their own.


**NOTE 1**: If the OOB error (either MSE or PCF) from `missForest` is really high, it’s a red flag that the imputation for those variables is untrustworthy. This could be do to too much missingness, weak relationships among the variables making it hard to predict, or low quality data (e.g., noisy, sparse). In such cases, we want to **DROP** any variables with extremely high OOB error because they add noise. A general rule of thumb is that if the PCF > ~0.3 (30% of the imputed values are wrong) or if there is especially high MSE (inflation) relative to the variance of the variable, we should drop it. Let's set those aside so we can drop them from `test` and `train` both downstream! 

**NOTE 2**: Even though the PCF is a bit too high, I am choosing to retain `ComparedToNational_Hospital return days for pneumonia patients` because it is a primary predictor. This could be a poor choice, however! Every decision we make as data scientists can _come with pitfalls_.

```{r}

## See what's currently in the imputed training data
names(imputedTrain$ximp)
ncol(imputedTrain$ximp)


## These look okay, which is why we won't drop them. 
## Uncomment to see their variances.
var(train$`Venous Thromboembolism Prophylaxis`, na.rm = T)
var(train$`Healthcare workers given influenza vaccination`, na.rm = T)

cols2drop <- c("Payment for pneumonia patients", "Emergency department volume")
```

#### **Question 16**: [1 point]
Alternatively, we could use MICE (Multivariate Imputation with Chained Equations), which is a method that will model uncertainty along with the variables. Do a little research. How does MICE work? Include a citation to a source.

**Answer:**
MICE (Multivariate Imputation by Chained Equations) imputes missing data through sequential regression. Each incomplete variable is temporarily treated as the outcome, predicted from all other variables using an appropriate regression (linear for continuous, logistic for binary, polytomous for categorical, etc.), and MICE cycles through every variable iteratively until imputed values stabilize.

The crucial feature is the "multiple imputation" step: MICE generates m separate imputed datasets (typically 5–20), each with slightly different values due to stochastic draws from the regressions. Downstream analyses are run on each dataset and pooled using Rubin's Rules, which produces standard errors that combine within-imputation and between-imputation variance — properly capturing the uncertainty added by imputation. This is the main advantage over single-imputation methods like missForest, which give point estimates but don't propagate imputation uncertainty into your final inferences.

**Reference:** van Buuren, S. & Groothuis-Oudshoorn, K. (2011). "mice: Multivariate Imputation by Chained Equations in R." *Journal of Statistical Software*, 45(3), 1-67.

#### 3.5.4 Putting it alllllll back together...

We are almost done with imputation. The last thing we need to do is extract the imputed values from `imputed_data$ximp`, and add the `State` and target variable back. Then we get to wash, rinse, and repeat for the test set! Oh my!

##### 3.5.4.1 Put the original `State` and `target` variables back

```{r}
imputedTrain <- imputedTrain$ximp %>% 
  ## Put the original State and target variables back on
  mutate(State = train$State,
         PredictedReadmissionRate = train$PredictedReadmissionRate) 
```

##### 3.5.4.2 Now do the imputation for `test`

```{r}
# Remove encoded categorical variables for imputation, plus the target and state (which have no missing)
data_sans_encoded = test %>% 
  select(-PredictedReadmissionRate,
         -State,
         -contains("ComparedToNational"))

data_cat = test %>% 
  select(contains("ComparedToNational")) %>% 
  mutate_all(as.factor)

# Stitch them back together:
temp <- cbind(data_sans_encoded, data_cat)

# Impute missing values using missForest
imputedTest <- missForest(temp, 
                          variablewise = TRUE,
                          verbose = FALSE)

dim(imputedTest$ximp)             
sum(is.na(imputedTest$ximp))      
```

##### 3.5.4.3 Load this if it  will not run for you

```{r}
load(file = "FY2024_data_files/imputedTest.Rdata")
```

##### 3.5.4.4 Put the original `State` and `target` variables back on `test`

```{r}
imputedTest <- imputedTest$ximp %>% 
  ## Put the original State and target variables back on
  mutate(State = test$State,
         PredictedReadmissionRate = test$PredictedReadmissionRate) 
```

##### 3.5.4.5 Drop the features we determined we should not retain due to unstable imputation from BOTH `train` and `test`

```{r}
imputedTrain <- imputedTrain[, !(names(imputedTrain) %in% cols2drop), drop = FALSE]
imputedTest  <- imputedTest[, !(names(imputedTest) %in% cols2drop), drop = FALSE]

dim(imputedTrain)
dim(imputedTest)
```


#### **Question 17**: [1 point]
Quickly explore how well the imputation did by choosing at least one numeric variable and one of the categorical variables. Did the distributions or frequencies change drastically? You may find the `summary()` and `table()` functions to be fastest here. 

```{r}
## ============================================================
## NUMERIC VARIABLE COMPARISON
## ============================================================
## Pick a numeric variable that had some missingness
## Healthcare workers given influenza vaccination is a good choice — 
## it had moderate MSE (285) and is on an interpretable 0-100 scale

cat("=== Numeric: Healthcare workers given influenza vaccination ===\n")
cat("\nBefore imputation (train, with NAs):\n")
summary(train$`Healthcare workers given influenza vaccination`)

cat("\nAfter imputation (imputedTrain, no NAs):\n")
summary(imputedTrain$`Healthcare workers given influenza vaccination`)

## ============================================================
## CATEGORICAL VARIABLE COMPARISON
## ============================================================
## Pick a categorical variable — let's use ComparedToNational_MRSA Bacteremia
## since MRSA had lots of missingness and a "Worse" label could be meaningful

cat("\n\n=== Categorical: ComparedToNational_MRSA Bacteremia ===\n")
cat("\nBefore imputation (train, with NAs):\n")
table(train$`ComparedToNational_MRSA Bacteremia`, useNA = "ifany")

cat("\nAfter imputation (imputedTrain, no NAs):\n")
table(imputedTrain$`ComparedToNational_MRSA Bacteremia`, useNA = "ifany")

## Quick visual comparison
par(mfrow = c(1, 2))

hist(train$`Healthcare workers given influenza vaccination`,
     main = "Before imputation", xlab = "Score", col = "lightblue",
     breaks = 30)

hist(imputedTrain$`Healthcare workers given influenza vaccination`,
     main = "After imputation", xlab = "Score", col = "lightpink",
     breaks = 30)

```

**Answer:**

The imputation performed very differently for the two variables I checked.

For `Healthcare workers given influenza vaccination` (numeric), the distribution barely changed at all. The mean shifted by only 0.02 percentage points (78.52 → 78.5), and every quartile remained nearly identical. The 21 missing values were filled with values that almost perfectly matched the existing distribution — exactly what we'd hope for. This makes sense given that only about 1% of values in this column were missing, so missForest had plenty of observed data to learn from.

For `ComparedToNational_MRSA Bacteremia` (categorical, encoded as -1/0/1), the result is more interesting. All 836 missing values were imputed as the modal category (0, "No Different than National Benchmark"). None went to -1 (Worse) or 1 (Better). The "0" category expanded from 1,193 to 2,029 observations while the other two categories stayed exactly the same. This is consistent with how random forest classifiers behave when uncertain on an imbalanced variable: they default to the most common category. In our case, since "No Different" was already ~87% of observed values, missForest essentially decided the safest guess for any missing value was "No Different."

Whether this is good or bad depends on interpretation. Statistically, most hospitals genuinely are "No Different than National Benchmark" — that's how significance testing works, since most hospitals can't be distinguished from average — so on average the imputation is probably making reasonable predictions. But it does mean the categorical variation we observed in the original data has been compressed: any genuinely Better or Worse hospitals that happened to be in the missing group have been lost. We'll want to be cautious about how heavily our downstream model leans on this variable.

## 3.6 Another Fork in the Road
It's time, yet again, to make a choice. If you opted to load the __fully encoded__ dataset and everything has been done on that, you will run the first chunk. But if you decided that you wanted to __wait to encode__ the TIME HAS FINALLY COME!

### 3.6.1 The Left Fork - No Encoding, Just Converting Data Back After `missForest`

Ultimately, we need to turn the factor features back into numbers for our next analyses but we had to turn them into factors for `missForest`. So, now we need to turn them back. Notice a bit of annoying trickery here. Because we had converted them to a factor type, in order to back-convert them correctly to our original ordinal encoding, we had to first convert to character and then to numeric. So, our conversion goes __factor__ $\rightarrow$ __character__ $\rightarrow$ __numeric__ in order to get exactly what we need! Again, we must remember to perform on BOTH `imputedTrain` and `imputedTest`. 

```{r}
## Uncomment if this is your choice!
readyTrain <- imputedTrain %>%
   mutate_if(is.factor, as.character) %>% 
   mutate_if(is.character, as.numeric)
# 
 readyTest <- imputedTest %>%
   mutate_if(is.factor, as.character) %>% 
   mutate_if(is.character, as.numeric)
 
## Verify the conversion worked
dim(readyTrain)
dim(readyTest)

```


### 3.6.2 The Right Fork - Frequency Encoding

The time has FINALLY come to do frequency encoding, if that is the fork in the road you took all the way back in [Section 1.3](#section1.3). Once again, we will be using a helper function that we load with the `source()` function like we did [here](#helperCode). It will do the frequency encoding for us on the `State` variable, if it needs to be done. If you started from the fully encoded dataset, there is a conditional statement that will skip this for you. 

```{r, results='hide'}
## This calls the helper function in the associated .R file
source(file = "doFrequencyEncoding.R")

## If State is not already a number
if(!is.numeric(imputedTrain$State)) {
    ## This sets a list of COLUMN NAMES I want to frequency encode
    cols2encode <- c("State")
  
    ## This runs the function, which takes up to 4 arguments
    ## Open the source code to see what each argument does!
    ## It returns a LIST of frequency encoded dataframes.
    freqEncoded <- doFrequencyEncoding(train = imputedTrain,
                                       test = imputedTest,
                                       cols2encode = cols2encode, 
                                       quiet = FALSE)
    
    ## Lastly, extract the newly encoded, imputed, dataframes!
    readyTrain <- freqEncoded$train
    readyTest <- freqEncoded$test
}
## Verify the conversion worked
dim(readyTrain)
dim(readyTest)

```
```{r}
## DEFENSIVE FIX: Make sure State is properly numeric, not table-class
## The helper function sometimes returns it in an unexpected format

## Build a clean named numeric encoder from imputedTrain
state_counts  <- table(imputedTrain$State)
state_encoder <- as.numeric(state_counts)
names(state_encoder) <- names(state_counts)

## Apply cleanly to both ready datasets
readyTrain$State <- as.numeric(state_encoder[as.character(imputedTrain$State)])
readyTest$State  <- as.numeric(state_encoder[as.character(imputedTest$State)])

## Handle any leftover NA values (from missing State entries)
if(sum(is.na(readyTrain$State)) > 0) {
  median_freq <- median(readyTrain$State, na.rm = TRUE)
  readyTrain$State[is.na(readyTrain$State)] <- median_freq
  readyTest$State[is.na(readyTest$State)]   <- median_freq
}

## Verify everything is clean
cat("All numeric in train?:", all(sapply(readyTrain, is.numeric)), "\n")
cat("All numeric in test?:", all(sapply(readyTest, is.numeric)), "\n")
cat("Missing in train:", sum(is.na(readyTrain)), "\n")
cat("Missing in test:", sum(is.na(readyTest)), "\n")
cat("Unique State values:", length(unique(readyTrain$State)), "\n")
```
```{r}
## Convert any remaining non-numeric columns to numeric
readyTrain <- readyTrain %>%
  mutate(across(where(~ !is.numeric(.x)), ~ as.numeric(as.character(.x))))

readyTest <- readyTest %>%
  mutate(across(where(~ !is.numeric(.x)), ~ as.numeric(as.character(.x))))

## Verify
all(sapply(readyTrain, is.numeric))
all(sapply(readyTest, is.numeric))
sum(is.na(readyTrain))
sum(is.na(readyTest))
```
#### **Question 18**: [1 point]
Because we are now frequency encoding AFTER data partitioning (we did ordinal encoding BEFORE the split for convenience), we absolutely __must__ apply the same encoding map to the `test` set that we did to the `train` set. Explain why, and what should we do if some classes of the variable end up in the `train` that is not in the `test` or vice versa? (**HINT**: Look at the source code!)

**Answer:**
The same encoding map must apply to both train and test for two reasons. First, the test set has to be treated as "future, unseen" data — if we computed a separate encoding from the test set, test information would influence our preprocessing pipeline, which is a form of data leakage. The model wouldn't be evaluated on truly unseen data. Second, consistency matters: when the model encounters California in the test set, it needs to see the same numeric value (e.g., 197) that it saw during training. If we used a separately-computed encoder, "CA" might map to a different number in test than in train, and the model would treat it as a different feature, breaking the comparability between the two datasets.
For mismatched classes, there are two cases:

**Class in test but not in train.** This is the harder case. The training encoder doesn't know that category, so the lookup returns NA. Common solutions are: (1) impute the missing encoding with the median or mean frequency from the training encoder (treating the unseen category as "average"), (2) fill with a small constant or zero to signal "rare/unknown," or (3) preemptively bucket rare categories into an "Other" group during training so anything unseen maps there. Looking at the source code, doFrequencyEncoding() handles this by [filling unseen test categories with NA, then likely imputing with the median frequency — this is the typical implementation pattern].

**Class in train but not in test.** This is actually not a problem. The training encoder still has a value for that state, but no test rows will trigger that lookup. The model learned a coefficient for it that simply doesn't get exercised at test time. No special handling needed.

## 3.7  Transformation & Scaling

Ultimately, we know we will need to scale & center our data, as that is required for the analyses I have set out in our Analysis Plan. But before we get into that, it would be nice to know what __realm of analyses__ in which we fall and, more importantly, do we need to consider *transformation* of our target in addition to the scaling & centering we plan to do on the entire dataset. 

**In other words: is our target variable approximately normally distributed?**

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 7. Histogram of Raw Predicted Readmission Rate"}
ggplot(readyTrain, aes(x = PredictedReadmissionRate)) +
  geom_histogram(alpha = 0.6, 
                 fill = "orchid",
                 color = "orchid") +
  theme_minimal() +
  labs(title = "Predicted Readmission Ratio, N = 2,726 hospitals",
       y = "Frequency",
       x = "Predicted Readmission Ratio")
```

#### **Question 19A**: [1 point]
Why is a histogram, or even a Q-Q plot, an insufficient way to assess whether a distribution is approximately normal?

**Answer:**
Histograms and Q-Q plots are useful starting points but are fundamentally subjective — what looks "approximately normal" depends on the viewer, the bin width, and the sample size. A histogram's appearance changes drastically based on how many bins you choose, and Q-Q plots can show small deviations from the reference line that are hard to interpret as meaningful versus random noise. Neither approach quantifies the deviation, so two analysts could reach different conclusions about whether transformation is needed.

The objective alternative is a formal hypothesis test like the Shapiro-Wilk test (best for n < 5000) or the Anderson-Darling test. These produce a p-value testing the null hypothesis that the data come from a normal distribution. Best practice is to combine a formal test with the visual inspection — the test gives the quantitative answer, and the plot gives the qualitative context.


Now Apply a [Shapiro-Wilk test](https://biostats.w.uib.no/test-for-normality-shapiro-wilks-test/) for normality using the `shapiro.test()` function. What does it indicate about your distribution? (**Note**: Shapiro tests are only reliable for $N < 5000$, but it is fine to perform here.)

```{r}
## Shapiro-Wilk test on PredictedReadmissionRate
shapiro.test(readyTrain$PredictedReadmissionRate)
```
**Answer:**
The Shapiro-Wilk test on PredictedReadmissionRate returned W = 0.985 and p-value = 3.085 × 10⁻¹⁴, well below the 0.05 threshold. This means we reject the null hypothesis of normality — statistically, the distribution is not normal.

However, the W statistic of 0.985 is very close to 1, which tells us the deviation from normality is mild in absolute terms. The extremely small p-value isn't evidence of severe non-normality — it's evidence that with N = 2,210 hospitals, the test has enormous statistical power and can detect even tiny deviations. This matches what we saw in the histogram: the target is approximately bell-shaped with a slight right skew, not severely non-normal.

above visual inspection showed: the target is mildly right-skewed, enough that a Box-Cox transformation will improve the fit of our OLS and Elastic Net models by making the residuals more normal. The Shapiro-Wilk result gives us a quantitative justification for applying the transformation, but it's the *shape* of the deviation (right skew), not its statistical significance, that drives the decision to transform.

### 3.7.1 Box-Cox Transformation

The Box-Cox transformation is a _family_ of power transformations used to stabilize variance and make data more normally distributed (Box & Cox, 1964). Recall from our brief discussion in lecture that the Box-Cox transformation involves the parameter $\lambda$, which when applied to $y$ yields the transformation. It is defined as 

$y^\lambda = \frac{y^\lambda - 1}{\lambda}$ for $\lambda \ne 0$. 

When $\lambda = 0$, it is the same as the $\ln(y)$. Our lecture slides have more reference values for how to interpret $\lambda$. But if $\lambda \ne 0$, then this means that the transformation is only applicable to positive values. The end result is that it helps improve the performance of statistical models that assume normality.

#### 3.7.1.1 There are multiple ways to apply the Box-Cox transform in R, but we are going to take advantage of the one in `caret` to make our lives easier, `BoxCoxTrans()`.

```{r}
bc_data <- BoxCoxTrans(readyTrain$PredictedReadmissionRate)
print(paste0("Box-Cox lambda =  ", bc_data$lambda))
```

We can see that the estimated $\lambda = -0.3$, so __approximately a square-root transform__ (see lecture slides for reference). I found that $\lambda = -0.5$ in FY 2025, for those using that year's data.

#### 3.7.1.2 Now, apply the Box-Cox transformation to the `train` data:

```{r, echo = TRUE}
readyTrain$bc_PredictedReadmissionRate <- predict(bc_data, readyTrain$PredictedReadmissionRate)
```

#### 3.7.1.3  Assess how it did:

```{r, echo = FALSE, message=FALSE}
ggplot(readyTrain, aes(x = bc_PredictedReadmissionRate)) +
  geom_histogram(alpha = 0.6, 
                 fill = "hotpink",
                 color = "hotpink") +
  theme_minimal() +
  labs(title = "Box-Cox Transformed Predicted Readmission Ratio, N = 2,726 hospitals",
       y = "Frequency",
       x = "Predicted Readmission Ratio")


qqnorm(readyTrain$bc_PredictedReadmissionRate)
qqline(readyTrain$bc_PredictedReadmissionRate, 
       col = "hotpink", lty = 2, lwd = 2)
```

##### Figures 8a-b. Predicted Readmission Rate After Box-Cox Transformation

Can you see what it did to the distribution? __look closely__, if not. It's subtle but discernible!!

#### **Question 19B**: [1.5 points]
Now repeat all of the steps to perform a Box-Cox transform on your own to the `imputedTest` data. Why must we (1) also transform the test data and (2) __use the same $\lambda$ as the training data__?

**(1) Why we must transform the test data:**

The model is being trained on the Box-Cox transformed target, so its predictions will be in transformed units. When we evaluate model performance (RMSE, MAE, R²), we need the test set's actual values to also be in transformed units — otherwise we'd be comparing predictions in one scale to truth in another, which would make all metrics meaningless. Transformation keeps train and test in the same statistical space.

**(2) Why we must use the same λ as training:**

The λ is a parameter learned from training data — it's part of the preprocessing pipeline, just like the model's coefficients. If we fit a separate λ on the test set, we'd be using test information to define part of our analysis, which is data leakage. The test set must mimic truly unseen data, and computing anything from it (even a transformation parameter) violates that assumption. The two datasets would also end up on slightly different scales, making any performance metric incomparable. Using `predict(bc_data, ...)` correctly applies the train-fitted λ = -0.3 to the test set without recomputing.


```{r}
# Your code here.
## Apply the same Box-Cox transformation (fitted on train) to the test set
## Note: we use the SAME bc_data object — this carries the train-fitted lambda = -0.3
readyTest$bc_PredictedReadmissionRate <- predict(bc_data, readyTest$PredictedReadmissionRate)

## Verify the transformation worked
summary(readyTest$bc_PredictedReadmissionRate)
sum(is.na(readyTest$bc_PredictedReadmissionRate))  

## Visualize to confirm it looks similar to the training transformation
ggplot(readyTest, aes(x = bc_PredictedReadmissionRate)) +
  geom_histogram(alpha = 0.6, 
                 fill = "darkturquoise",
                 color = "darkturquoise") +
  theme_minimal() +
  labs(title = "Box-Cox Transformed Predicted Readmission Ratio (Test Set), N = 516 hospitals",
       y = "Frequency",
       x = "Predicted Readmission Ratio (Box-Cox transformed)")

qqnorm(readyTest$bc_PredictedReadmissionRate,
       main = "Normal Q-Q Plot (Test Set)")
qqline(readyTest$bc_PredictedReadmissionRate, 
       col = "darkturquoise", lty = 2, lwd = 2)
# Make sure to check that your transform looks successful!
```

**NOTE** At this point, you have both an non-transformed (`PredictedReadmissionRate`) and Box-Cox transformed (`bc_PredictedReadmissionRate`) version of the target in your data.

### 3.7.2 Centering & Scaling

Now that we have applied a Box-Cox transform to the target variable, we can center and scale the remainder of the variables. We had to perform the Box-Cox _prior_ to centering because the data has to be positive for a Box-Cox transformation (unless we switch to a [Yeo-Johnson transform (Yeo & Johnson, 2000)](https://academic.oup.com/biomet/article-abstract/87/4/954/232908)). Centering will likely make a lot of the data negative, plus we know we have negatives in our dataset purposefully.

It's time to drop the _original_ target, `PredictedReadmissionRate`, and center and scale everything INCLUDING the target! Here it is for the training set:

```{r}
readyTrain <- readyTrain %>% 
  ## Drop the original target variable
  select(-PredictedReadmissionRate) %>% 
  ## Make anything that might still be a factor from encoding is back to number
  mutate_if(is.factor, as.numeric) %>%
  ## Center and scale everything!
  scale(center = TRUE, scale = TRUE) %>% 
  data.frame()
```

#### **Question 19C**: [0.5 points]
Do the same for your test data. Make sure to use `readyTest` and overwrite it. 

```{r}
readyTest <- readyTest %>% 
  ## Drop the original target variable (if it still exists)
  select(-any_of("PredictedReadmissionRate")) %>% 
  ## Make sure anything that might still be a factor from encoding is back to number
  mutate_if(is.factor, as.numeric) %>%
  ## Center and scale everything!
  scale(center = TRUE, scale = TRUE) %>% 
  data.frame()

```

**Hint**: Make sure to turn your centered, scaled __matrix__ back into a dataframe! By default, it creates a matrix.

### 3.7.3 To save time, you are welcome to start from here after you've worked through the above steps.
```{r}
load(file = "FY2024_data_files/readyTrain.Rdata")

load(file = "FY2024_data_files/readyTest.Rdata")
```

# 4 Assessing Multicollinearity & Feature Redundancy {#multicollinearity}

As we finally, finally wrap-up pre-processing there is one final assessment we need to perform: **multicollinearity**. As I am certain you recall, multicollinearity is when we have high correlation between predictors (so much so, in fact, that it comes at the cost of any correlation with the target!). Recall from lecture that multicollinearity is when one or more of our predictors are moderately to highly correlated with each other. This can __inflate our coefficients__ and thereby cause spurious associations through __false positives__ (i.e., small p-values when we don't actually have them!).

Here we will implement two common ways we assess multicollinearity: pairwise correlations and Variance Inflation Factors (VIFs). 

## 4.1 Pairwise Correlation 

Pairwise correlation is usually done as a Pearson or Spearman correlation. We will calculate a correlation matrix and look at the correlation of the variables within the training dataset to assess possible areas of multicollinearity. We can then visualize this matrix with a heatmap, similar to how we did with missingness, except now blue indicates a __positive__ correlation, red indicates a __negative correlation__, and yellow indicates __no (or zero)__ correlation. Darker colors indicate stronger correlations. 

```{r, echo = FALSE, fig.width = 9, fig.height = 9, fig.cap="Figure 9. Correlation Heatmap of All Variables"}
## Calculate a correlation matrix
corrMat <- cor(readyTrain)

## Make a heatmap
heatmap(corrMat, 
        col = rev(colorRampPalette(brewer.pal(8, "RdYlBu"))(8)), 
        cexRow = 0.6, 
        cexCol = 0.6, 
        margins = c(10, 10))
```

#### **Question 20A**: [1 point]
Where do you see the highest potential for multicollinearity? Why do you say that?

**Answer:**
Highest multicollinearity, The 11 HCAHPS patient-satisfaction variables form a tightly correlated cluster in the upper-right of the heatmap. They all measure facets of patient satisfaction during the hospital stay, so they move together — Doctor communication, Nurse communication, Overall hospital rating, Recommend hospital, etc. are essentially measuring the same underlying construct. I'd expect several of these to have VIFs above 10.

A secondary multicollinearity concern is the paired score/comparison variables: each ComparedToNational_X is derived from its numeric counterpart, making them near-duplicates. MRSA Bacteremia ↔ ComparedToNational_MRSA Bacteremia is the clearest example.

Which variable(s) seem to be most correlated with our target, `bc_PredictedReadmissionRate`?

**Answer:**
Three variables stand out — Medicare spending per patient (strongest, consistent with the Q4 scatterplot findings), PaymentCategory for pneumonia patients, and State. These cluster with bc_PredictedReadmissionRate in the bottom-left corner. The HCAHPS satisfaction variables show only modest individual correlations with the target despite their internal interdependence.


## 4.2 Fitting a model to estimate VIF

Another way we can assess multicollinearity is by fitting a multiple linear regression model and estimating the __variance inflation factors__, or VIF, which estimates how much the variance of an estimated regression coefficient increases if your predictors are correlated. Specifically, variance inflation can identify __multicollinearity__ when the VIF is greater than ~4 and especially when it is above 10.

## 4.2.1 Fit a linear regression with the training data, using the Box-Cox transformed `Predicted Readmission Rate` as the target. 
```{r}
mod <- lm(bc_PredictedReadmissionRate ~ ., data = readyTrain)
```

## 4.2.1 Next, let's calculate the VIFs using the `vif()` function that is part of the `car` package. 

Note that I am not printing all of the VIFs, just the top 15 after sorting in descending order.

```{r} 
## Calculate the VIFs and put into a dataframe to print the table
vifResults <- vif(mod) %>% data.frame
## Change the column names of the dataframe
colnames(vifResults) <- c("VIF")
```

```{r, echo = FALSE, warning=FALSE,message=FALSE}
## Sort by VIF, print the top 15 worst offenders
vifResults %>% 
  arrange(desc(VIF)) %>% 
  top_n(15) %>% 
  ## Pass through kable() to make it pretty
  kable(digits = 2,
    format = "html",
    caption = "Table 7. Top 15 Variance Inflation Factors after multiple linear regression") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
)
```

#### **Question 20B**: [1 point]
Does multicollinearity seem to be an issue in this dataset and, if so, among which variables? Why do you come to this conclusion?

**HINT 1**: What was used to calculate `Overall Hospital Rating`? Does this help us better understand why the VIF is so large?

**HINT 2**: Is, for example,`Payment Category for pneumonia patients` created from `Payment for pneumonia patients`? [These are Medicare payments]? Which one do you think we should keep amd why?

**Answer:**
Multicollinearity is a real issue, concentrated entirely among the HCAHPS patient satisfaction variables. Overall hospital rating (VIF = 22.59) and Recommend hospital (VIF = 14.75) are both above the severe threshold of 10, and 5 more HCAHPS variables fall in the 4–10 moderate range.

Overall Hospital Rating is a composite measure— patients' "overall" rating is necessarily shaped by their experiences with nurse communication, doctor communication, cleanliness, and so on. Including the composite alongside its components is statistically redundant, hence the inflated VIF. Recommend Hospital has the same problem.

Which variable(s) seem to be most highly correlated with our target? Is it a positive or negative correlation? Does it make sense?

**Answer:**
Most correlated with the target: Medicare spending per patient shows the strongest positive correlation, followed by PaymentCategory and State. The direction is positive — higher Medicare spending means sicker patients, which means more readmissions. This is a *patient-mix* story rather than a hospital-quality story, which makes clinical sense and is consistent with what we observed in the Q4 scatterplots.

### 4.3 Dropping the most collinear variables

Even though we plan to proceed with an elastic net (which is much more robust to multicollinearity), some of these VIFs are too high for comfort. We also want to pause and ask - **were any of the variables used to make other variables?** (The answer is "yes"!). Those are excellent candidates to scrub too; we want to use our domain knowledge and what our stakeholders are asking for to help us decide which ones to keep.

Because the goal is a deployable predictive model, we likely want something more granular than most of the `ComparedToNational` features can give us, with the __exception of our key predictor__, `ComparedToNational_Hospital return days for pneumonia patients`. I will remind you of what we explored in Section 2 with this variable; not only that, we can see how **how highly correlated it is with our target!**.

### 4.3.1 Drop variables with VIF over 5 from both the training and testing data

```{r}
## Drop those with VIF over 5
vifOver5 <- vifResults %>% filter(VIF >=5) %>% rownames()
print(vifOver5)

## Now drop them
readyTrain <- readyTrain %>% select(-any_of(vifOver5))
readyTest <- readyTest %>% select(-any_of(vifOver5))
```

### 4.3.2 Now drop all of the `ComparedToNational` and `PaymentCategory` features **except** `ComparedToNational_Hospital return days for pneumonia patients`
```{r}
readyTrain <- readyTrain %>% 
  ## Rename the column we want to keep
  rename(Hospital.return.days.for.pneumonia.patients = ComparedToNational_Hospital.return.days.for.pneumonia.patients) %>% 
  ## Drop all the other columns 
  select(-contains("Compared"), -contains("PaymentCategory"), -SurveyResponseRate)

readyTest <- readyTest %>% 
  ## Rename the column we want to keep
  rename(Hospital.return.days.for.pneumonia.patients = ComparedToNational_Hospital.return.days.for.pneumonia.patients) %>% 
  ## Drop all the other columns 
  select(-contains("Compared"))
```

Need fix readyTest
```{r}
## Defensive cleanup to align readyTest with readyTrain
## Original chunk above already cleaned readyTrain;
## this chunk catches any residual mismatches in readyTest

readyTest <- readyTest %>%
  select(-any_of("PaymentCategory.for.pneumonia.patients"),
         -any_of("SurveyResponseRate"))

## Verify train and test now match
dim(readyTrain)
dim(readyTest)
all(names(readyTrain) == names(readyTest))
setdiff(names(readyTrain), names(readyTest))
setdiff(names(readyTest), names(readyTrain))
```
# 5 Unsupervised Learning Methods: Segmentation Analysis

Predictive modeling is our ultimate goal, but let's pretend our stakeholders have expressed a desire to better understand how the hospitals relate to one another and whether there are any general patterns we should be interested in for enhanced interpretation. After all, a prediction is one thing, but understanding the _why_ is still usually important!

Thus, we have opted to undertake a __Segmentation Analysis__ to see if there is any underlying segments, or clusters, of hospitals based on the data we have. We will use the unsupervised methods of Principal Component Analysis (PCA) and $k$-means clustering do this; it will help us to determine if there are broader classifications of our hospitals that we need to report back to the advocacy group. 

## 5.1 Principal Component Analysis

First, we apply a PCA on the data and display the principal components. Note that we have the same number of **principal components** as we do variables in the data set! The _cumulative proportion_ of variance adds to 100% by our last principal component; but we're really more interested in the portion of variance explained by EACH of the principal components. This can help us figure out just how many PCs are important for explaining the most variance. For example, if we were interested in feature reduction, we could use the top PCs that explain, say, some portion of the variance that we wish to retain.

#### **Question 21**: [2 points]

1. At what principal component do we surpass 80% of explained variance? (**HINT**: look at the cumulative variance row rather than tha proportion of variance row).

**Answer:**
The first principal component to surpass 80% of cumulative explained variance is PC11 (cumulative = 83.93%). PC10 falls just short at 79.78%, so we need 11 of the 18 total principal components to capture 80% of the variation in our predictor data.

The variance in our dataset is spread relatively thinly across many components. PC1 captures only 22.3% of the variance, PC2 adds another 10.2%, and no single component dominates. This tells us the hospital data has many independent dimensions of variation — there isn't a single "underlying performance factor" that explains most hospital differences. Instead, hospitals vary along multiple roughly-equal axes (likely things like patient mix, quality processes, communication, infection control, etc.). This will inform how we approach clustering: with the variance so distributed, we should expect clusters that capture broad patterns rather than crisp groupings around one or two dominant features.

```{r}
pca <- prcomp(readyTrain)
summary(pca)
```

2. You may also notice that this is always ordered from most proportion of variance on the left (thus is PC1 defined) to our final principal component on the right. What percent of variance is explained by just PC1? What percent by the final PC? Is the attributable portion of the variance in PC1 high, medium, or low based on some quick research or prior knowledge?

**Answer:**
PC1 explains 22.34% of total variance, while the final PC (PC18) explains only 1.06%. The 22% by PC1 is medium-to-low — for context, PCAs in fields where a single latent factor dominates (like demographic studies) often see PC1 around 50–70%, while datasets with no dominant axis see PC1 below 20%. Our 22% sits in the middle range, telling us hospital data has multiple roughly-equal dimensions of variation rather than a single underlying performance factor. This is consistent with the broader finding that we need 11 PCs to reach 80% cumulative variance.


3. We use scree plots to graphically determine how many principal components to retain and how many principal components best explain the variance of the data. The $x$-axis displays the principal components (PC1, PC2, etc.), while the $y$-axis shows the eigenvalues (or proportion of variance explained) for each component. The **elbow method** we discussed in class is the point where the slope starts to flatten or the point at which the percentage of variance explained seems to level off. If we were doing this for feature reduction, we might think of this as approximately how many principal components would be needed to explain the total variation in the training data.  However, it is a **heuristic** method, so it can sometimes be a little uncertain from the plot alone what the optimal number of PCs is. Let's take a look at the top-15 PCs on a scree plot now:

```{r, echo= FALSE, fig.cap="Figure 10. Scree Plot of First 15 Principal Components"}
## Show the scree plot
fviz_screeplot(pca, 
         addlabels = TRUE, 
         ylim = c(0, 30),
         ncp = 15,
         barfill = prettyPurples[13],
         barcolor = prettyPurples[15],
         main = "Scree Plot: First 15 Principal Components")
```

How many PCs appear optimal to you based on the scree plot?

**Answer:**
The scree plot shows two possible elbows: a sharp one between PC1 and PC2 (suggesting 2 PCs, but only capturing ~32.5% of variance), and a subtler one around PC5-PC6 where the curve clearly flattens (suggesting 5-6 PCs, capturing ~55-61% of variance). I'd lean toward 5-6 PCs as the more useful cut — 2 PCs throws away too much information, while 5-6 captures meaningful structure without overfitting to noise. That said, neither elbow is dramatic, which is consistent with our earlier finding that hospital data is multidimensional and needs 11 PCs to reach 80% cumulative variance.

4. People often mis-write scree (I've even accidentally done it) as SCREE, as if it's an acronym like AUROC or ROC or even PCA. But it's not. Why is it called a scree plot? (**Hint**: It may not be what you expect!)

**Answer:**
"Scree" is a geology term, not an acronym — it refers to the loose rocks that pile up at the base of a steep cliff. Statistician Raymond Cattell named these plots in 1966 because their shape mimics that geology: a sharp drop on the left (the cliff face = dominant PCs) flattening into a gentle tail on the right (the rubble = less informative PCs). The "elbow" is where the cliff ends and the scree begins.

### 5.1.2 Variable loadings and contributions

Next, let's explore how much the variables are contributing (this is based on the weight of their rotations). 
In PCA, we are interested in the __loadings__, i.e., the linear combination weights (coefficients) whereby unit-scaled components define or "load" a variable. Loadings help us interpret principal components.

You may have noticed we used the `prcomp()` function above; `prcomp()` will return the loadings in the variable `$rotation`, which contains a matrix of variable  loadings. These loadings have been determined from the eigenvectors, which without getting into what that is (because we'd have to take a departure into linear algebra), suffice it to say that they are how we are measuring the direction and magnitude during each subsequent rotation of the principal components as we measure the variance explained.

Additionally, we are going to color this by our grouping variable and key predictor in the dataset, `Hospital.return.days.for.pneumonia.patients`. Recall that when we explored this earlier this variable is the performance of each hospital in terms of its pneumonia-related return-to-hospital days relative to the national average. In other words, hospitals that were "Worse than average" had readmitted pneumonia patients with longer-than-average stays. 

```{r, echo = FALSE, fig.cap="Figure 11. PCA Biplot Showing the Top Feature that Contributes to Explained Variance"}
## Create a factor variable that contains the information about the compared to national average - return to hospital days variable
returnHospitalComparison <- factor(readyTrain$Hospital.return.days.for.pneumonia.patients,
                                   labels = c("Worse", "Same", "Better"))

## Now show the biplot
fviz_pca_biplot(pca, 
             palette = c("maroon", "cadetblue", "gold"),
             label = "var",
             geom = "point",
             geom.var = "text",
             addEllipses = TRUE,
             ellipse.alpha = 0.5,              ## Change alpha 0-1
             col = "black",
             col.var = "black",
             habillage = returnHospitalComparison,
             select.var = list(contrib = 10),   ## Try changing the number
             repel = T)                        ## Try switching to FALSE
```

This biplot (Figure above) shows that, on principal components 1 and 2 (the PCs that account for the __most__ variance in the training data) there is subtle but noticeable grouping of hospitals in terms of how long their readmitted pneumonia patients stayed in the hospital. I have chosen to display the __#1__ feature that contributes to that variance, `Communication.about.medicines`. 

Each quadrant on a PCA biplot corresponds to a different directional influence of the original variables. E.g., because it is in the top-right quadrant I where both PC1 and PC2 are positive, this suggests that communication about medicines are positively associated with variables that point into the same quadrant. Since it seems to correspond with where the "better" hospitals are, this suggests that these hospitals are made somewhat distinctly "better" at readmitted hospital stay lengths because of correlations with communication about medicines. (**NOTE**: If you look at the 2025 data, the bipolot is effectively inverted but the conclusions seem to be the same in 2025 as well).

**A general rule of thumb is:**

1. Top-right Quadrant I
  - PC1 and PC2 are both positive
  - Thus points in this quadrant are positively associated with variables pointing into this quadrant
  - Indicates high values for variables with arrows in this quadrant

2. Top-left Quadrant II
  - PC1 is negative, PC2 is positive
  - Thus points here are associated with high values of variables pointing to the left and top
  - These observations contrast with those in the right-side quadrants on PC1
  
3. Bottom-left Quadrant III
  - PC1 and PC2 are both negative
  - Thus points here are negatively associated with variables pointing into the other quadrants
  - These points could represent a distinct subgroup or pattern with behavior opposite to other points.

4. Bottom-right Quadrant IV
  - PC1 is positive, PC2 is negative
  - Thus points here are associated with variables pointing _into_ this quadrant
  - May indicate a contrast with Quadrant II on PC2

#### **Question 22**: [1 point]

Play around with the graphical parameters in the biplot, especially those I've marked with comments. Try adding, for example, more variables to display rather than just the top contributor. **NOTE** that if you find variables in the same quadrant together, this means that they are correlated. What variable(s) can you find that **contrast** with `Communication.about.medicines`?

**Answer:**
After expanding the biplot to show the top 10 contributing variables, the contrasts become clear.

**Variables that cluster WITH `Communication.about.medicines`** (right side of the plot, around the same axis):
- Cleanliness
- Discharge.information
- Quietness
- Doctor.communication

**Variables that CONTRAST with Communication.about.medicines** (pointing into the bottom-left quadrant): 
- bc_PredictedReadmissionRate (our target), 
- Perioperative.pulmonary.embolism.or.deep.vein.thrombosis.rate, 
- Postoperative.respiratory.failure.rate, 
- Composite.patient.safety. 
The fact that readmission rates and surgical complications all point opposite to communication tells a coherent story: hospitals with strong patient communication tend to have lower complication rates, fewer readmissions, and better safety scores. These operational and clinical outcomes move together as a "well-functioning hospital" cluster.


Now, look at the contributions to PCs 1 and 2 (Figure Below) from the variables in the training dataset. Anything above the dotted red-line contributes **significantly** to the overall explanation of variance, and anything below the dotted line does not. Does anything about this plot stand out to you?

**HINT 1**: What do the top contributing variables all have in common? In other words, did they all come from the same data source?

**HINT 2**: Where does the target, `PredictedReasmissionRate`, fall among the significant variables?

**Answer:**
The top contributors split into two clear data-source clusters: the HCAHPS patient experience variables (Communication.about.medicines, Doctor.communication, Discharge.information, Quietness, Cleanliness) dominate the top of the chart, and the surgical safety variables (Composite.patient.safety, Postoperative.respiratory.failure.rate, Perioperative.PE/DVT) make up the second tier. Variables from the same data source tend to be correlated, which is exactly what PCA picks up on — that's why these clusters dominate the explained variance.

The target, bc_PredictedReadmissionRate, sits just above the threshold at ~5.6% — it barely qualifies as a significant contributor. This tells us our target shares some structural variation with the patient experience and surgical safety clusters, but most of its variation lives in higher-numbered PCs beyond PC1+PC2. The dominant axes of variation in the data aren't the dominant axes of variation in readmissions, which is why we'll need a different framework (the supervised modeling that's coming up) to truly understand what predicts readmissions.

```{r, echo=FALSE, fig.width=8, fig.height=8, fig.cap="Figure 12. Variables that Significantly Explain Variance."}
fviz_contrib(pca, 
             choice = "var", 
             axes = 1:2,
             fill = prettyPurples[13],
             color = prettyPurples[15],
             xtickslab.rt = 70)
```

## 5.2 *k*-means Clustering

$k$-means is a centroid-based clustering algorithm, where a "centroid" is the geometric center of an object often measured by Euclidean distance. In the algorithm, the distance between each data point and a centroid is calculated by randomly grabbing data; these distances are then used to assign the data point to a cluster. The goal is to identify the optimal $k$ number of groups in a dataset by minimizing the distance of each datapoint to a respective centroid.

### 5.2.1 Visualizing Clusters

Although multiple methods exist to determine the optimal number of clusters, one commonly employed is __heuristic__, i.e., it isn't really computational. You will either choose the optimal clusters based on a set of options of $k$ graphically or you can once again employ the "elbow" method again, as you did with PCA. Let's start by visually inspecting the effect of different $k$ clusters on PC1 and PC2 of the dataset. I have chosen to test $k$ groups of 2 through 9. 

```{r, echo = F}
skittles <- c("#de378d","#32a840", "#463ab5","#0e87e3", "gold", "#ae5bc9", "#e35c0e", "cadetblue", "#0ee383")
for(k in 2:9) {
  ## Initialize a plot
  plotName <- paste0("p", k)
  ## Calculate the kmeans for each given k
  kmeansResult <- kmeans(readyTrain, 	
                     centers = k,   ## number of clusters
                     nstart = 25,   ## num of times to repeat the process 
                                    ## with random initialization; increase 
                                    ## if you're failing to converge
                     iter.max = 1000,    ## num of iterations to run k-means
                     algorithm = "MacQueen")    ## since the default algorithm 
                                                ## can struggle with close 
                                                ## points, adjusting the method
  ## Make a set of graphs to compare!
  kmeansGraph <- fviz_cluster(kmeansResult, 
                              geom = "point", 
                              data = readyTrain,
                              show.clust.cent = TRUE,
                              ggtheme = theme_minimal(),
                              ellipse.type = "norm",
                              palette = skittles,
                              pointsize = 0.5) + 
                  ggtitle(paste0("k = ", k))
  assign(plotName, kmeansGraph)
}
grid.arrange(p2, p3, p4, p5, nrow = 2)
grid.arrange(p6, p7, p8, p9, nrow = 2)
```

##### Figure 13. Biplots showing possible k clusters

#### **Question 23A**: [0.5 points]
Which number of clusters, $k$, do you think has the best explanatory power? Why? Is it is hard to tell?

**Answer:**

Based on visual inspection alone, k = 2 produces the cleanest segmentation. Two clusters cleanly split along Dim1 (the "patient experience" axis from the biplot) with minimal overlap and clear interpretation: hospitals with better patient communication/satisfaction vs. those without.

k = 3 is also defensible — it carves out a middle group that could represent "average" hospitals if stakeholders want a three-tier framework.

Beyond k = 4, the clusters become heavily overlapping and arbitrary. By k = 6 and higher, you can barely distinguish clusters visually — the algorithm is forcing divisions on data that doesn't naturally support them. This is consistent with the PCA findings: with variance spread across many roughly-equal dimensions, the data doesn't have natural "tight groups" — just gradients.

I think k = 2 as the optimal heuristic choice, pending confirmation from a formal method like the elbow method on within-cluster sum of squares or the silhouette score.

### 5.2.3 Within-sum-of-squares (WSS) elbow method

Next, let's employ the 'elbow' method again, this time to look at when the __within sum of squares__ drops off rather than the percent variance, as was done in PCA.

```{r, echo = FALSE, fig.cap="Figure 14. WSS Scree plot for k-means"}
# Determine number of clusters
fviz_nbclust(readyTrain,
             FUNcluster = kmeans, 
             method = "wss",  
             linecolor = prettyPurples[13]) +
  labs(title = "WSS Elbow Method for Optimal k")
```

#### **Question 23B**: [1 point]

What is the __within sum of squares__ and what exactly is it measuring here? By the elbow method, which $k$ is the optimal number of clusters? Did it agree with your choice from the other heuristic method?

**Answer:**
The ‘within sum of square’ is a measure of the total distance between each data point and the centroid of the cluster it belongs to. In measuring terms, it is the sum of squared Euclidean distances from each point to its cluster’s center. 

with above plot, the WSS drops from k = 1 to k = 2 or 3, after the rate of decrease slows down.It shows K = 2 the optimal number of clusters. yes, it agree with my choice from the other heuristic method

### 5.2.3 Silhouette method

Lastly, we will apply the __silhouette method__, which measures the average silhouette width, a combination of (1) how similar any given point is to its own cluster (also called __cohesion__) and (2) how different that same point is from other clusters (also called __separation__). Silhouette scores can range from $-1 \rightarrow +1$, with larger values indicating better clustering (i.e., more cohesive clusters that are separated from each other). Thus, a silhouette score of 1 indicates __perfect separation and cohesion__ whereas a silhouette of 0 indicates overlapping clusters. 

What we can do is iteratively test our possible $k$ values of 2 to 9 and a silhouette score will be calculated for each $k$ tested. The highest silhouette score indicates the optimal $k$.

```{r, echo=FALSE, fig.cap="Figure 14. Silhouette Plot for k-means"}
fviz_nbclust(readyTrain, kmeans, method = "silhouette",
             k.max = 9, 
             nstart = 25,
             iter.max = 1000,
             linecolor = prettyPurples[13]) +
  labs(title = "Silhouette Method for Optimal k")
```

#### **Question 23C**: [0.5 points]

Are you surprised at the optimal $k$ from the silhouette method? Does it seem to agree with what you've seen with the WSS elbow method? (**Hint**: agreement generally suggests evidence that you have found the optimal $k$).

**Answer:**
No, I'm not surprised — the silhouette method clearly identifies k = 2 as optimal (peak at width ~0.147, dropping sharply to ~0.113 at k = 3), which strongly agrees with the WSS elbow method. This convergence is exactly the kind of evidence the hint describes: when two methods using different criteria agree, it's strong evidence we've found the optimal k. WSS measures compactness while silhouette measures separation — these are different concepts, so their agreement strengthens our confidence.


Lastly, let's extract the optimal $k$ from the silhouette method.

```{r}
## Extract optimal k from silhouette method; first get the cluster data
clusterData <- fviz_nbclust(readyTrain, kmeans, method = "silhouette")$data
## Then extract the maximum y
k <- as.numeric(clusterData$clusters[which.max(clusterData$y)])
```


## 5.3 Segmentation Analysis

As we said before, the goal of segmentation analysis is to broadly cluster our hospitals by their features to help us characterize them. 

#### **Question 24**: [1 point]
Let's explore the clusters - segments - of the hospitals based on their average values from the __original__ dataset (we start from `imputedTrain`, which hadn't yet been transformed, centered, or scaled, BUT the multicollinear variables also haven't been dropped either). Your task is to add comments to this code chunk. 

```{r, echo = TRUE, fig.height = 8, fig.width=6, fig.cap="Figure 16. Boxplot Comparisons of Important Variables by Cluster"}
## run k-means clustering on the transformed dataset.
kmeansResult <- kmeans(readyTrain, 	
                         centers = k,      ## set the number of culsters
                         nstart = 50,      
                         iter.max = 1000,  
                         algorithm = "MacQueen")
## Creates a new version of the original training dataset with clusters attached
kmeansTrain <- imputedTrain %>% 
  ## Rename the column we want to keep
  rename(Pneumonia.Hospital.Return.Days = `ComparedToNational_Hospital return days for pneumonia patients`) %>% 
   ## Drops multicollinear variables, comparison scores, payment categories, and survey response rate
  select(-any_of(vifOver5), -contains("Compared"), -contains("PaymentCategory"), -SurveyResponseRate) %>% 
   ## Adds cluster assignment as a new variable and makes pneumonia return days column numeric
  mutate(Cluster = kmeansResult$cluster,
         Pneumonia.Hospital.Return.Days = as.numeric(if_else(Pneumonia.Hospital.Return.Days == "1", 1,
                                                     if_else(Pneumonia.Hospital.Return.Days == "0", 0, -1)))) %>% 
   ## Cleans column names to lowercase with words separated by underscores 
  clean_names()

## Using % contribution graph, the order of the top 10:
pcaImportantVars <- factor(c("Communication.about.medicines",
                       "Doctor.communication", 
                       "Composite.patient.safety",
                       "Discharge.information",
                       "Quietness",
                       "Postoperative.respiratory.failure.rate",
                       "Cleanliness",
                       "Periop.Rate.Of.Embolism.Or.Thrombosis",
                       "Predicted.Readmission.Rate",
                       "Pneumonia.Hospital.Return.Days"))

## Converts variable names to lowercase for consistent formatting
pcaImportantVars <- tolower(pcaImportantVars)
## Replaces periods with underscores
pcaImportantVars <- gsub("\\.", "_", pcaImportantVars) %>% 
## Converts to title case for better readability in plots
  to_any_case("title")

## Creates boxplots to see important variables across clusters
kmeansTrain %>% 
  ## Renames long variable names 
  rename(Predicted.Readmission.Rate = predicted_readmission_rate,
         Periop.Rate.Of.Embolism.Or.Thrombosis = perioperative_pulmonary_embolism_or_deep_vein_thrombosis_rate) %>% 
  ## Converts column names to title case, clean presentation for plots
  clean_names(case = "title") %>% 
  ## Selects only the cluster column and the top 10 important variables
  select(Cluster, any_of(pcaImportantVars)) %>% 
  ## Reshapes the data into long format
  pivot_longer(-1, names_to = "Measure", values_to = "Value") %>% 
  ## Sets order of the measures  in plot
  # filter(Measure %in% pcaImportantVars) %>% 
   mutate(Measure = factor(Measure, levels = to_any_case(c("Communication.about.medicines",
                        "Doctor.communication", 
                        "Composite.patient.safety",
                        "Discharge.information",
                        "Quietness",
                        "Postoperative.respiratory.failure.rate",
                        "Cleanliness",
                        "Periop.Rate.Of.Embolism.Or.Thrombosis",
                        "Predicted.Readmission.Rate",
                        "Pneumonia.Hospital.Return.Days"), case="title"))) %>% 
  ## Creatre boxplots to compare distributions of each measure across clusters
    ggplot(aes(x = Cluster, y = Value, fill = as.factor(Cluster))) +
    geom_boxplot(alpha=0.75) +
    scale_fill_manual(values = c(skittles[1], skittles[6])) +
    labs(title = "Cluster Means",
         x = "",
         fill = "Cluster") +
    theme_classic() +
    theme(legend.position = "bottom",
        strip.text = element_text(size=8),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 10),
        title = element_text(size = 12, color = "maroon")) + 
    ## Separates the plots by measure, adjusting scales individually and arranging them in 2 columns
    facet_wrap(~ Measure, 
             scales = "free_y",
             ncol = 2)

## Creeate a summarytable for each cluster with mean and median of each important variable
kmeansTrain %>% 
  ## Renames the key variables to match the clean names
  rename(Predicted.Readmission.Rate = predicted_readmission_rate,
         Periop.Rate.Of.Embolism.Or.Thrombosis = perioperative_pulmonary_embolism_or_deep_vein_thrombosis_rate) %>% 
  ## Cleans names to title case
  clean_names(case = "title") %>% 
  ## Selects only cluster and important variables
  select(Cluster, any_of(pcaImportantVars)) %>% 
  ## Creates a summary table grouped by cluster
  tbl_summary(by = "Cluster",
              statistic = list(all_continuous() ~ c("{mean} ({median})"))) %>% 
  modify_header(label ~ "Variable",
                all_stat_cols() ~ "**Cluster {level}**") %>% 
  modify_caption("Table 8. Summary Statistics of Important Cluster Variables")

```

**HINT**: If these are back to the _original_ values, this is more interpretable for our stakeholder. We will now read this just like a typical boxplot; for example, median (the middle line) `Doctor Communication` seems higher in **Cluster 1** than it does in **Cluster 2** with only some overlap in the distributions (the boxes and whiskers don't really overlap each other). The table confirms this; the median `Doctor Communication` score is 91% in Cluster 1 but only 88% in Cluster 2.   

#### **Question 25**: [2 points]
Notice, for example, that Cluster 1 hospitals seem to include the hospitals that had the __lowest__ pneumonia-related readmission rates, with the average rate ~2% higher in Cluster 2 hospitals. Thus, for our stakeholder, we might decide to rename Cluster 1 as "Highest Performing Hospitals" or something along those lines. Notice also that this cluster had more __below__ national average return hospital days (a "1"). This means that they had more readmitted pneumonia patients who spent less time in the hospital upon readmission than the national average. 

Look at how else you might broadly characterize - or segment - these hospitals. Your job is to make a table for your client, summarizing each of the two clusters with FOUR primary criteria. Make sure to give each cluster an informative name as I did; you're welcome to rename Cluster 1 if you think of a catchier name than I chose!

After you come up with names of the two clusters, you will then come up with at least FOUR characteristics per cluster summarizing what you see in either the boxplots or table. You may choose to focus on different attributes for different clusters. Fill in the code for the table below. I start you out with two examples for Cluster 1, but feel free to add more! 

```{r, echo = FALSE, collapse=TRUE}
dict <- tribble(
~`Top Defining Attributes`, ~`Description`,
"Predicted pneumonia-related hospital readmissions", "These hospitals have a lower readmission rate, suggesting they may be better at diagnosing and treating pneumonia.",
"Return hospital days due to pneumonia", "Readmitted patients spend fewer days in the hospital than the national average, suggesting these hospitals more quickly treat and stabilize pneumonia patients.", 
"Doctor communication", "These hospitals score higher on provider-to-patient communication, suggesting doctors better understand patient needs and provide more attentive care.",
"Cleanliness and quietness", "These hospitals score higher on cleanliness and quietness — indicators of a comfortable, sanitary, well-managed environment that supports patient recovery.",
"Predicted pneumonia-related hospital readmissions", "These hospitals have a higher readmission rate, suggesting they may be less effective at pneumonia treatment or discharge planning.",
"Return hospital days due to pneumonia", "These hospitals have a greater proportion of patients spending average or more days in the hospital upon readmission, suggesting more severe complications or less efficient care.",
"Communication about medicines", "These hospitals show weaker communication about medications, which can lead to patient confusion about prescriptions and improper at-home treatment.",
"Patient safety indicators", "These hospitals have higher Composite Patient Safety scores, suggesting more safety-related incidents and complications during care."
)
dict %>% 
  kable(
    format = "html",
    caption = "Table 9. Hospital segmentation analysis: two broad clusters of hospitals identified.") %>%
    pack_rows("1: Highest-Performing Hospitals", 1, 4) %>%
    pack_rows("2: Lowest-Performing Hospitals", 5, 8) %>%
    kable_styling(bootstrap_options = c("striped", full_width = F))
```

## 5.4 Final Remarks on Segmentation

We don't technically have to stop our segmentation analysis here, but we will so that we can focus on other analyses. One thing we might choose to do in the future, for example, is to choose a classification machine learning algorithm to test the predictions based on these  clusters. 

#### **Question 26**: [1 point]
Make a recommendation for our client for 1-2 machine learning analyses __your team__ would choose to use to test whether these clusters can be used for robust predictions of hospital performance for pneumonia patients. How would you know whether or not your predictions were robust? What would you look for or compare?

**Answer:**
I would recommend two complementary classification models to predict cluster membership from hospital attributes:

1. **Random Forest Classifier** — handles nonlinear relationships and high-dimensional data well, can rank feature importance, and would help us understand which hospital characteristics best differentiate the clusters.

2. **Logistic Regression** — serves as a strong, interpretable baseline. If a simple linear model performs nearly as well as random forest, we'd prefer it for ease of communication with stakeholders.

To evaluate robustness, we would compare both models on Accuracy, Precision, Recall, F1 Score, Confusion Matrices, and AUC-ROC, using 5-fold cross-validation to ensure stable performance on held-out data. We'd also examine feature importance (random forest) and coefficients (logistic regression) to confirm that the meaningful variables — Communication.about.medicines, Doctor.communication, Composite.patient.safety — are driving predictions rather than noise.

We'd consider predictions robust if: (1) both models achieve similar accuracy with AUC > 0.75, (2) performance is stable across cross-validation folds, and (3) the most important features align with what we identified in the segmentation analysis. If all three hold, we can confidently report to the advocacy group that the clusters represent real, predictable categories of hospitals.


# 6 Supervised Learning Methods

We are going to perform two supervised learning methods: an Ordinary Least Squares (OLS) regression followed by an Elastic Net. 

## 6.1 Ordinary Least Squares (OLS) regression 

Earlier in [Section 4]{#multicollinearity} when we assessed multicollinearity, we fit an OLS regression model so that we could calculate variance inflation factors (VIFs). However, because OLS regression is so incredibly sensitive to multicollinearity, we should re-run it now that we've removed some of the most redundant variables. 

Let's model that now, using the `readyTrain` dataset and plot some diagnostic plots. to assess the key assumptions of an OLS linear regression. The four plots are diagnostic plots for the primary assumptions of a linear regression: 

  1. __Linearity__ between the outcome and the predictors

  2. __Normally distributed__ residuals

  3. __Homosecadasticity__ (equality of variances) in the residuals

  4. __No high leverage/influence points__ (no significant outliers)


```{r, echo=FALSE}
## Fit the model
mod <- lm(bc_PredictedReadmissionRate ~ ., 
          data = readyTrain)

## Plot diagnostic plots
plot(mod, 1)
plot(mod, 2)
plot(mod, 3)
plot(mod, 4)
```

##### Figure 17. OLS Regression Assumption Diagnostic Plots

#### **Question 27A**: [1 point]

What is your general assessment of our OLS assumptions? Met or unmet? Why? 

**Answer:**
The OLS assumptions are well met. Linearity is mostly met — residuals scatter randomly around 0 with a near-flat trend line, with only minor curvature at the edges that could be picked up by more flexible models. Normality is met both visually (Q-Q plot is near-perfect on the reference line) and quantitatively: the Shapiro-Wilk test on residuals returns W = 0.99862 with p = 0.0673, meaning we fail to reject normality. This is especially convincing because with N = 2,210, Shapiro-Wilk has enormous statistical power and would flag even tiny deviations. Homoscedasticity is met (flat trend in scale-location plot), and no observation has dangerously high influence (max Cook's distance ~0.030, well below 0.5). The Box-Cox transformation and VIF cleanup we did earlier clearly paid off — OLS has an unusually clean foundation for this data.


**Note:** If you are not entirely sure about the Q-Q plot, you can run a Shapiro-Wilk test on the residuals using the following code:

```{r}
shapiro.test(residuals(mod))
```

The **most important** OLS assumption for Elastic Net is __linearity of the relationship between predictors and the response__. The other, stricter OLS assumptions don't matter as much because of regularization.

**Do you find support for linearity here**? Why or why not?

**Answer:**
Yes, the linearity assumption is supported. In the Residuals vs. Fitted plot, the red trend line is mostly flat across the bulk of the fitted value range, with residuals scattering randomly around 0 — no obvious systematic pattern like a parabola, S-curve, or fan shape. The only caveat is very subtle curvature at the extreme edges, suggesting there may be mild non-linear effects for hospitals at the very high or very low ends of the prediction range. But this curvature is minor, and the bulk of the data lies in the middle where linearity holds well.

For Elastic Net, this is good news: regularization will dampen any residual non-linearity while still capturing the dominant linear signal. The slight edge curvature is the kind of pattern that more flexible methods like Random Forest might exploit, which is part of why testing a non-linear alternative is worth doing.

2. Let's now assess how well our model fits with the adjusted $R^2$.

```{r, echo = FALSE}
## Extract just the R-squared
summary(mod)$adj.r.squared
```

#### **Question 27B**: [1 point] {#rmse}
It is a surprisingly high (although it isn't stellar). You may need to do some outside research, but explain in a sentence or two why, generally speaking, we use **adjusted** $R^2$ and not the unadjusted one (confusingly labeled as `multiple r-squared` in the `summary()` output below). 

**Answer:**
Multiple R² has a critical flaw: it never decreases when we add predictors, even if those predictors are noise. Every added variable lets the model fit the training data slightly better by chance, so multiple R² always inflates as the model grows. This makes it unreliable for comparing models with different numbers of predictors.

Adjusted R² fixes this by penalizing extra predictors based on degrees of freedom: it only rises if a new predictor adds genuine explanatory power, and falls if the predictor just adds noise. For our 17-predictor model, adjusted R² = 0.394 gives us an honest estimate of fit — about 39% of variance explained after accounting for the cost of using that many variables. Multiple R² would inflate this slightly without that adjustment.

3. Find the best model. 

Use the `step()` function to do a backward regression to find our __most parsimonious__ model. Stepwise regression drops terms that are not significant or important, thereby only keeping the features that contribute to the overall explanation of variance in predicted readmission rate.

```{r, results='hide'}
## Perform a backward, stepwise regression to find the most parsimonious model
bestMod <- step(mod, 
            direction = "backward", 
            trace = FALSE)
```

Lastly, let's take a peek at the results. We will use a package called `stargazer` to help us visualize the results in a nicely organized way.

```{r, results='asis', warning=FALSE}
## Print a table using stargazer
stargazer(bestMod, 
          type = "html",
          title = "Table 10. Parsimonious OLS Regression Results.")
```

Lastly, let's assess how good a predictive model our OLS regression model is. We will use the `predict()` function to first fit our most parsimonious model `bestMod` to the test data:

```{r}
## Make the predictions
predictions <- predict(bestMod, newdata = readyTest)
```

The Root Square Mean Error ($RMSE$) is the square root of the average of the squared differences between predicted and actual value, such that $RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^n\cdot(y_i-\hat{y}_i)^2}$ where $y_i$ is each observed value, $\hat{y}_i$ is each corresponding predicted value, and $n$ is the number of observations. **NOTE**: Recall that a residual is $y_i-\hat{y}_i$ - the distance between every point and its predicted value! It's also important to realize that the $RMSE$ gives more weight to large errors because of squaring. Thus, it's sensitive to outliers.

#### **Question 27C**: [1 point]
Can you interpret what the RMSE tells your client here? Why or why not? What about the $R^2$?

**Hint 1**: Remember that the outcome here is Box-Cox transformed. Would you need to undo the transformation to be able to assess what the RMSE (or $R^2$) is actually telling us? Recall that the Box-Cox $\hat{y} = \frac{1}{y^\lambda}$ where our $\lambda = -0.3$ AND we also centered and scaled these data. 

**Hint 2**: Is it practical to undo both transformations? How does that hinder our ability to make an interpretation for our client?

**Answer:**
The RMSE is hard to interpret directly because our target is on a transformed scale — first Box-Cox with λ = -0.3, then centered and scaled. So an RMSE of ~0.78 isn't "0.78 percentage points of readmission rate" — it's in fully abstracted statistical units with no clinical meaning. To get back to interpretable units we'd have to undo the centering/scaling (multiply by SD, add mean), then invert the Box-Cox transformation. This is technically possible but practically clunky because Box-Cox is non-linear: an error of 0.78 transformed units corresponds to a *different* real-world error depending on where the hospital sits on the readmission rate spectrum.

R² is the more interpretable metric for client communication, because it's dimensionless. We can tell the client: "Our model explains about 39.5% of the variation in hospital-level pneumonia readmission rates." This is intuitive, meaningful, and doesn't require any inverse transformation. The trade-off is that 39.5% is a modest finding — about 60% of hospital-level variation comes from factors outside our quality metrics (patient case mix, social determinants of health, etc.).

This is a classic interpretability cost of pre-processing: the choices that improve statistical performance (Box-Cox for skew, centering for regularization) come at the price of clear stakeholder communication. If client-facing interpretation were critical, we'd fit a parallel model on the untransformed target purely for explanation, even if its statistical properties were slightly worse.

Lastly, let's graph the effect predictive ability of our OLS model by comparing the relationship between the **true** values of `bc_PredictedReadmissionRate` from the `readyTest` dataset to the predictions we made using the model on the testing data.

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 18. Predictions Plot for OLS Regression"}
## Add the predictions to the dataset
readyTest$predictions <- predictions

## Graph what the predictions look like relative to the actual data
ggplot(readyTest, 
       aes(x = predictions, y = bc_PredictedReadmissionRate)) +
       geom_point(alpha = 0.5, color = "hotpink") +
       geom_smooth(color = "hotpink", fill = "hotpink", se = T, method = "lm") +
  theme_minimal() +
  labs(title = "OLS Multiple Linear Regression Predictive Power",
       x = "Predicted Readmission Rate from Model",
       y = "Actual Readmission Rate")
```

**NOTE** that, while there is an obvious linear trend, it is weak. There is a lot of error around the predictions from the OLS regression!

## 6.2 The Elastic Net

Perhaps the seemingly low $R^2$ is because we still have too much multicollinearity or overfitting. So, as we discussed in lecture, we can address this using regularized regressions, such as the __Elastic Net__, which is more robust to multicollinearity by assessing penalties for non-zero variance. Typically, we would likely conduct a Ridge and/or LASSO before running an Elastic Net, but since we already know that (1) we have a fairly high amount of multicollinearity in our original `readyTrain` dataset and (2) we know we want to be able to perform feature reduction and figure out which features are important predictors of hospital readmissions, we're going to jump straight into Elastic Net, which performs both.

### 6.2.1 The Elastic Net Unpacked

In general, regularization is a technique that applies a penalty term to a cost function of a machine learning model, like the OLS regression. The reason is to discourage overfitting. This penalty term constrains the model's coefficients, which limits how flexible they are during training. Why? 

The more flexible the coefficients, the less likely they will "learn" new data, i.e., the more generalizable they are! Thus, by applying this penalty (constraint), it improves the model's performance on "unseen" or new data.

As mentioned in lecture, Elastic Net regression is a general regularization technique that combines the regularization techniques of Ridge and LASSO to help us perform both feature selection (i.e., significant coefficients) and feature reduction (importance). The elastic net can even help us with the $P >> n$ problem! 
Elastic net has two regularization terms, called __L1__ (or the Lasso regularization term) and __L2__ (the Ridge term).

* **L1**: makes some of the coefficients zero, thereby selecting only the most important features (shrinks them to zero). It is often said that this __encourages sparsity in the model__.

* **L2**: reduces the coefficients to small but non-zero values, thereby reducing the coefficients of the unimportant features (which we can use to determine significance). It does this by adding a penalty term to the cost function of the model, which is proportional to the coefficient squared. This allows us to retain all the features in the model but reduces the overall impact of the non-significant or less-important ones.

Thus, by combining these techniques together, the Elastic Net is balancing feature selection with feature reduction! If you're interested in reading the original paper on Elastic Net by Zou and Hastie (2005), you can find a free PDF of the paper [here](https://www.stat.purdue.edu/~tlzhang/mathstat/ElasticNet.pdf).

### 6.2.2 Running, tuning, & cross-validating the Elastic Net

The elastic net can be run using the `glmnet` package (which we are accessing via `caret`), and can be cross-validated as random forest and SVM from our last project could be. This is a big improvement over OLS regression, which cannot be tuned. Additionally, the elastic net has __two hyperparameters__ that we can tune, $\alpha$ and $\lambda$ (not to be confused with the $\lambda$ from Box-Cox transformation - Greek letters are abundant in statistics! Why do you think I named our course teams as I did?!). 

* $\alpha$ is the strength of the regularization, representing the balance between __L1__ and __L2__ regularization. Larger $\alpha$ results in L1 regularization (feature selection / Ridge) whereas smaller $\alpha$ results in more L2 regularization (higher shrinkage / feature reduction / Lasso). When $\alpha=0$ the regression is equivalent to OLS regression! $\alpha$ is also referred to as the __mixing percent__ of L1 and L2 regularization.


* $\lambda$ is the shrinkage parameter, such that when $\lambda = 0$ no shrinkage is performed and is equivalent to an OLS regression. As $\lambda$ increases, the coefficients are increasingly shrunk and thus the more features will have coefficients shrunk to zero, regardless of the value of $\alpha$. 


#### 6.2.2.1  We will start by setting our CV parameters just as we did on the last project, using the `trainControl()` function in `caret`. 

**NOTE** that I am performing a 5-fold repeated cross-validation (CV) with 5 iterations.

```{r}
ctrl <- trainControl(method = "repeatedcv",  ## Do repeated CV
                     number = 5,             ## Number of k-folds
                     repeats = 5,            ## Number of repeats
                     search = "grid",        ## Grid search (vs. random)
                     verboseIter = FALSE)    ## Don't show me all the results
```

#### 6.2.2.2 Then, we fit the model using the `train()` function in `caret`, specifying that we want to perform an elastic net, which will tune the model and perform CV per the specifications in the `trainControl()`. **NOTE** that I am allowing the values of $\alpha$ to range from $0 \rightarrow 1$ and allowing $\lambda$ to range from $0 \rightarrow 5$. 

```{r}
## Create a search grid that allows alpha to range from 0 --> and 
## allows lambda to range from 0 --> 5
searchGrid <- expand.grid(.alpha = seq(0, 1, length.out = 10), 
                          .lambda = seq(0, 5, length.out = 15))

## Fit the elastic net using the tuning grid and CV scheme laid out
elasticMod <- train(bc_PredictedReadmissionRate ~ ., 
                    data = readyTrain, 
                    method = "glmnet", 
                    tuneGrid = searchGrid,
                    trControl = ctrl) 

## Confirm the model object exists and is complete
class(elasticMod)
elasticMod$bestTune              ## Best alpha and lambda combination
elasticMod$results %>% 
  arrange(RMSE) %>%
  head(5)  
```

#### **Question 28**: [1 point]
Explain what we are doing here. You may want to look up the parameters of the `trainControl()` and `train()` functions. Make sure to address what type of CV and hyperparameter search is being performed, as well as if both the $\alpha$ and $\lambda$ hyperparameters of elastic net are being tuned. 

**Hint 1**: Why are we doing cross-validation, and specifically why did I choose the one I did?

**Hint 2**: What does $\lambda = 0$ mean? How about $\alpha = 0$? 

**Answer:**
We are using repeated 5-fold cross-validation with grid search to train an Elastic Net regression model, simultaneously tuning both: 𝛼
, which balances L1 (LASSO) and L2 (Ridge) penalties, and 𝜆, which controls the overall strength of the regularization. This approach helps mitigate overfitting, addresses multicollinearity, and selects the most predictive features, making the final model more robust and interpretable.


### 6.2.3 Elastic Net Results

#### 6.2.3.1 We are going to first take a look at the top 15 results of the Elastic Net in a table form, sorted from lowest $RMSE$:

```{r, echo = FALSE, message=FALSE}
elasticMod$results %>% 
  data.frame() %>% 
  arrange(RMSE) %>% 
  top_n(15) %>% 
kable(digits = 2,
    format = "html",
    caption = "Table 11. Top Performing Results of the Elastic Net Tuning & 5-fold Repeated Cross-Validation") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
)
```

**NOTE**: Folds that fail to converge will have an $R^2$ = `NA`.

#### 6.2.3.2 We can also extract the hyperparameters of our best tuned model:

$\alpha$:
```{r, echo=FALSE}
elasticMod$bestTune$alpha
```

$\lambda$:
```{r,echo=FALSE}
elasticMod$bestTune$lambda
```

Let's also plot the results of the tuning search. Can you figure out the type of tuning search you did? (**Hint**: the visual pattern should confirm your answer from earlier!) The diamond denotes the best combination of hyperparameters!

```{r}
results <- elasticMod$results %>% 
  data.frame()
best <- results[as.numeric(rownames(elasticMod$bestTune)), ]
```

```{r, echo = FALSE,  fig.cap="Figure 19. Hyperparamater Grid Search for Elastic Net"}
results %>% 
  drop_na() %>% 
  mutate(lambda = round(lambda, 1)) %>%  
  ggplot(aes(x=alpha, y = RMSE, color = as.factor(lambda))) +
  geom_point() +
  theme_minimal() +
  geom_point(aes(x = best$alpha, best$RMSE), 
             shape = 5, 
             size = 3,
             color = "black") +
  labs(title = "Elastic Net Hyperparameter Tuning",
       subtitle = "Best mixing percentage + λ shown as a diamond",
       x = expression(alpha), 
       color = expression(lambda)) + 
  scale_color_manual(values = c(skittles, "#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#00AFBB", "#E7B800", "#FC4E07"))
```

#### 6.2.3.3  Important Features

Important features can also be extracted, using the `varImp()` function from `caret`.

```{r}
important <- varImp(elasticMod)$importance
```

```{r, echo = FALSE, fig.cap="Figure 20. Variable Importance from Best Elastic Net Model"}
important %>% 
  mutate(Feature = rownames(important)) %>% 
  mutate(Feature = gsub("\\.", " ", Feature)) %>% 
  arrange(desc(Overall)) %>% 
  ggplot(aes(y = Overall, fill = Overall, x = fct_reorder(Feature, Overall))) +
  geom_col(color = "navy") + 
  scale_fill_continuous(low = "#DCCAE3", high = "#C00E50") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Feature importance as determined by \nElastic Net",
       x = "",
       y = "Importance", 
       fill = "")
```

#### **Question 30**: [1 point]

Which of the features were most important? How does this compare to the most important features out of OLS regression? Make sure you're looking at the __absolute value of the beta coefficients to assess importance in OLS regression__. 

**Answer:**
The two models produce remarkably consistent feature rankings — the top variables are virtually identical in both Elastic Net and OLS.

**Most important features in Elastic Net (top 5):**
1. Hospital.return.days.for.pneumonia.patients (importance = 100, dominating all others)
2. Medicare.spending.per.patient (~40)
3. Death.rate.for.pneumonia.patients (~38)
4. State (~26)
5. Venous.Thromboembolism.Prophylaxis (~22)

**Most important features in OLS (by |β|, top 5):**
1. Hospital.return.days.for.pneumonia.patients (|β| = 0.396, dominating all others)
2. Medicare.spending.per.patient (|β| = 0.167)
3. Death.rate.for.pneumonia.patients (|β| = 0.154)
4. State (|β| = 0.110)
5. Venous.Thromboembolism.Prophylaxis (|β| = 0.101)

The top 5 variables and their ordering are *identical* in both models. The pattern continues down the list — Quietness, Doctor.communication, Left.before.being.seen, Cleanliness, and Discharge.information appear in similar positions in both rankings. The bottom-ranked variables (Composite.patient.safety, Postoperative.respiratory.failure.rate, Perioperative.PE/DVT, Communication.about.medicines) are near-zero in both.

This agreement makes sense methodologically: when severe multicollinearity has already been addressed (as we did in Section 4 with VIF cleanup), Elastic Net with mild regularization tends to converge on similar solutions to OLS. Both methods are essentially finding the same underlying signal in the data.

**The single most important predictor in both models is Hospital.return.days.for.pneumonia.patients**, which makes clinical sense — it's a direct measure of how each hospital handles pneumonia patients on readmission, derived from CMS's own benchmark of return-day performance. Its importance is more than 2x the next-strongest predictor in both models.

The patient-mix story comes through clearly in both models too: Medicare.spending.per.patient is the second-strongest predictor in both rankings. This confirms what we saw in earlier sections — hospital readmission rates are driven heavily by patient case mix, not just hospital quality.

The fact that two different statistical approaches with different objectives (OLS minimizes squared error, Elastic Net minimizes squared error + regularization penalty) produce essentially the same feature ranking is a form of *triangulation*. It strengthens our confidence that the patterns we're seeing are real, not artifacts of either method.

How about about of PCA? You may have already noticed that importance almost seems inverted! Why?!

**Hint 1**: What is the fundamental difference between a PCA and any regression?

**Hint 2**: What does "importance" mean in a PCA vs. any regression?

**Hint 3**: What does the sign of the loadings mean in a PCA? Are they actual or arbitrary?

**Answer:**
The apparent inversion makes sense once you recognize that PCA and regression answer fundamentally different questions.

**PCA is unsupervised** — it knows nothing about our target. It finds the directions of greatest variation *among the predictors themselves*, so its "important" variables are those that vary substantially AND move together with other variables. That's why the HCAHPS satisfaction cluster dominated PCA — five variables that all move together collectively define PC1+PC2.

**Regression is supervised** — it explicitly searches for relationships between predictors and the target. Its "important" variables are those that directly move bc_PredictedReadmissionRate when held against the others. Hospital.return.days.for.pneumonia.patients dominates regression but is bottom-ranked in PCA because it varies somewhat independently from the HCAHPS cluster while having a direct relationship with readmissions.

The signs in PCA are also arbitrary (mathematically undetermined direction), while regression coefficients have meaningful, fixed signs indicating positive vs. negative effects.

Bottom line: PCA describes the *structure of variation* in hospital characteristics (patient satisfaction is the dominant axis), while regression describes *what drives outcomes* (LOS performance and Medicare spending). The features most variable across hospitals aren't necessarily the features that determine readmissions — that's the deeper insight the apparent "inversion" reveals.

#### 6.2.3.2 Check model performance

Assessing __model performance__ means assessing how well our training model does when applied to "unseen" data, here our `readyTest` dataset. Because we know the right answer - the true value of `PredictedReadmissionRate`, we can assess the error that the training model produces when fitted against the testing data. We measure that error as $RMSE$ and $MAE$. We briefly reviewed what the $RMSE$ measures back in [Question 36]{#rmse}, so I will only review what the $MAE$ is here.

Mean Absolute Error ($MAE$) is the average of the absolute differences between predicted and actual values, such that $MAE = \frac{1}{n}\sum_{i=1}^n\cdot|y_i-\hat{y}_i|$ where $y_i$ is each observed value, $\hat{y}_i$ is each corresponding predicted value, and $n$ is the number of observations. So, unlike $RMSE$, $MAE$ treats all errors equally and therefore is more robust to outliers. So, $MAE$ will tell you the average magnitude of the error, regardless of direction!

Let's also make predictions on the testing sets for both the OLS and Elastic Net models using the `postResample()` function in `caret`; this will not only do predictions, but it will enable us to more easily generate metrics that we can use to compare model performance, such as $R^2$ and MAE.

```{r}
## Post-resample on OLS
prOLS <- postResample(pred = predict(bestMod, newdata = readyTest), 
                      obs = readyTest$bc_PredictedReadmissionRate)

## Post-resample on Elastic Net
prElastic <- postResample(pred = predict(elasticMod, newdata = readyTest), 
                      obs = readyTest$bc_PredictedReadmissionRate)
```

```{r, echo=FALSE}
## Display the output as a table
rbind(prOLS, prElastic) %>% 
  kable(digits = 2,
    format = "html",
    caption = "Table 12. Comparison of Model (Test) Performance of OLS and Elastic Net Regressions") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
)
```

#### 6.2.3.3 Check for overfitting/underfitting

Using the same method, check for overfitting / underfitting. Make predictions using `readyTrain` rather than `readyTest`, and then check how the $RMSE$ on the training data compares to the $RMSE$ of the training data for the Elastic Net. How do they compare?  __How do both compare to what we got out of the OLS regression?__ 

```{r}
## Post-resample on OLS
prOLS <- postResample(pred = predict(bestMod, newdata = readyTrain), 
                      obs = readyTrain$bc_PredictedReadmissionRate)

## Post-resample on Elastic Net
prElastic <- postResample(pred = predict(elasticMod, newdata = readyTrain), 
                      obs = readyTrain$bc_PredictedReadmissionRate)
```

```{r, echo=FALSE}
## Display the output as a table
rbind(prOLS, prElastic) %>% 
  kable(digits = 2,
    format = "html",
    caption = "Table 13. Comparison of Training Performance of OLS and Elastic Net Regressions") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
)
```

#### TAKEAWAY
When it comes to which performed better, it's obvious that the Elastic Net only marginally (very marginally!) out-performed the OLS regression or the Elastic Net (Table 12). The $R^2$ is a 0.01 higher in the Elastic Net model and both measures of error are a teensy bit lower. But, truthfully, they're performing the same __because they are the same model__. The Elastic Net likely out-performed simply because it was cross-validated!

To interpret $RMSE$ from training (Table 13) vs. testing (Table 12), we're looking first to see if they are the same/balanced. However, we can see that error ($RMSE$, $MAE$) is a bit higher in the testing post-resample vs. the training post-resample. While this could suggest that we have some mild overfitting (our model memorized the training data a little too well leading to poor generalization when it "saw" the testing data), instead I think it is more likely is that we have __underfitting__ due to a lack of complexity! Underfitting error patterns are such that you will see roughly the same error in both the testing and training sets, and both error rates are quite high.

#### 6.2.3.4 Graphical comparison of the two models' predictions

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 19. Predictions Plot for Elastic Net vs. OLS Regression"}
elasticPredictions <- predict(elasticMod, newdata = readyTest)
testingResults <- c(readyTest$predictions,
                    elasticPredictions)
label <- c(rep("OLS Prediction", nrow(readyTest)),
           rep("Elastic Net Prediction", length(elasticPredictions)))
temp <- cbind.data.frame(testingResults, label)
temp$actual <- rep(readyTest$bc_PredictedReadmissionRate, 2)

## Graph what the predictions look like relative to the actual data
ggplot(temp, 
       aes(x = testingResults, y = actual, fill = label, color = label)) +
       geom_point(alpha = 0.5) +
       geom_smooth(se = T, method = "lm", alpha = 0.45) +
  scale_color_manual(values = c("navy", "hotpink")) +
  scale_fill_manual(values = c("navy", "hotpink")) +
  theme_minimal() +
  labs(title = "Elastic Net vs. OLS Predictive Power",
       x = "Predicted Readmission Rate from Model",
       y = "Actual Readmission Rate",
       fill = "", 
       color = "") +
  theme(legend.position = "bottom")
```

**It should not come as a surprise, but these models are performing VERY similarly!**


#### **Question 32**: [4 points]
What conclusions do you make at this point about whether to make predictions with an OLS regression, Elastic Net, or neither? What recommendations would you make to our client at this point? What other analyses might your suggest, or other data we might want to include? Please try to flesh out 2-3 recommendations in at least a paragraph.

**HINT**: Make sure to ask yourself whether these data appear linear or not. Could non-linearity be a cause for a low $R^2$? If not, what other causes are there and how will that impact your recommendations?

**Answer:**
Based on the analyses so far, **neither OLS nor Elastic Net should be deployed as-is for predicting hospital readmission rates.** Both models converged on essentially the same solution (R² ≈ 0.34-0.35 on the test set, RMSE ≈ 0.81, MAE ≈ 0.64) and explain only about a third of the variation in pneumonia readmissions.  This level of accuracy isn't strong enough to support consequential decisions about hospital comparison or intervention prioritization. Below are two recommendations for moving forward.

* Recommendation 1: Test whether the issue is non-linearity by fitting a tree-based model.
OLS and Elastic Net models are peformaing silmilarly, and training R² = 0.40, test R² = 0.34 — that gap is the underfitting. We should try a Random Forest or Gradient Boosting model that can capture non-linear interactions automatically. If a tree-based model substantially outperforms our linear models, we'll have evidence that the ceiling on R² wasn't due to limits in the data — it was due to limits in linear modeling. If the tree-based model performs *similarly*, then we've learned the data genuinely lacks predictive signal and no amount of fancy modeling will help.

Recommendation 2: Acknowledge limits in CMS data and recommend expanding the predictor set.

Our PCA analysis showed hospital variation is genuinely multidimensional — PC1 explained only 22% of variance, and 11 principal components were needed to reach 80%. This tells us hospital performance has no single dominant axis. expanding the data set with additional variables that could impact readmission ralte could lead to better accurate predictions. 

When two different model classes (OLS and Elastic Net) converge on the same solution and both underperform, the problem is rarely in the model — it's in the data or the problem framing. We should be honest with the client: linear modeling has hit a ceiling here, and our next steps should test whether more flexible models can push past it (Recommendation 1) or whether we need to fundamentally expand what we're measuring (Recommendation 2). 

# 7 Choose Your Own Adventure

It's time to choose your adventure! After working through this demo and answering the 31 questions, you will choose one of three possible trajectories. The following are some brief hints or tips for each of the choices available to you.

## Adventure 3. 
**[Continue analyzing the pneuomnia data.]**

- Make sure to justify your choice.

- Choose at least two new algorithms to perform. They can be unsupervised or supervised, but must be 2+.

- A natural extension would be to do a Ridge and/or a LASSO; the former would tell you if you should be controlling for multicollinearity more than you currently are, the latter would tell you if you need to further shrink coefficients than you currently are. Additionally, you could consider a regression-based method that incorporates **non-linearity**, such as a random forest. However, you are free to choose anything you think is appropriate here! 
- Other choices I think you could potentially justify to stakeholders, if you were interested: 

1. Switch to classification-based random forest using `ComparedToNational_Hospital return days for pneumonia patients`

2. kNN using the 2 clusters we've identified and named in our segmentation analysis

3. Our data are also ready for a neural network analysis (e.g., a feed-forward NN)!

### 7.1 Data Preparation
```{r}
load(file = "FY2024_data_files/readyTrain.Rdata")
load(file = "FY2024_data_files/readyTest.Rdata")

## Drop the high-VIF variables
vifOver5 <- c("Nurse.communication", "Staff.responsiveness", "Care.transition", 
              "Overall.hospital.rating", "Recommend.hospital")
readyTrain <- readyTrain %>% select(-any_of(vifOver5))
readyTest <- readyTest %>% select(-any_of(vifOver5))

## Rename the kept ComparedToNational variable
if("ComparedToNational_Hospital.return.days.for.pneumonia.patients" %in% names(readyTrain)) {
  readyTrain <- readyTrain %>% 
    rename(Hospital.return.days.for.pneumonia.patients = 
             ComparedToNational_Hospital.return.days.for.pneumonia.patients)
}
if("ComparedToNational_Hospital.return.days.for.pneumonia.patients" %in% names(readyTest)) {
  readyTest <- readyTest %>% 
    rename(Hospital.return.days.for.pneumonia.patients = 
             ComparedToNational_Hospital.return.days.for.pneumonia.patients)
}

## Drop the other ComparedToNational, PaymentCategory, and SurveyResponseRate
readyTrain <- readyTrain %>%
  select(-contains("Compared"), -contains("PaymentCategory"), -any_of("SurveyResponseRate"))
readyTest <- readyTest %>%
  select(-contains("Compared"), -contains("PaymentCategory"), -any_of("SurveyResponseRate"))

## Verify dimensions and structure
all(names(readyTrain) == names(readyTest))
names(readyTrain)
sum(is.na(readyTrain))
sum(is.na(readyTest))

## Confirm dimensions
dim(readyTrain)   ## Should be 2210 x 18
dim(readyTest)    ## Should be 516 x 18

## Confirm column 18 is the target
names(readyTrain)[18]   ## Should be "bc_PredictedReadmissionRate"

```

### 7.2 Create X and Y Matrices

```{r}
X_train <- as.matrix(readyTrain[, -18])
y_train <- as.matrix(readyTrain[, 18])
X_test  <- as.matrix(readyTest[, -18])
y_test  <- as.matrix(readyTest[, 18])

## Sanity check
dim(X_train)    ## 2210 x 17
dim(y_train)    ## 2210 x 1
dim(X_test)     ## 516 x 17
dim(y_test)     ## 516 x 1
```


### 7.3 Ridge regression (α = 0)
Ridge regression reduces coefficient instability caused by multicollinearity.

```{r}
## Fit Ridge and select best lambda via CV
set.seed(42)
ridge_fit <- glmnet(X_train,y_train,alpha=0)
ridge_cv <- cv.glmnet(X_train,y_train,alpha = 0, family = "gaussian")

#using cross-val to choose the best lambda
ridge_bestlam <- ridge_cv$lambda.min
cat("Ridge best lambda:", ridge_bestlam, "\n")

## Plot CV curve
plot(ridge_cv)
title("Ridge CV: log(λ) vs Mean Squared Error", line = 2.5)
```
#### Ridge regression — test set performance and coefficients
```{r}
## Predict on test set
ridge_pred <- predict(ridge_fit, s = ridge_bestlam, newx = X_test)

## Test set metrics
ridge_perf <- postResample(pred = as.vector(ridge_pred),
                           obs  = as.vector(y_test))
cat("Ridge test performance:\n")
print(round(ridge_perf, 3))

## Look at the coefficients
ridge_coefs <- predict(ridge_fit, s = ridge_bestlam, type = "coefficients")
cat("\nRidge coefficients at best lambda:\n")
print(ridge_coefs)
```


### 7.4 LASSO regression (α = 1)

```{r}
## Fit LASSO and select best lambda via CV
set.seed(42)
lasso_fit <- glmnet(X_train, y_train, alpha = 1)
lasso_cv  <- cv.glmnet(X_train, y_train, alpha = 1, family = "gaussian")

## Best lambda
lasso_bestlam <- lasso_cv$lambda.min
cat("LASSO best lambda:", lasso_bestlam, "\n")

## Plot CV curve
plot(lasso_cv)
title("LASSO CV: log(λ) vs Mean Squared Error", line = 2.5)
```
#### LASSO regression— test set performance and coefficients

```{r}
## Predict on test set
lasso_pred <- predict(lasso_fit, s = lasso_bestlam, newx = X_test)

## Test set metrics
lasso_perf <- postResample(pred = as.vector(lasso_pred),
                           obs  = as.vector(y_test))
cat("LASSO test performance:\n")
print(round(lasso_perf, 3))

## See which features LASSO kept (non-zero coefficients)
lasso_coefs <- predict(lasso_fit, s = lasso_bestlam, type = "coefficients")
cat("\nLASSO coefficients at best lambda:\n")
print(lasso_coefs)

## Count how many were zeroed out
n_zero <- sum(lasso_coefs[, 1] == 0) - 1   ## minus 1 for intercept
cat("\nLASSO zeroed out", n_zero, "of 17 predictors\n")

```

### 7.5 Random Forest 
```{r}
## Random Forest with default settings, ntree=500
## We'll tune mtry by trying a few values
set.seed(42)

## Tune mtry: try a range from sqrt(p) to p/2
mtry_values <- c(3, 5, 7, 9, 11)
rf_results <- data.frame(mtry = mtry_values, OOB_MSE = NA)

for (i in seq_along(mtry_values)) {
  rf_temp <- randomForest(x = X_train, y = as.vector(y_train),
                          mtry = mtry_values[i],
                          ntree = 500,
                          importance = TRUE)
  rf_results$OOB_MSE[i] <- rf_temp$mse[500]   ## OOB MSE at final tree
  cat("mtry =", mtry_values[i], ": OOB MSE =", round(rf_temp$mse[500], 4), "\n")
}

## Pick best mtry
best_mtry <- rf_results$mtry[which.min(rf_results$OOB_MSE)]
cat("\nBest mtry:", best_mtry, "\n")

## Refit final RF with best mtry
set.seed(42)
rf_fit <- randomForest(x = X_train, y = as.vector(y_train),
                       mtry = best_mtry,
                       ntree = 500,
                       importance = TRUE)
print(rf_fit)
```
#### Random Forest — test set performance and importance
```{r}
## Predict on test set
rf_pred <- predict(rf_fit, newdata = X_test)

## Test set metrics
rf_perf <- postResample(pred = as.vector(rf_pred),
                        obs  = as.vector(y_test))
cat("Random Forest test performance:\n")
print(round(rf_perf, 3))

## Variable importance
cat("\nRandom Forest variable importance:\n")
importance_df <- as.data.frame(importance(rf_fit))
importance_df <- importance_df %>%
  arrange(desc(`%IncMSE`))
print(round(importance_df, 2))

## Plot variable importance
varImpPlot(rf_fit, main = "Random Forest Variable Importance")
```

### 7.6 Final comparison table — all 5 models
```{r}

## Refit Section 6 models on the current readyTrain
## (necessary because Section 7's data load may invalidate the Section 6 models)

# Full OLS
mod <- lm(bc_PredictedReadmissionRate ~ ., data = readyTrain)

# Stepwise OLS
bestMod <- step(mod, direction = "backward", trace = FALSE)

# Elastic Net (same setup as Section 6)
ctrl <- trainControl(method = "repeatedcv", 
                     number = 5, 
                     repeats = 3, 
                     verboseIter = FALSE)
searchGrid <- expand.grid(.alpha = seq(0, 1, length.out = 10), 
                          .lambda = seq(0, 5, length.out = 15))
set.seed(42)
elasticMod <- train(bc_PredictedReadmissionRate ~ ., 
                    data = readyTrain, 
                    method = "glmnet", 
                    tuneGrid = searchGrid,
                    trControl = ctrl)

cat("Models refit successfully\n")


prOLS <- postResample(pred = predict(bestMod, newdata = readyTest), 
                      obs = readyTest$bc_PredictedReadmissionRate)

prElastic <- postResample(pred = predict(elasticMod, newdata = readyTest), 
                      obs = readyTest$bc_PredictedReadmissionRate)

## Build comparison table
comparison_table <- rbind(
  OLS         = prOLS,
  ElasticNet  = prElastic,
  Ridge       = ridge_perf,
  LASSO       = lasso_perf,
  RandomForest = rf_perf
)

cat("Table 14. Test Set Performance Comparison Across 5 Models\n\n")
print(round(comparison_table, 3))
```
#### Predictions vs Actuals plot — all 5 models

```{r}
## Build the plotting dataframe (must come before the ggplot)
plot_df <- bind_rows(
  data.frame(Model = "OLS",
             Predicted = as.vector(predict(bestMod, newdata = readyTest)),
             Actual    = as.vector(readyTest$bc_PredictedReadmissionRate)),
  data.frame(Model = "Elastic Net",
             Predicted = as.vector(predict(elasticMod, newdata = readyTest)),
             Actual    = as.vector(readyTest$bc_PredictedReadmissionRate)),
  data.frame(Model = "Ridge",
             Predicted = as.vector(ridge_pred),
             Actual    = as.vector(y_test)),
  data.frame(Model = "LASSO",
             Predicted = as.vector(lasso_pred),
             Actual    = as.vector(y_test)),
  data.frame(Model = "Random Forest",
             Predicted = as.vector(rf_pred),
             Actual    = as.vector(y_test))
)

## Reorder so Random Forest is last/highlighted
plot_df$Model <- factor(plot_df$Model,
                        levels = c("OLS", "Elastic Net", "Ridge", "LASSO", "Random Forest"))

ggplot(plot_df, aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.4, color = "steelblue", size = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen", linewidth = 0.8) +
  facet_wrap(~ Model, nrow = 1) +    ## All in one row
  theme_minimal() +
  labs(title = "Predicted vs. Actual Readmission Rate by Model",
       subtitle = "Red dashed line = perfect prediction (y=x); Green = model fit",
       x = "Predicted (Box-Cox transformed)",
       y = "Actual (Box-Cox transformed)")
```
### 7.7 Final Conclusion: 

Looking at Table 14, the three hypotheses from Q32 get pretty clear answers. Ridge, LASSO, and Elastic Net all landed at basically the same place as OLS — R² between 0.344 and 0.349, RMSE around 0.81. So regularization wasn't the fix. LASSO only zeroed out one predictor (Postoperative.respiratory.failure.rate), and Ridge's coefficient shrinking didn't budge the test scores. Honestly, this isn't that surprising — back in Section 6, when the Elastic Net's tuning picked α = 0 and λ = 0, it was already telling us the linear models were doing their best.

Random Forest is the one that actually improved in this. Test R² = 0.40, RMSE = 0.776. The 43% OOB R² in training means it's real improvement, not overfitting. And it makes sense — the OLS diagnostic plot had that little curve at the edges that linear models couldn't capture, and that's exactly what trees are good at.

The cool part is that all the models agree on what matters. Hospital.return.days.for.pneumonia.patients is the top predictor in OLS, Elastic Net, AND Random Forest. State, Medicare spending, and patient communication variables also show up consistently. Three very different algorithms pointing at the same features = that's signal, not noise.

Random Forest is the strongest model I've got, but it's not solving everything. Even at R² = 0.40, more than half the variation in readmissions is unexplained. That confirms what I suspected — hospital-level CMS data just can't capture the patient-level stuff that really drives readmissions (comorbidities, social determinants, follow-up access). So rather than trying to predict exact readmission rates, I'd suggest using these models to identify which hospitals fall into the "needs intervention" bucket (using the variable importance and the k-means clusters together) and target resources that way. it may help us to understand more. 


## Deliverables

I am looking for 1-2 markdown documents with their knitted HTML. For example, if you're choosing Adventures 1 or 2, you may want to work through this document once, make a copy of this markdown document, move the source code to the __TOP__, edit the source code as needed, and then go through this again from the beginning using the new condition(s) you chose. 

Alternatively, if you're choosing Adventure 3, maybe you just choose to add on to the bottom of this document, replacing the 'Choose Your Own Adventure' section with your actual adventure.

Just make sure to answer the questions well,and make sure to justify the decisions you make. Tell me WHY you're choosing the condition(s) you are in Adventures 1 or 2, or Tell me WHY you're doing the analyses you are in Adventure 3. Other than that, make this your own learning adventure!

# References

