---
title: 'Demo 1: Biomedical & Clinical Informatics'
author: "Lucia Adeola"
date: "18 May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
always_allow_html: true
subtitle: 'Merrimack College DSE6630: Healthcare & Life Sciences Analytics'
bibliography: references.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      cache.comments = FALSE,
                      size = 13)
```

```{r, echo = FALSE, message=TRUE, include=FALSE}
## ====== SAFETY WARNING ======
## This script will clear your environment if you answer 'YES' to the prompt. 
## This only applies to interactive sessions, not knitting.

## Prompt user for confirmation in interactive sessions
if (interactive()) {
  message("This script will clear your environment. Proceed with caution.")
  confirm <- readline(prompt = "Type Y to clear the environment: ")
  if (toupper(confirm) == "Y") {
    rm(list = ls())
    gc()  ## Garbage collector - frees up memory
    cat("Environment cleared.\n")
  } else {
    #cat("Environment NOT cleared.\n")
  }
} else {
  ## In non-interactive sessions (e.g., RMarkdown), assume clean start is intended
  rm(list = ls())
  gc()
}
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Turn off scientific notation
options(scipen=999)

# Set seed
set.seed(50009)

# Install pacman if it's missing:
if (!requireNamespace("pacman", quietly = TRUE)) { install.packages("pacman") }

# Clean, set up, and load
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
pacman::p_load(tidyverse,
               janitor,
               naniar,
               stringr,
               ggplot2, 
               kableExtra,
               RColorBrewer,
               gridExtra,
               ggrepel,
               e1071,
               caret
)
```

###This was a combined effort between Lucia and Alison for all questions, however Lucia took the lead for questions 1-11, and Alison on questions 11-22.###


# Introduction

You have just been brought on as a data science consultant for a patient advocacy watchdog. They have asked you to source information about hospital readmissions, because they would like to flag specific geographic areas or types of hospitals where hospital readmission is unexpectedly high. They have heard that datasets exist to investigate these types of questions, but they are not really sure where to begin.

You remember this one data science class you took - and the dataset from the __Centers for Medicare & Medicaid Services__ managed by [Medicare.gov](https://www.medicare.gov). You start poking around the [Hospitals](https://data.cms.gov/provider-data/topics/hospitals) dataset and decide it is exactly what you need.

The problem is, this dataset is massive, so you are not only tasked with identifying a reasonable question you can ask from the dataset but managing the cleaning, pre-processing, and exploratory data analysis on the merged datasets you choose to explore.

## Question

For __Project 1__, you are going to choose your own question! But for this demo analysis, we are focusing on __pneumonia-related hospital readmissions__.

### Why pneumonia-related readmissions?

Hospital readmission rates for patients with pneumonia is almost unbelievably high, with about $\frac{1}{5}$ or 20% of patients hospitalized with pneumonia re-admitted within 30 days (de Alba and Amin, 2014). Further, pneumonia is both a leading cause of death among the elderly as well as a commonly communicated disease in hospitals (as in, people may come into a hospital for a different condition but be readmitted for pneumonia). Thus, given that pneumonia is the only communicable disease on this list, it is serving as our best proxy for diseases spread person-to-person within hospitals because of lack of availability.

#### Our research question:

Can we predict pneumonia-related hospital readmissions and, more generally, disease transfer within hospitals, based on hospital-level characteristics, including patient-ratings, Medicare money received, and other indicators of general performance in disease diagnosis and prevention?

#### Our hypothesis & prediction:

Hospitals are, by their very nature, hot zones for increased spread of disease, including pneumonia. We hypothesize that hospitals with lower indicators of patient care performance and lower Medicare funding rates will generally be unable to provide sufficient control of specific disease diagnosis and prevention, like pneumonia. This is because the same intrinsic and extrinsic factors that impact general patient quality of care is likely to extend to pneumonia, a _common, preventable, and treatable_ disease. Thus, we specifically predict that factors indicating lower quality-of-care will predict higher than average rates of pneumonia-related hospital readmissions.

## Objective

Our objective for the patient advocacy watchdog is clean, merge, explore, and predict which hospitals tend to have higher-than-average rates of pneumonia-related readmissions. This is to enable the watchdog group to put together a dashboard or otherwise deploy this information to share with patients so they know how their local hospitals are performing, allowing them to make healthier, more informed choices.

# Data & Data Cleaning

1. Download the [Hospitals](https://drive.google.com/drive/u/0/folders/1Z8VloDDl3_Gl9LhTQyxOS4v34r7xvkgq) dataset as a zip directory. NOTE: These 2024 and 2025 data were originally downloaded from the [CMS repository](https://data.cms.gov/provider-data/topics/hospitals). **Make sure to download it to your local machine NOT your GitHub repo!** I have given you both years' data, but for now we are still using ***2024***. All archived datasets are found [here](https://data.cms.gov/provider-data/archived-data/hospitals).
2. Unzip the directory. 
3. Make sure to adjust the name and/or path of the directory in your directory if it is different from mine. *Note that mine is living on my Desktop.*

**WARNING!** Do not try to put these data onto GitHub. They are too large to push. If you need help adjusting your path or have questions, please let me know.

```{r}
## Make sure to adjust your path!
filepath <- "~/Desktop/DSE66300M/Module 1/hospitals_current_data_FY2024/"
```



### Read in the hospital-level data iteratively, while dynamically naming & storing each file as a dataframe.

You may have to dig through the accompanying data dictionary **HOSPITAL_Data_Dictionary.pdf** to get a better handle on what is in the hospital-level files. 

##### **Question 1**: [1 point] 
Take a look at the chunk below, which performs **dynamic variable assignment** as it reads in each of the files. Your task is to add a comment to each line of code to describe what that line does.

> Your answer in the code chunk as comments. Make sure to use the `#` sign!

```{r}
#This creates a list of all the files in the unzipped folder "hospitals_current_data_FY2024 whose names contain "Hospital.csv"
files <- list.files(path = filepath, pattern = "Hospital.csv")

#starts a loop through each file in the files list one at a time
for(f in 1:length(files)) {
  
  #reads the current csv file via combining folder path and the file name
    dat <- read_csv(paste0(filepath, files[f]),
                    show_col_types = FALSE,
                    ## HOT FIX of the "N/A" characters that were problematic
                    na = c("", "NA", "N/A", "NULL", ".")
                    ) %>%
      #clean column names into upper camel format
        clean_names(case = "upper_camel") %>%
        
        ## HOT FIX to convert character to numeric
        type_convert() %>%
        ## HOT FIX to convert date columns based on month/day/year
        mutate(across(contains("Date"), ~ mdy(.)))
    
#remove "Hospital.csv" from the file name so that the dataframe name is the remaining texts
    filename <- gsub(".Hospital\\.csv", "", files[f])
    
    #create new dataframe using the cleaned file name and store "dat" inside of it
    assign(filename, dat)
}

#clean the file names via removing "Hospital.cvs" and convert into a dataframe
files <- gsub(".Hospital\\.csv", "", files) %>% data.frame()

#rename the column in the files dataframe to the unique "File Name'
names(files) <- "File Name"
```

#### Now, let's look at a list of the files that we've read in:
```{r, echo = FALSE}
files %>% 
  kable(
    format = "html",
    caption = "Table 1. List of hospital-level data files.") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
  )
```

##### **Question 2**: [1 point] 
Take a look at some of the dataframes, e.g., `HCAHPS` and `FY_2024_Hospital_Readmissions_Reduction_Program`. What do you notice about the format / structure of the data? Does the data have an issue with the dimensionality? 
**Hint: **Check to see if each row represents a single line of data or not!

> Your answer here.

```{r}
#view first few columns and rows
head(HCAHPS)
head(FY_2024_Hospital_Readmissions_Reduction_Program)
```
###After inspecting HCAHPS and FY_2024_Hospital_Readmissions_Reduction_Program, the first thing I noticed was that the data is not tidy because the same facility appears multiple times throughout the dataframe. Each row represents an individual measure or condition instead of one complete hospital-level observation. For example, Southeast Health Medical Center appear several times for different readmission measures and patient metrics. It would make more sense for each hospital to have a single row, while the different measures and conditions would become separate columns instead of thier own individual rows. This would make the dataset more organized and easier to analyze or compare across hospitals.




### Tidy the data response variables' dataframe: Fiscal Year 2024 Hospital Readmissions Reduction Program.
We are going to focus on the dataset `FY_2024_Hospital_Readmissions_Reduction_Program` to work on best practices to [tidy our data](https://vita.had.co.nz/papers/tidy-data.pdf). Although I do not personally choose to follow everything that Hadley Wickham and the __Tidy Data__ Movement has espoused, I do think a lot of the best practices are useful principles especially when we are just getting use to working with data. Further, we can leverage a lot of the functions to make our lives MUCH easier! And who doesn't like easier?!

##### **Question 3**: [1 point] 
Take a deeper look at `FY_2024_Hospital_Readmissions_Reduction_Program` at some of the numeric variables in the dataset, e.g., `NumberOfReadmissions`? 

> Your answer here.

**Hint: **Columns 5 & 7:10 have problems with the `NA` class.
```{r}
#preview the dataframe
glimpse(FY_2024_Hospital_Readmissions_Reduction_Program)

#check class(type of each column)
sapply(FY_2024_Hospital_Readmissions_Reduction_Program, class)
```

###After inspecting the dataset more closely, I noticed that NumberOfReadmissions, which should be a numeric variable, was instead imported as a character column. This is likely because it contains inconsistent missing value labels such as “NA” and other non-numeric entries such as “Too Few to Report.” 



##### **No Longer Do Question 4**: [1 point] 
Investigate the `replace_with_na_all()` function that is part of the `naniar` package. Use this function to fix the `NA` class in the aberrant columns in `FY_2024_Hospital_Readmissions_Reduction_Program`.

```{r}
# ?replace_with_na_all
# Your code here.
```

You will notice that this aberrant class of missing value __caused__ the numeric columns to be coerced into character columns when they were imported by `read_csv()`. Although we could try using `as.numeric()` by itself, which would assign `NA` to any non-numeric entry in `FY_2024_Hospital_Readmissions_Reduction_Program`, the problem is that we could lose important information that way. Thus, as annoying as this is, we should fix it a little more specifically...

##### **Question 5**: [1 point] 
Look more closely at the `NumberOfReadmissions` column in the `FY_2024_Hospital_Readmissions_Reduction_Program`. What other issue do you see with this specific column that is coercing it to the character type, and why does it exist? 

**Hint: ** Make sure to use the accompanying data dictionary **HOSPITAL_Data_Dictionary.pdf** to also examine **WHY** this problem exists!

> Your answer here.

```{r}
unique(FY_2024_Hospital_Readmissions_Reduction_Program$NumberOfReadmissions)
```
###Using unique() allowed me to further investigate the specific NumberOfReadmissions column by showing all of the distinct values contained in that variable. The first thing I noticed was the text value “Too Few to Report,” while the remaining values were numeric. This is likely the main reason the column was imported as a character type instead of a numeric type. According to the CMS data dictionary, some hospital results are not reported when the number of cases is too small to reliably compare against national averages.   Because of this, those entries are labeled as “Too Few to Report” instead of as numeric values.


##### **Question 6**: [1 point] 
Investigate the `gsub()` function that is part of base `R` or any other function of your choosing. Use the function you choose to **replace** the problematic text in  the `NumberOfReadmissions` column in `FY_2024_Hospital_Readmissions_Reduction_Program` with a **5**.
```{r}
#updating dataframe with the fixed version
FY_2024_Hospital_Readmissions_Reduction_Program <- 
  FY_2024_Hospital_Readmissions_Reduction_Program %>%
  #modify column by substituting "To Few to Report" with "5"
  mutate(NumberOfReadmissions = gsub(
    "Too Few to Report",
    "5",
    NumberOfReadmissions
  ))
```

##### **Question 7**: [1 point] 
Speculate as to why I asked you to replace the text with a **5** rather than a **0** or an `NA`. Can you think of any other logical choices for the substitution? Do you agree with my choice? (You're free to disagree, but please defend your decision either way!)

> Your answer here.

###I believe the text was replaced with 5 instead of 0 because “Too Few to Report” does not mean that there were no readmissions. According to the data dictionary, the value exists because the number of cases was too small to reliably compare against national averages. Therefore, using 0 would incorrectly suggest that no readmissions occurred at all. Replacing the value with 5 instead preserves the idea that some readmissions were documented, while still using a small enough number to reflect that the count was limited. This also allows the column to be converted into a numeric type so the dataset can be analyzed more effectively.



##### **Question 8**: [1 point] 
Now, imagine an alternative scenario where instead of replacing it with a **5**, I instead asked you to replace it with a randomly sampled integer from 1 to 10. What would this do? Would `gsub()` work here or would I need to execute it a different way, e.g., with an `lapply()` function or a `for()` loop?

> Your answer here.

###No, I do not think gsub() would work the same way in this scenario because we are no longer replacing the text with one exact value. Instead, we would first need some type of randomization or sampling logic to generate different integers from 1 to 10 for each occurrence of “Too Few to Report.” Because of this, I believe the workflow would likely require a function such as sample() along with some type of conditional replacement logic.


##### **Question 9**: [3 points] 
You probably anticipated this! We just decided that, rather than a **5** we want a randomly sampled integer from 1 to 10. 

**Hint 1: ** You might find it easier to read the data in from fresh and, after commenting out your `gsub` from Question 5, pattern match on the problematic text.

**Hint 2: ** This is NOT the only way to do this, but I solved this using `gregexpr()`, `regmatches()`, and `lapply()`.

```{r}
# makes the random numbers reproducible
set.seed(123) 

# Find every where "Too Few to Report" appears
matches <- gregexpr(
  "Too Few to Report",
  FY_2024_Hospital_Readmissions_Reduction_Program$NumberOfReadmissions
)

# Replace each matched text with a randomly sampled integer from 1 to 10
regmatches(
  FY_2024_Hospital_Readmissions_Reduction_Program$NumberOfReadmissions,
  matches
  #loops over each matched text
) <- lapply(
  regmatches(
    FY_2024_Hospital_Readmissions_Reduction_Program$NumberOfReadmissions,
    matches
  ),
  function(x) sample(1:10, length(x), replace = TRUE)
)
```


##### **Question 10**: [1 point] 
It is finally time to fix those aberrant character columns by making them  numeric. However, just using `as.numeric()` would require us to do it over and over on each of the columns in multiple lines of code. Instead, investigate the function `mutate_at()` which will allow you to pass it a list of columns you want to convert with the function you want to use.
```{r}
FY_2024_Hospital_Readmissions_Reduction_Program <- 
  FY_2024_Hospital_Readmissions_Reduction_Program %>%
  #modifying the specified columns from character to numeric
  mutate_at(vars(5, 7:10),
    as.numeric
  )

#check
glimpse(FY_2024_Hospital_Readmissions_Reduction_Program)
```

#### Let's look more closely at Measure Names.
Notice, that what I have done here first is used the `gsub()` function to `mutate()` the `MeasureName` column to remove the text that flanked the medical conditions.
```{r}
FY_2024_Hospital_Readmissions_Reduction_Program <-  FY_2024_Hospital_Readmissions_Reduction_Program %>%
  mutate(MeasureName = gsub("READM-30-", "", MeasureName)) %>% 
  mutate(MeasureName = gsub("-HRRP", "", MeasureName)) 
```

**Note:** If at this point you've had any issues with any of the previous problems 2-10, you can load the cleaned readmissions data, `readmissionsClean.Rdata` so that you can proceed with the assignment.

```{r}
load(file = "readmissionsClean.Rdata")
```

##### **Question 11**: [1 point] 
Investigate the function `pivot_wider()` or `spread()` from the `tidyr` package, which comes bundled with `tidyverse`. Try pivoting the `FY_2024_Hospital_Readmissions_Reduction_Program` dataset wider. Make sure you correctly identify the `names_from` and `values_from` column(s). **Make sure to save it as a separate dataframe called `wideDF`.** Use the `dim()` function to prove to me that you successfully pivoted wider.

```{r}
# create a new dataframe "wideDF" from the cleaned readmissions dataset
wideDF <- FY_2024_Hospital_Readmissions_Reduction_Program %>%
  
  # change the dataset from long format to wide format
  pivot_wider(
    
    # use the values in MeasureName as the names of the new columns
    names_from = MeasureName,
    
    # use readmission-related variables as the values that will be in the new columns
    values_from = c(
      NumberOfDischarges,
      ExcessReadmissionRatio,
      PredictedReadmissionRate,
      ExpectedReadmissionRate,
      NumberOfReadmissions
    )
  )

# check the dimensions of the new wide dataframe
dim(wideDF)
```

### Filtering for specific conditions.
Ideally, though, before we'd pivot wider we would decide if we were going to filter for any specific hospital readmission conditions. For your __Project 1__, you will be choosing which medical condition(s) you want to focus on for predicting hospital readmission. However, for the remainder of this demo, we're going to focus on just __pneumonia__. 

__Ideas you might consider for your Project_1 are:__

* surgical interventions (`HIP-KNEE`, `CABG`)
* heart-related conditions (`HF`, `AMI`, & `CABG`)
* all conditions (warning: this might be too unwieldy!)
* any other condition(s) that you motivate with a statement or two in support

Therefore, let's filter for just pneumonia-related readmissions which will actually eliminate the need to pivot wider in just this case (because we only chose a single condition).
```{r}
readmissionsClean <- 
  readmissionsClean %>% 
  filter(MeasureName == "PN")
```

```{r, echo = FALSE}
dict <- tribble(
  ~Acronym, ~Definition,
  "HIP-KNEE", "Total Hip/Knee Arthroplasty",
  "HF", "Heart Failure",
  "COPD", "Chronic Obstructive Pulmonary Disease",
  "AMI", "Acute Myocardial Infarction",
  "CABG", "Coronary Artery Bypass Graft",
  "PN", "Pneumonia"
)
dict %>% 
  kable(
    format = "html",
    caption = "Table 2. Acronyms of medical conditions for which hospital readmissions are tracked.") %>%
    kable_styling(bootstrap_options = c("hover", full_width = F)
  )
```

It's important to understand that these medical conditions are the **ones that Medicaid/Medicare tracks for hospital readmissions**. These are not the only conditions in which a patient might be readmitted, but these are the ones that the agency uses to keep track of hospital performance.

### Dynamic creation of merged dataframe objects: supporting dataframes.

We will often be asked to deal with large, unwieldy datasets and sometimes we need to be able to handle them dynamically. __Dynamic variable assignment__, as I demonstrated when we read in the files, is going to come in handy - but it can sometimes be tricky to accomplish. The whole point, though, is to create code that can handle ANYTHING you throw at it. 

In a similar vein, we can also perform __dynamic dataframe creation__ when we need to modify and stitch together more than one pre-existing dataframe - ESPECIALLY when we might decide down the line to alter our choices by adding or removing other dataframes.

#### Writing your own function to create objects dynamically.

We are going to take all the cleaning steps we did above that was not specific to the `FY_2024_Hospital_Readmissions_Reduction_Program` dataframe and tidy, pivot, & filter for specific readmission conditions, and iteratively join as many dataframes as needed. The goal is to make something flexible enough that, if you change a set of inputs, it will create a joined table from ANY set of dataframe inputs. There is more than one way to do this; but leveraging some of the existing `tidyverse` and `tidyr` functions will likely help you make this very flexible. Please note, however, that you do not have to use those packages.

I start you out with something I find helpful when joining together dataframes that sometimes have *redundant* columns BUT not every column is in every dataframe. I start by making a dataframe called `hospitalInfo` that contains the pertinent information on each hospital in the dataset, so you can drop those columns from the other dataframes to prevent inflation or issues when you later join the dataframes together. You may want to join each of the other dataframes to `hospitalInfo`. 

As I was writing my own version of this function, I found that the `Payment and Value of Care` table must be separated into two tables to join well, at least with how I wrote my function. So, I do that for you below. I also clean up `HCAHPS` to make it easier to work for the function I wrote, so it may also be useful for you. Note that what I'm doing in these cases is making sure that `MeasureName` is the title of the column **I will join my tables on.** In this way, I'm using the `MeasureName` as the **key** for my table joins. 


**First, make sure to separate the Payment and Values tables, giving each facility IDs:**

```{r}
paymentOnly <- Payment_and_Value_of_Care %>% 
  select(FacilityId, PaymentMeasureName, PaymentCategory, Payment) %>% 
  mutate(Payment = gsub("\\$", "", Payment)) %>% # Remove the dollar sign
  mutate(Payment = gsub("\\,", "", Payment)) %>% # Remove the comma 
  rename(MeasureName = PaymentMeasureName)  # Make consistent with other tables

valueOnly <- Payment_and_Value_of_Care %>% 
  select(FacilityId, ValueOfCareDisplayName, ValueOfCareCategory) %>% 
  rename(MeasureName = ValueOfCareDisplayName)  # Make consistent with other tables
```


**HCAHPS is also a mess. Remove the columns that we will never use in any analysis, make a `MeasureName` column, and also drop the " - linear mean score" from some rows in the newly minted `MeasureName`**

```{r}
HCAHPS <- HCAHPS %>% 
  select(-HcahpsMeasureId, -PatientSurveyStarRatingFootnote, 
         -HcahpsAnswerPercentFootnote, -NumberOfCompletedSurveysFootnote,
         -SurveyResponseRatePercentFootnote, -HcahpsAnswerDescription,
         -PatientSurveyStarRating) %>%  ## Drops the columns we don't need
  rename(MeasureName = HcahpsQuestion) %>%  ## Makes a new MeasureName column
  mutate(MeasureName = gsub(" - linear mean score", "", MeasureName))
  ## Takes off the phrase that we don't want
```

**Lastly, pull the hospital information off of the Payment and Value table because it is complete there. We can then drop this information from all the other tables:**

```{r}
hospitalInfo <- Payment_and_Value_of_Care %>% 
              ## Information about the hospitals
                select(FacilityId, FacilityName, Address, CityTown, 
                         State, ZipCode, CountyParish, TelephoneNumber)
```


##### **Question 12**: [5 points] 

Write a function that will:

1. Fix any issues identified previously, e.g., issues with the `NA` class or pivoting, if needed

2. Join any number of a list of dataframes you give it. For example, it should be able to join these 8 tables:

```{r}
datList <- list(Healthcare_Associated_Infections,
                paymentOnly,
                Outpatient_Imaging_Efficiency,
                Complications_and_Deaths,
                Medicare_Hospital_Spending_Per_Patient, 
                Timely_and_Effective_Care, 
                Unplanned_Hospital_Visits,
                HCAHPS)
```

3. Filters the data for given criteria and condition(s) prior to pivoting. For example, for the 8 tables listed above, perhaps to filter for these possible measures before pivoting, so that I'm not pivoting ALL of those measures! (That would likely crash R!). Notice that what we're filtering for - is really a selection criteria, but because the data are in **wide format**, we are selecting them by row (i.e., "filter") rather than selection by column (i.e., "select"). 

```{r}
filterList <- list("MRSA Bacteremia", 
                   "Payment for pneumonia patients", 
                   "Abdomen CT Use of Contrast Material", 
                   
                   c("Death rate for pneumonia patients", 
                     "Perioperative pulmonary embolism or deep vein thrombosis rate",
                     "CMS Medicare PSI 90: Patient safety and adverse events composite", 
                     "Postoperative respiratory failure rate"),
                   
                   "Medicare spending per patient",
                   
                   c("Healthcare workers given influenza vaccination", 
                     "Percentage of healthcare personnel who completed COVID-19 primary vaccination series", 
                     "Average (median) time patients spent in the emergency department before leaving from the visit A lower number of minutes is better", 
                     "Left before being seen",
                     "Venous Thromboembolism Prophylaxis", 
                     "Intensive Care Unit Venous Thromboembolism Prophylaxis", 
                     "Emergency department volume"),
                   
                   "Hospital return days for pneumonia patients",
                   
                   c("Nurse communication",
                     "Doctor communication",
                     "Staff responsiveness",
                     "Communication about medicines",
                     "Discharge information",
                     "Care transition",
                     "Cleanliness",
                     "Quietness",
                     "Overall hospital rating",
                     "Recommend hospital")
)
```

4. For full credit, make sure that your function can clean up and join together at least three of the dataframes. **Note**: Make sure to join with `readmissionsClean` at the end!

Perhaps you will choose to start your function like this (although you do not have to):
```{r, eval=FALSE, echo=TRUE}
tidyNjoin <- function(datList, filterList, hospitalInfo) {

}
```

* **Hint 1: ** Feel free to drop the `startDate` and `endDate` columns, as well as the `footnote` column. I would even suggest dropping `MeasureId`.

* **Hint 2: ** `select(-any_of(c(...)))` can be used to drop any of a list of columns, regardless of whether it exists in every dataframe or not.

* **Hint 3: ** If you choose to drop all those extraneous columns, then the column you will need to merge/join on will always be in the second position, `MeasureName`.

* **Hint 4: ** `full_join()` in `dplyr` is likely what you will need; you will want to execute the join on `FacilityId`.

* **Hint 5: ** Make sure to check for and/or remove duplicate rows.

* **Hint 6: ** Write ***pseudocode***, which is using English to write out our programming logic. Then, use that pseudocode to have generative AI help you craft the function. Make sure to test it well! GenAI is often very wrong!

* **Hint 6: ** If you are really stuck, move ahead and I give you a cleaned, merged data set to work with.

Okay, take a stab at it!

##Pseudocode:
#Input:
# - list of dataframes
# - list of filters

#For each dataframe:
  # - remove unnecessary columns
  # - apply the requested filter
  # - pivot wider if needed
  # - prepare dataset for joining

#Join all cleaned datasets together using FacilityId

#Remove duplicate rows

#Return final merged dataframe


```{r eval=FALSE, echo=TRUE}
# function to clean, filter, reshape, and join multiple hospital datasets
tidyNjoin <- function(datList, filterList, hospitalInfo) {
  
  # create an empty list where we will store each cleaned/wide dataset
  cleanedList <- list()
  
  # loop through each dataframe in datList
  for(i in 1:length(datList)) {
    
    # pull out one dataframe from the list
    tempDF <- datList[[i]]
    
    # drop columns that are not needed for joining or analysis
    # any_of() prevents errors if one of these columns does not exist in a dataframe
    tempDF <- tempDF %>%
      select(-any_of(c(
        "StartDate", "EndDate", "Footnote", "MeasureId",
        "FacilityName", "Address", "CityTown", "State",
        "ZipCode", "CountyParish", "TelephoneNumber"
      )))
    
    # keep only the measure(s) listed in the matching filterList entry
    tempDF <- tempDF %>%
      filter(MeasureName %in% filterList[[i]])
    
    # remove duplicate rows before pivoting
    tempDF <- tempDF %>%
      distinct()
    
    # pivot the dataframe wider so MeasureName values become columns/features
    tempDF <- tempDF %>%
      pivot_wider(
        names_from = MeasureName,
        values_from = -c(FacilityId, MeasureName)
      )
    
    # store the cleaned and widened daaset in cleanedList
    cleanedList[[i]] <- tempDF
  }
  
  # join all cleaned datasets together using FacilityId
  mergedDF <- reduce(cleanedList, full_join, by = "FacilityId")
  
  # join the merged dataset to hospitalInfo so hospital details are included once
  mergedDF <- hospitalInfo %>%
    full_join(mergedDF, by = "FacilityId")
  
  # join pneumonia readmissions data at the end
  mergedDF <- readmissionsClean %>%
    full_join(mergedDF, by = "FacilityId")
  
  # remove duplicate rows from the final merged dataset
  mergedDF <- mergedDF %>%
    distinct()
  
  # return the final cleaned and joined dataset
  return(mergedDF)
}

#final merged pneumonia dataset
pneumoniaFull_test <- tidyNjoin(datList, filterList, hospitalInfo)

dim(pneumoniaFull_test)
```

##### **Question 13**: [1 point] 
Why do you think I just made you work through this function? (**Hint**: What are you going to be asked to do in Project 1?)

> Your answer here.

##I believe we worked through this function because, in Project 1, we will need to create and answer our own healthcare-related analytical question using multiple datasets instead of relying on just one source of data. This function helped demonstrate how to clean, filter, reshape, and merge healthcare datasets into one final analytical dataset using a shared hospital identifier. Although the function initially seemed complex, writing out the pseudocode first made the workflow much easier to understand. Practicing this process now will also help us independently build and preprocess our own datasets later in the project.


### The full, merged, tidied dataset for pneumonia. 

Please load if you are struggling to get a function that performs the task or if you aren't sure if it does what's expected.

```{r}
load("pneumoniaFull.Rdata")
dim(pneumoniaFull)
```

As I mentioned in Question 12, I ultimately chose to merge 8 of the datasets together for this dataset, although I selectively filtered to focus on my question: **Can we predict pneumonia-related hospital readmissions based on factors that might indicate how well the hospital is doing at diagnostics and disease prevention?**

Whew! Notice that this dataset currently has **100 features** and **4,816 rows**. That's definitely too many features, so let's perform some additional feature selection.

**Let's start by examining the "big picture" of what's in this giant dataset**:

```{r, echo = FALSE, collapse=TRUE}
dict <- tribble(
~Dataset, ~`Information Used`, ~Rationale,
"Healthcare_Associated_Infections", "MRSA Bacteremia", "Hospital-transmitted disease rate",

"paymentOnly", "Mean Medicare payment per patient received by hospital", "",

"Outpatient_Imaging_Efficiency", "Abdomen CT Use of Contrast Material", "Proxy for imaging ability to diagnose pneumonia appropriately",

"Complications_and_Deaths", "Death rate for pneumonia patients", "",

"Complications_and_Deaths", "Perioperative pulmonary embolism or deep vein thrombosis rate", "Diagnostic ability of hospital",

"Complications_and_Deaths", "CMS Medicare PSI 90: Patient safety and adverse events composite", "General hospital safety",

"Complications_and_Deaths", "Postoperative respiratory failure rate", "General ability for respiratory diagnostics",

"Medicare_Hospital_Spending_Per_Patient", "Score", "Score rather than $ amount",

"Timely_and_Effective_Care", "Healthcare workers given influenza vaccination", "Likelihood of influenza transmission",

"Timely_and_Effective_Care", "% healthcare personnel completed COVID-19 primary vaccination series", "Likelihood of COVID-19 transmission",

"Timely_and_Effective_Care", "Average (median) minutes patients spent in the ED before leaving (lower is better)", "Proxy for overall effective diagnosis & treatment",

"Timely_and_Effective_Care", "Left before being seen in ED", "Proxy for overall effective diagnosis & treatment",

"Timely_and_Effective_Care", "Venous Thromboembolism Prophylaxis", "Diagnostic ability for another critical and silent disease",

"Timely_and_Effective_Care", "ICU Thromboembolism Prophylaxis", "Diagnostic ability for another critical and silent disease",

"Timely_and_Effective_Care", "Emergency department volume", "Proxy for how overwhelmed the hospital is",
    
"Unplanned_Hospital_Visits", "Hospital return days for pneumonia patients", "",

"HCAHPS", "Nurse communication", "Overall rating as linear score",

"HCAHPS", "Doctor communication", "Overall rating as linear score",

"HCAHPS", "Staff responsiveness", "Overall rating as linear score",

"HCAHPS", "Communication about medicines", "Overall rating as linear score",

"HCAHPS", "Discharge information", "Overall rating as linear score",

"HCAHPS", "Care transition", "Overall rating as linear score",

"HCAHPS", "Cleanliness", "Overall rating as linear score",

"HCAHPS", "Quietness", "Overall rating as linear score",

"HCAHPS", "Overall hospital rating", "Overall rating as linear score",

"HCAHPS", "Recommend hospital", "Overall rating as linear score",
)

dict %>% 
  kable(
    format = "html",
    caption = "Table 3. List of hospital-level data sets chosen for the pneumonia-related readmissions analysis.") %>%
    kable_styling(bootstrap_options = c("striped", full_width = F)
  )
```


# Pre-processing & Feature Selection (Round 1)

**WARNING**: Typically we would do a lot of this work AFTER splitting into a training and testing set. I'm not having you do that today so that you can write code that will help you with your Project1. Just be aware that this goes against the typical workflow a bit.

Feature selection can take many forms, from what we've already done (choosing features relevant to our question) to culling non-informative columns to using more machine-guided approaches. We're going to leverage all types here.

Additionally, we are going to perform **encoding** to our categorical variables. It will be important to perform encoding **BEFORE** using automated methods for feature selection.

## Culling Features (and, yes, more tidying!)

Although we ideally could just include these steps (and perhaps you did!) in our tidying & joining function from Question 12, I wanted to make sure we talk about this explicitly. 

##### **Question 14**: [1 point] 
Looking through the feature names, you probably notice names like `LowerEstimate_Death rate for pneumonia patients`. We do not need the higher and lower estimates in this case; just the point estimates. Use the `contains()` function to **drop** any columns that contain `LowerEstimate` or `HigherEstimate`. Let's also go ahead and drop any columns containing `Denominator` and `HcahpsAnswerPercent` as well. How many features did you drop? 

> Your answer here.

```{r}
pneumoniaFull <- pneumoniaFull %>%
  select(
    -contains("LowerEstimate"),
    -contains("HigherEstimate"),
    -contains("Denominator"),
    -contains("HcahpsAnswerPercent")
  )

dim(pneumoniaFull)
```

**Note:** Please drop `FacilityName`, `TelephoneNumber`, `Address`, & `CityTown` at this stage as well, if they're still in your full dataset.

```{r}
pneumoniaFull <- pneumoniaFull %>%
  select(
    -FacilityName,
    -TelephoneNumber,
    -Address,
    -CityTown,
  )

dim(pneumoniaFull)
```

## Encoding

### Mystery encoding! (Well, it won't be a mystery for long.)

##### **Question 15**: [1 point] 
We need to perform encoding on all of the categorical columns, ultimately. However, we will focus first on the columns that begin with ``ComparedToNational_`, e.g., `ComparedToNational_Death rate for pneumonia patients`. What kind of encoding should we perform on these columns? Why? 

> Your answer here.

```{r}
glimpse(pneumoniaFull$`ComparedToNational_Death rate for pneumonia patients`)

```

###We should use ordinal encoding for the columns that begin with 'ComparedToNational_'  because there is a clear hierarchy within the categories. For example, when looking at the ComparedToNationalDeathRateForPneumoniaPatients column, the values include categories such as “Worse than National Rate” and “No Different than National Rate,” which represent ranked performance levels rather than completely separate groups. Since modeling methods generally require numeric values instead of text labels, we need to convert these categorical values into numeric form. 





##### **Question 16**: [1 point] 
What other column do we also need to perform this kind of encoding on?

> Your answer here.

```{r}
glimpse(pneumoniaFull)
```

##Other columns that also need ordinal encoding are:
#ComparedToNational_MRSA Bacteremia
#ComparedToNational_Postoperative respiratory failure rate
#ComparedToNational_Perioperative pulmonary embolism or deep vein thrombosis rate
#ComparedToNational_CMS Medicare PSI 90: Patient safety and adverse events composite



##### **Question 17**: [3 points] 
You probably expected this - attempt to encode those columns the method you selected. It's okay if you can only think of a way to brute-force this for now; I will show you code to help you do it faster after the assignment. :)

* **Hint 1: **You could use the `contains()` or `startsWith()` functions to help you quickly grab those columns so you can make the changes you want to make.

* **Hint 2: **You may want to explore the `grepl()` function or regular expressions, as they can allow us to match in a "fuzzy" way.

* **Hint 3: **You may want to create a temporary dataframe that you execute the changes on and then replace the original columns from there. That way, you don't have to keep reloading the original data if you make mistakes!

* **Hint 3: **For full credit, don't forget to double check that you managed to preserve all the data! The `table()` function is sufficient here.

```{r}
# Create a temp dataset so we can test encoding without damaging pneumoniaFull
temp_pneumonia <- pneumoniaFull

# Find all columns that start with ComparedToNational_
cols2encode <- names(temp_pneumonia)[startsWith(names(temp_pneumonia), "ComparedToNational_")]

# Check which columns we are about to encode
cols2encode
```
```{r}
# encode all ComparedToNational_ columns in the temporary dataframe
temp_pneumonia <- temp_pneumonia %>%
  mutate(across(
    all_of(cols2encode),
    ~ case_when(
      grepl("Worse|More Days", ., ignore.case = TRUE) ~ 0,
      grepl("No Different|Average Days", ., ignore.case = TRUE) ~ 1,
      grepl("Better|Fewer Days", ., ignore.case = TRUE) ~ 2,
      TRUE ~ NA_real_
    )
  ))

# check the encoded values
lapply(temp_pneumonia[cols2encode], unique)

```


**Note**: If you struggled with Question 17, load the data below and keep going:
```{r}
load("pneumoniaFullEncoded.Rdata")
```



### Frequency encoding state and county.
We also decide that we want a way to try to preserve and analyze the geographic information without causing terrible overfitting. So, we will try to apply **frequency encoding** to those two categorical columns.


##### **Question 18**: [1 points] 

Go through the **frequency encoding** code chunk below and comment each line. What does it do? *Make sure to comment why you think the choice was made*, if appropriate.

> Your answer in comments in the code chunk.

```{r}
# identify the two geographic categorical columns that we want to encode
#frequency encoding to preserve geographic information
cols2encode <- c("State", 
               "CountyParish")

# create a temp dataset containing only the columns we want to encode
# this protects the original dataset incase we need to go back
temp <- pneumoniaFullEncoded[, names(pneumoniaFullEncoded) %in% cols2encode]

# create a function that replaces each category in a column with its frequency count
add_freq <- function(data, column_name) {
  
# count how many times each category appears in the selected column, including missing values
  frequency_map <- table(data[[column_name]], useNA = "always")
  
# replace each original category with its matching frequency count
  data[[column_name]] <- frequency_map[match(data[[column_name]],
                                             names(frequency_map))]
# return the updated dataset
   return(data)
}
# apply the frequency encoding function to each column in the temp dataset
for (col in names(temp)) {
  temp <- add_freq(temp, col)
}
# copy the encoded State and CountyParish columns back into pneumoniaFullEncoded
for (c in 1:length(cols2encode)) {
  pneumoniaFullEncoded[, cols2encode[c]] <- temp[, cols2encode[c]]
}
```             


##### **Question 19**: [1 point] 

Use your finally & freshly encoded version of `pneumoniaFullEncoded` to answer this vital question for `State` and `CountyParish` features: are we justified using these features? Why or why not? You will need to write code to answer this question.

**Hint 1: **Use your original `pneumoniaFull` dataset to compare the features in `pneumoniaFullEncoded`.

**Hint 2: ** **Heavily** duplicated frequency values can make it ill-advised to proceed with a frequency encoded feature. Low to moderate duplication can be kept if there is a strong justification for doing so.

**Hint 3: **If you decide it is ill-advised for either or both columns, make sure to remove those features from the dataset!

```{r}
#check original uniqueness
length(unique(pneumoniaFull$State))
length(unique(pneumoniaFull$CountyParish))
```


```{r}
#compare uniqueness with encoded dataset
length(unique(pneumoniaFullEncoded$State))
length(unique(pneumoniaFullEncoded$CountyParish))
```

```{r}
#investigate the duplicates
table(pneumoniaFullEncoded$State)
table(pneumoniaFullEncoded$CountyParish)
```


> Your answer here.

###We are justified in using the State feature, but not the CountyParish feature, because CountyParish did not retain much of its uniqueness after frequency encoding. Originally, there were 1,549 unique county/parish values, but after frequency encoding, the feature collapsed down to only 36 unique frequency values. This indicates heavy duplication, where many different counties now share the same encoded value. As a result, much of the geographic specificity and uniqueness was lost, making the feature less useful and potentially misleading for analysis.
However, the State feature appears more justifiable to retain because it originally contained 57 unique values and still retained 50 unique frequency values after encoding. This suggests that most of the geographic information was preserved with relatively low duplication compared to CountyParish.

```{r}
#remove the CountyParish column
pneumoniaFullEncoded <- pneumoniaFullEncoded %>%
  select(-CountyParish)
```

```{r}
#check
glimpse(pneumoniaFullEncoded)
```

### Finishing touches

#### 1. Ensure that every feature is numeric.

```{r, warning=FALSE, message=FALSE}
load("pneumoniaAnalyze.Rdata")

pneumoniaAnalyze <- pneumoniaAnalyze %>% 
  mutate(across(where(is.character), as.numeric))
```

#### 2. Collapse the `NumberOfCompletedSurveys_`... and `SurveyResponseRatePercent_`... features into a single one, as they are identical / redundant:

```{r}
pneumoniaAnalyze <- pneumoniaAnalyze %>% 
  ## arbitrarily chose as the rep as they are identical
  mutate(NumberSurveysCompleted = NumberOfCompletedSurveys_Cleanliness,              
         SurveyResponseRate = SurveyResponseRatePercent_Cleanliness/100) %>%       ## Turned into an actual rate 
  select(-contains(c("NumberOfCompletedSurveys_", "SurveyResponseRatePercent_")))   ## drop the others
```

Excellent! Now we are down to just 51 features...

```{r, echo=FALSE}
## We can remove the datasets we are no longer using:
rm(pneumoniaFull, pneumoniaFullEncoded, Timely_and_Effective_Care, Unplanned_Hospital_Visits, valueOnly, Outpatient_Imaging_Efficiency, Payment_and_Value_of_Care, paymentOnly, readmissionsClean, valueOnly, hospitalInfo, Maternal_Health, Medicare_Hospital_Spending_Per_Patient, FY_2024_HAC_Reduction_Program, FY_2024_Hospital_Readmissions_Reduction_Program, files, HCAHPS, Healthcare_Associated_Infections, Complications_and_Deaths, dat, temp)
```


#### 3. Identify the target variable.

__Our very, very last steps!__ Your task is to identify the appropriate target variable from the dataset. I have purposefully not gone into great detail about what it would be because choosing the best target is sometimes tricky. 

##### **Question 20**: [2 points] 

Use `pneumoniaAnalyze` to calculate an  `observed_readmission_rate` as the number of readmissions divided by the number of discharges, multiplied by 100, then drop the two columns used to make this new one. Now, using the data and the data dictionary, try to understand the relationship between `observed_readmission_rate` that we just created, the `PredictedReadmissionRate`, `ExpectedReadmissionRate`, and the `ExcessReadmissionRatio`. What is the relationship? What is the appropriate target to choose and why? 

**Hint: **Think about (or perhaps try to investigate) why the **predicted** rate exists. What would be the (dis)advantage to using the predicted vs. the newly created `observed_readmission_rate`? 

> Your answer here.

```{r}
# create observed_readmission_rate using the raw readmission count
# divided by the number of discharges, then multiply by 100.
# after creating the new rate, remove the two raw count columns used to calculate it.
pneumoniaAnalyze <- pneumoniaAnalyze %>%
  mutate(
    observed_readmission_rate = 
      (NumberOfReadmissions / NumberOfDischarges) * 100
  ) %>%
  select(
    -NumberOfReadmissions,
    -NumberOfDischarges
  )

# compare the raw observed rate, CMS predicted rate,
# expected rate, and excess readmission ratio.
summary(
  pneumoniaAnalyze[, c(
    "observed_readmission_rate",
    "PredictedReadmissionRate",
    "ExpectedReadmissionRate",
    "ExcessReadmissionRatio"
  )]
)

# observe  the relationships between the possible target variables.
# use correlation to show whether the variables move together or not
cor(
  pneumoniaAnalyze[, c(
    "observed_readmission_rate",
    "PredictedReadmissionRate",
    "ExpectedReadmissionRate",
    "ExcessReadmissionRatio"
  )],
  use = "complete.obs"
)

```
###I believe observed_readmission_rate is the best target variable because it represents the actual hospital readmission outcome rather than a predicted or expected estimate. The predicted and expected readmission rates are already analytically created using hospital data, adjustments, and statistical modeling. Because of this, using them as the target variable would be similar to modeling another model rather than independently analyzing actual hospital outcomes. In contrast, observed_readmission_rate is calculated directly from the real number of readmissions and discharges, which makes it the most interpretable and realistic target variable for analysis. Additionally, observed_readmission_rate also shows moderate to strong correlations with the other potential target variables, suggesting that it still captures much of the same overall readmission behavior while remaining the most direct measure of what is actually happening in hospitals.




##### **Question 21**: [1 points] 

Why did I have you drop the two columns used to make `observed_readmission_rate`?

> Your answer here.

###We dropped the two columns 'NumberOfReadmissions' and 'NumberOfDischarges' because they were used directly to calculate observed_readmission_rate. If we kept them in the dataset, they would create data leakage because the model could use those original columns to easily reconstruct the target variable. Dropping them helps make sure the model is learning from other hospital features rather than from the exact variables used to create the outcome.


# Exploratory Data Analysis

For your final question, you will kick start our EDA by focusing on the target variable that you identified in Question 20. **You will not be penalized for choosing the wrong target as long as you made an attempt to choose and defend a rational choice.**

##### **Question 22**: [3 points] 

Explore the target variable you identified in Question 20 with at least one other variable. Try to push yourself to try a new style of plot; for example, have you ever made a `density` or `violin` plot? These can be excellent choices when exploring the distribution of a target variable against another variable. Note that, if you choose to compare the target against one of the encoded categorical variables, you will need to properly re-label the categories for your plot.

Full points are awarded for professional plots with axis labels, labeled legends (if appropriate), and creative use of multivariable information. **One bonus point will be awarded if you make a faceted plot or otherwise include information from at least 3 variables.** 

Need inspiration that comes with code?!? Check out the [R Graph Gallery](https://r-graph-gallery.com)!

```{r}
# create a smaller dataset just for eda analysis and plotting
# focusing mainly on the observed_readmission_rate as the target variable selected in Question 20
eda_plot <- pneumoniaAnalyze %>%
  
  # select only the variables needed for this visualization
  select(
    observed_readmission_rate,
    `ComparedToNational_Death rate for pneumonia patients`,
    `ComparedToNational_MRSA Bacteremia`
  ) %>%
  
  # remove missing values so the graph is easier to interpret
  drop_na() %>%
  
  # convert the encoded numeric categories back into readable labels
  mutate(
    death_rate_comparison = factor(
      `ComparedToNational_Death rate for pneumonia patients`,
      levels = c(0, 1, 2),
      labels = c(
        "Worse than National Rate",
        "No Different than National Rate",
        "Better than National Rate"
      )
    ),
    
    mrsa_comparison = factor(
      `ComparedToNational_MRSA Bacteremia`,
      levels = c(0, 1, 2),
      labels = c(
        "Worse than National Benchmark",
        "No Different than National Benchmark",
        "Better than National Benchmark"
      )
    )
  )

# create a violin plot to compare how observed readmission rates are distributed across the national pneumonia death-rate categories
ggplot(
  eda_plot,
  aes(
    x = death_rate_comparison,
    y = observed_readmission_rate
  )
) +
  
  # violin plot helps show the spread and density of the readmission rates
  geom_violin(trim = FALSE) +
  
  # add a boxplot inside the violin plot to make the median/spread easier to see
  geom_boxplot(width = 0.15) +
  
  # split the graph further by MRSA benchmark comparison so we can compare another hospital quality-related feature
  facet_wrap(~ mrsa_comparison) +
  
  # add labels and titles
  labs(
    title = "Observed Pneumonia Readmission Rates by National Death Rate Comparison",
    subtitle = "Faceted by MRSA Bacteremia National Benchmark Comparison",
    x = "Pneumonia Death Rate Compared to National Rate",
    y = "Observed Readmission Rate (%)"
  ) +
  
  # reotate labels slightly so they do not overlap
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

Lastly, make sure to **interpret** your graphic.

> Your answer here.

###Observing the violin plots, one thing I noticed is that the median observed readmission rate is relatively similar across most of the groups, generally ranging between about 16% and 19% for the “Worse than National Benchmark,” “No Different than National Benchmark,” and even the missing/NA categories. This suggests that most hospitals tend to have observed pneumonia readmission rates somewhere between 15% and 20%.

However, the hospitals in the “Worse than National Benchmark” category appear to have a much wider spread and more extreme high and low outliers compared to the other groups. This suggests there may be some association between lower hospital quality benchmarks and  higher readmission rates.

At the same time, there is still substantial overlap between the groups. For example, hospitals categorized as “No Different than National Benchmark” still have readmission rates similar to some hospitals in the worse benchmark categories, and the medians across groups are still fairly close. Because of this, although there appears to be some relationship between hospital quality measures and observed pneumonia readmission rates, these variables alone likely do not fully explain the variation in readmissions, and there are probably additional factors contributing to the differences across hospitals.

# References